Yaml

---
title: "Chapter A5: Technical Report for Likelihood Subgradient Based Samplers"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter A5: Technical Report for Likelihood Subgradient Based Samplers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(glmbayes)
```
## 1. Introduction

## 2. Theoretical background


## 3 Implementation


### 3.1 Mean/offset shift and reparametrization

### 3.2- posterior optimization

### 3.3 — model standardization - glmb_Standardize_Model(y, x, P, bstar, A1)

This functions starts with basic information about the model in the argument list and then uses the following steps to further standardize the model (the model is already assumed to have a 0 prior mean vector when this step is applied).

1) An eigenvalue composition is applied to the posterior precision matrix, and the model is (as an interim step) standardized to have a posterior precision matrix equal to the identity matrix. Please note that this means that the prior precision matrix after this step is "smaller" than the identity matrix.

2) A diagonal matrix epsilon is pulled out from the standardized prior precision matrix so that the remaining part of the prior precision matrix still is positive definite. That part is then treated as part of the posterior for the rest of the standardization and simulation and only the part connected to epsilon is treated as part of the prior. Note that the exact epsilon chosen seems not to matter. Hence there are many possible ways of doing this standardization and future versions of this package may tweak the current approach if it helps improve numerical accuracy or acceptance rates.

3) The model is next standardized (using a second eigenvalue decomposition) so that the prior (i.e., the portion connected to epsilon) is the identity matrix. The standardized model then simutaneously has the feature that the prior precision matrix is the identity matrix and that the data precision A (at the posterior mode) is a diagonal matrix. Hence the variables in the standardized model are approximately independent at the posterior mode.


### 3.4 — Envelope construction - EnvelopeBuild


The implementation of `EnvelopeBuild` follows the envelope construction in Nygren & Nygren (2006) for models in standard form (see Section 3–3.3). Each computational step corresponds to a theoretical guarantee:


#### 3.4.1 Envelope Sizing and Optimization

When `GridType = 2` the grid size is chosen to balance envelope build cost against sampling efficiency (acceptance rate). A larger grid typically increases acceptance but costs more time and memory to construct; a smaller grid builds quickly but yields more rejected proposals. The optimizer uses information about per-dimension curvature and the requested number of draws to find a pragmatic trade-off.

##### How automatic sizing is invoked

The wrapper scales the requested draw count `n` by the number of detected OpenCL cores (when available) before calling the optimizer. This encourages richer envelopes when envelope construction can be parallelized, since build cost can be amortized across cores


##### How the optimization is performed

This section explains, step by step and with the exact formulas used in the code, how EnvelopeOpt converts the inputs — the diagonal precisions `a1` (you called this `a_2` in the wrapper) and the desired draw count `n` — into a per-dimension grid specification (a vector of 1s and 3s). The explanation follows the code literally so you can copy it into the vignette and understand every intermediate quantity.

###### Inputs
- `a1` — numeric vector of length `p` containing the diagonal elements of the posterior precision matrix in standardized coordinates. Larger `a1[i]` means stronger curvature (narrower posterior) in coordinate `i`.
- `n` — integer: target number of posterior draws (possibly scaled by core count before the call).

###### Top-level idea

EnvelopeOpt evaluates a small set of candidate grids. Candidate r (r = 0..p) splits the r "most variable" coordinates from 1 component → 3 components; all other coordinates remain 1 component. For each candidate the function computes a two-term cost:
- a build-cost proxy (`intest`), roughly the number of grid cells; and
- a sampling-cost proxy (`n * slopeest`), proportional to n and an aggregate scale factor `slopeest`.

EnvelopeOpt returns the candidate whose sum (build-cost + sampling-cost) is minimal:
evalest[r] = intest[r] + n * slopeest[r].


##### EnvelopeOpt: purpose, inputs, and outputs

- **Purpose:** choose how many components to allocate per parameter dimension (currently `1` or `3`) to minimize an estimated total computational cost.  
- **Inputs:**
  - `a1` (called `a_2` where invoked): diagonal elements of the posterior precision matrix in standardized coordinates (per-dimension curvature).  
  - `n` (possibly scaled): target number of posterior draws.  
- **Output:** integer vector of length `p` giving per-dimension component counts (`1` or `3`), consumed by `EnvelopeBuild` to construct the grid.

##### EnvelopeOpt algorithm (high-level)

- Rank dimensions by variability using `rank(1/(1 + a1))`, so dimensions with larger curvature are prioritized for splitting.  
- Construct `p + 1` candidate grids:
  - Candidate 0: no splits (all components = 1).  
  - Candidate i (i = 1..p): split the i-th ranked dimension(s) from `1 → 3` components.  
- For each candidate compute:
  - `intest`: approximate number of grid cells (`1, 3, 3^2, ...`).  
  - `slopeest`: product of per-dimension scale factors (initially `1 + a1`; replaced by `2/sqrt(pi)` for split dims).  
  - `evalest := intest + n * slopeest` (heuristic total cost = build cost + expected sampling cost).  
- Select the candidate with minimal `evalest` and return its `dimcount` row.

##### Interpretation of the cost heuristic

- `intest` approximates envelope build complexity (grows rapidly with the number of split dimensions).  
- `n * slopeest` approximates cumulative sampling cost (larger `slopeest` means more candidate proposals are needed per accepted draw).  
- The algorithm selects the split pattern that minimizes the sum of build cost and expected sampling cost under this heuristic.

##### Behavior when GridType != 2

If `GridType` is not `2`, the optimizer is bypassed and a deterministic grid policy is applied. The chosen `GridType` uniquely determines grid size and componentization; automatic scaling and core-count adjustments are not used.

##### Practical recommendations

- **Default:** keep `GridType = 2` so the package tunes the grid automatically.  
- For debugging or single-draw runs, set `GridType != 2` or disable parallel/OpenCL (`use_parallel = FALSE`, `use_opencl = FALSE`) to force a small, deterministic grid.  
- Record detected core count and `GridType` when running large simulations for reproducibility; for strict reproducibility across machines, use a fixed `GridType` or disable OpenCL/parallelism.

##### Illustrative example

- With moderate `p` and large `n` plus GPU enabled, the wrapper scales `n` and `EnvelopeOpt` commonly returns more `3`-component splits, yielding a richer grid and higher acceptance at the cost of longer build time.  
- For `n = 1` and no parallelism, `EnvelopeOpt` will usually prefer smaller grids the minimal grid (all ones), leading to fast envelope construction and one-off sampling.

#### 3.4.2 Compute width parameters** \(\omega_i\) **from the diagonal precision matrix.**  
   Let \(\theta^{\ast}\) denote the posterior mode. For each dimension \(i\), define
   \[
     \omega_{i} :=
     \frac{\sqrt{2} - \exp\!\big(-1.20491 - 0.7321\,\sqrt{0.5 - \frac{\partial^{2}\log f(\theta^{\ast}\mid y)}{\partial\theta_{i}^{2}}}\big)}
          {\sqrt{1 - \frac{\partial^{2}\log f(\theta^{\ast}\mid y)}{\partial\theta_{i}^{2}}}}.
   \]
   The widths \(\omega_i\) are derived from the local curvature of the log-likelihood at the posterior mode. This choice ensures the three-interval construction per dimension yields an envelope whose efficiency does not deteriorate with sample size.

#### 3.4.3 Construct per-dimension intervals around the posterior mode.**  
   Set
   \[
     \ell_{i,1} = \theta^{\ast}_{i} - 0.5\,\omega_{i}, \qquad
     \ell_{i,2} = \theta^{\ast}_{i} + 0.5\,\omega_{i},
   \]
   and define three intervals per dimension:
   \[
     A_{i,1} = (-\infty,\ell_{i,1}), \quad
     A_{i,2} = [\ell_{i,1},\ell_{i,2}], \quad
     A_{i,3} = (\ell_{i,2},\infty).
   \]
   For each dimension \(i\) let \(J_{i}=\{1,2,3\}\) and \(J=\prod_{i=1}^p J_i\), so \(|J|=3^p\). Each \(j\in J\) is \((j_1,\dots,j_p)\) and define
   \[
     A^{\ast}_{j} = \prod_{i=1}^{p} A_{i,j_{i}}.
   \]
   The collection \(\{A^{\ast}_{j}:j\in J\}\) forms a partition of \(\Theta\).

#### 3.4.4 Select tangency points \(\bar{\theta}_{j}\) for each partition element.**  
   For each \(j\in J\) define index subsets
   \[
     C_{j1} = \{i : j_{i} = 1\}, \quad
     C_{j2} = \{i : j_{i} = 2\}, \quad
     C_{j3} = \{i : j_{i} = 3\}.
   \]
   Define the tangency point componentwise by
   \[
     \bar{\theta}_{j,i} =
     \begin{cases}
       \theta^{\ast}_{i} - \omega_{i}, & i \in C_{j1}, \\
       \theta^{\ast}_{i},              & i \in C_{j2}, \\
       \theta^{\ast}_{i} + \omega_{i}, & i \in C_{j3}.
     \end{cases}
   \]
   These tangency points ensure the envelope touches the log-likelihood at representative points in each interval, guaranteeing dominance and tightness.

#### 3.4.5 Build the Cartesian grid of tangency points (full grid).**  
   The Cartesian product of per-dimension tangency choices yields the complete set of \(3^p\) tangency points. These represent the restricted densities discussed in the paper and guarantee coverage of the parameter space.

#### 3.4.6 Evaluate negative log-likelihood and gradients at each grid point.**  
   Compute \(h_{\bar{\theta}}(\cdot)\) and the subgradients \(c(\bar{\theta})\) at each tangency point. These quantities define the Likelihood Subgradient Densities and are used in the accept–reject procedure. Implementations:
   - CPU routines: `f2_*` (neg-log-likelihood) and `f3_*` (gradient).  
   - GPU/OpenCL: `f2_f3_opencl` computes both together for efficiency.

#### 3.4.7 Evaluate restricted multivariate normal log-densities via `Set_Grid_C2_pointwise`.**  
   Each restricted density corresponds to a subset of the partition and is normalized as in Remark 5 of the paper. This step yields log-density values needed to form the mixture envelope (Claim 2).

#### 3.4.8 Compute component log-probabilities and normalization constants via `setlogP_C2`.**  
   Compute mixture constants \(\tilde{a}\) and mixture weights \(\tilde{p}_i\) as in Remark 6 so the mixture envelope is properly normalized and usable as a dominating density for accept–reject sampling.

#### 3.4.9 Normalize component probabilities (PLSD) and optionally sort grid components.**  
   Normalize so the mixture of restricted densities is a valid probability distribution (Claim 2). Sorting components is an implementation optimization to improve sampling efficiency and early-rejection performance.



### 3.5 — simulation (serial vs parallel)

### 3.6 — undo-standardization and postprocessing




## 4 Diagnostics, testing, and developer notes
Diagnostic checks and outputs
Numerical stability caveats and remedies
Suggested unit tests
Roxygen and API documentation tips
Extension points and future improvements

## 5. Appendix
API reference (arguments and returned list)
Troubleshooting checklist
Suggested figures and tables
References



