---
title: "Chapter C01: The Gaussian example -- model, data and design"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter C01: The Gaussian example -- model, data and design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(lmebayesCore)
```

```{r load-stored}
C01 <- readRDS(system.file(
  "extdata", "Chapter-C01-bwc-setup.rds", package = "lmebayesCore"
))
meta <- C01$meta
```

The chapters that follow each take one model class and work through its
posterior algebra, its convergence rate, and simulation evidence that the
sampler reaches the distribution it claims to. Comparing those results
across chapters only means something if they are computed on the same
problem, so the Gaussian chapters share one theoretical model, one dataset,
and one estimated model. This chapter sets out all three. Nothing is
sampled here; that begins in `Chapter-C02`.

# The theoretical model

A mixed model describes data that arrive in groups — pupils within schools,
patients within hospitals, measurements within subjects. Its defining idea
is that each group gets its own regression coefficients, and those
coefficients are themselves given a model rather than estimated freely.

That gives two levels of unknowns, and the distinction between them runs
through the whole series.

## Group effects

Write \(j = 1, \dots, J\) for the groups. Group \(j\) contributes \(n_j\)
observations \(y_j\) with its own design matrix \(D_j\), and has its own
coefficient vector \(\beta_j\):

\[
y_j \mid \beta_j \sim N\!\left(D_j \beta_j,\; \sigma^2 I_{n_j}\right).
\]

The \(\beta_j\) are the **group effects**. There are \(J \times
p_{\mathrm{re}}\) of them, which is far too many to estimate
independently from the data available, and \(\sigma^2\) is the observation
variance describing how tightly a group's observations cluster around its
own fitted line.

## Population effects

The group effects are not free parameters. They are assumed to be drawn
from a common distribution, whose mean is itself a regression on
group-level characteristics \(\mathcal{W}_j\):

\[
\beta_j \mid \gamma \sim N\!\left(\mathcal{W}_j \gamma,\; \Psi\right).
\]

The vector \(\gamma\) holds the **population effects**: the coefficients
that explain how group effects vary with group characteristics. The matrix
\(\Psi\) is the between-group variance, describing how far groups scatter
around the population pattern after those characteristics are accounted
for. Finally \(\gamma\) receives a prior of its own,

\[
\gamma \sim N(\mu_0, V).
\]

Two features of this parameterization matter for what follows. First, it is
*centered*: \(\beta_j\) is the group's own coefficient vector, not a
deviation from an average. Second, both levels have the same shape — a
normal likelihood paired with a normal prior — which is what allows a
single sampling machine to serve the group level and the population level
alike. This is the Lindley--Smith formulation [@LindleySmith1972], and
`Chapter-B01` describes how the engine exploits it.

## What is fixed and what has a prior

The group effects \(\beta_j\) and the population effects \(\gamma\) always
have priors and are always sampled; they are what a fit is *for*. The two
variance quantities are the ones that may be either estimated or held at a
known value, and choosing between them is what separates the chapters of
this series:

| Chapter | \(\beta_j\) | \(\gamma\) | \(\Psi\) | \(\sigma^2\) | Sampler |
|---------|-------------|------------|----------|--------------|---------|
| `C02` | prior | prior | fixed | fixed, pooled | exact iid |
| `C03` | prior | prior | fixed | fixed, pooled | two-block Gibbs |
| `C06` | prior | prior | fixed | prior, per group | two-block Gibbs |
| `C07` | prior | prior | prior | fixed, pooled | two-block Gibbs |
| `C09` | prior | prior | prior | prior, per group | two-block Gibbs |

Fixing both variances is what makes the posterior exactly normal, so that
it can be drawn from directly, which is why `C02` comes first. `C03` then
runs an iterative sampler at that same posterior, where the correct answer
is already known and any error is visible. Each later row relaxes one more
assumption, and the price of each relaxation is a convergence rate that has
to be certified rather than assumed.

# The Big Word Club data

The running example is `big_word_club` from the **bayesrules** package
[@JohnsonLauEtAl2022]. It comes from an evaluation of Big Word Club, a
digital vocabulary program aimed at narrowing the gap in vocabulary between
children from higher- and lower-income households [@KalilMayerOreopoulos2020].

The grouping structure is the reason this dataset suits the series: pupils
are nested within schools, schools differ in both intake and circumstances,
and the number of pupils per school is small enough that pooling across
schools genuinely matters.

The version used here is a prepared analysis dataset, restricted to
complete records with a valid assessment score:

```{r dims}
c(pupils = meta$n_obs, schools = meta$J, columns = meta$n_cols)
```

Each school contributes only a handful of pupils:

```{r sizes}
summary(C01$sizes)   # pupils per school
```

Estimating three coefficients per school from around a dozen observations
is exactly the situation in which pooling across schools earns its keep,
and it is why the shrinkage that later chapters display is substantial.

## Variables used

The full dataset carries `r meta$n_cols` columns covering pupils, teachers
and schools. The model below uses seven of them:

```{r vars}
knitr::kable(C01$model_vars[, c("variable", "level", "meaning", "values")])
```

The `level` column is worth noting. `private_school`, `title1` and
`free_reduced_lunch` are properties of the *school*, so they take one value
for all pupils in a school and are natural candidates for explaining
group effects. The `distracted_*` ratings vary from pupil to pupil within a
school.

Other columns not used here include a treatment indicator, pupil age and
sex, two earlier assessment scores, and a set of teacher-level measures:

```{r other-cols}
setdiff(C01$all_columns, C01$model_vars$variable)
```

# The model being estimated

```{r formula, eval = FALSE}
form <- score_ppvt ~
  private_school + title1 + free_reduced_lunch +
  distracted_ppvt + distracted_a1 +
  free_reduced_lunch:distracted_a1 +
  (1 + distracted_ppvt + distracted_a1 || school_id)
```

The bracketed term specifies the group effects. Each school gets its own
intercept and its own two distraction slopes, so schools may differ both in
average score and in how strongly distraction depresses scores. That makes
\(p_{\mathrm{re}} = `r meta$p_re`\):

```{r re-names}
meta$groupef_names
```

The `||` requests that these three be uncorrelated, which makes \(\Psi\)
diagonal. Everything outside the brackets specifies the population effects
that explain those three quantities.

## How the population effects are indexed

Because \(\gamma\) explains group effects, it is not a flat vector. It is
grouped by *which* group effect each coefficient explains:

```{r popef-names}
C01$popef_names
```

Reading these three blocks in turn: the intercept block holds the
school-level predictors of a school's average score. The `distracted_ppvt`
block holds a single entry, its own intercept, which is the average
`distracted_ppvt` slope across schools. The `distracted_a1` block holds
that same average plus a `free_reduced_lunch` term — a cross-level
interaction, asking whether distraction matters more in poorer schools.

This indexing is why the summary tables in later chapters label rows
`component::parameter`, and why lining them up against a flat `lmer`
coefficient vector takes some care.

## The priors

`Prior_Setup_GLMM()` calibrates priors from the data rather than asking for
hyperparameters directly:

```{r prior-setup, eval = FALSE}
ps <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01)
pf <- pfamily_list(ps)
```

For \(\gamma\) it produces the normal prior of Section 1, with `pop.pwt`
setting its weight. At `0.01` that is about one percent of the weight of
the data, so it regularizes without materially moving the answer:

```{r pop-prior}
C01$pop_prior
```

The prior means are zero apart from the intercept, and the standard
deviations are large relative to the coefficients they cover, which is what
makes this prior weak.

`pfamily_list(ps)` then wraps each component of \(\Psi\). A `dNormal`
component declares \(\tau^2_k\) known, so no sampling step is needed for
it; `dIndependent_Normal_Gamma` estimates it instead. The same choice
arises one level down for \(\sigma^2\). Which are fixed is precisely the
table in Section 1.

Where a quantity is fixed, it is worth knowing what it is fixed *at*. The
natural assumption is the REML estimate, and that assumption is wrong:

```{r fixed-vs-reml}
tab <- C01$fixed_vs_reml
tab$ratio <- round(tab$calibrated / tab$reml, 3)
tab
```

The between-school variances are calibrated to values well below the REML
ones, while the residual runs the other way, because `Prior_Setup_GLMM()`
uses a pooled marginal estimate rather than the classical residual:

```{r dispersion-source}
c(source = C01$dispersion_source,
  classical_reml = round(C01$dispersion_classical, 2))
```

This does not affect whether a sampler is correct: any fixed \(\Psi\) and
\(\sigma^2\) still define a legitimate posterior, and reaching that
posterior is what the later chapters verify. It does mean the draws are not
interchangeable with an `lmer` plug-in analysis, which is why the reference
comparisons there are judged against `lmer`'s own standard errors rather
than against exact agreement. `Chapter-B05` covers the calibration itself.

With the model specified and the priors calibrated, the first question is
what happens when both variances are simply held at these values. That is
`Chapter-C02`.

# See also

- `vignette("Chapter-C02")` — the exact iid sampler on this design.
- `vignette("Chapter-B01")` — the two-block architecture built on the
  two-stage form.
- `vignette("Chapter-B05")` — prior calibration and estimated variance
  components.
- `vignette("Chapter-B00")` — notation and the engine roadmap.
- `?model_setup`, `?Prior_Setup_GLMM`, `?pfamily_list`
- `data-raw/make_bwc_full_rank.R` and
  `data-raw/chapter_C_gaussian_setup.R` in the package sources for the code
  that prepares this dataset and design.
