---
title: "Chapter B08: Design, identifiability and rank conditions"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B08: Design, identifiability and rank conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B08: Design, identifiability and rank conditions

`model_setup()` does three jobs: it turns a formula into matrices, it fits
a reference model, and it runs a two-level identifiability check. This
chapter is about the third, because the conditions it tests are the same
ones the ergodicity theory of Chapter B02 assumes, and because the two
levels fail in importantly different ways.

The headline distinction, since it is the thing most often gotten
backwards:

- **A rank-deficient group is survivable.** It stays in the model, and in
  the direction its data cannot identify it is drawn from its prior.
- **A failed level-2 condition is not.** No group can supply the missing
  information, and the chain is null recurrent in that direction of
  \(\gamma\) even under a proper prior.

# What `model_setup()` returns

```{r data}
data(sleepstudy, package = "lme4")
dat <- sleepstudy
dat$Days_c <- dat$Days - mean(dat$Days)

design <- model_setup(Reaction ~ Days_c + (1 + Days_c || Subject), data = dat)
design
```

The call hierarchy is short:

1. `extract_re_hyper_matrices()` builds `y`, `D`, `group`, `W`, `weights`,
   `offset`, `groupef.names`, and `popef.moderation`.
2. A reference fit runs — `lme4::lmer()` or `lme4::glmer()`, plus
   `glmmTMB::glmmTMB()` when `dispformula` is per-group and Gaussian.
3. `check_identifiability()` computes the rank and estimability
   diagnostics and copies them onto the result.

`subset`, `weights`, `na.action`, `offset` and `contrasts` are forwarded to
*both* the design extraction and the reference fit, so rows and factor
coding cannot drift apart between them.

The fields that matter for this chapter:

```{r fields}
str(design[c("groupef.rank", "groupef.estimable",
             "popef.rank", "popef.deficient", "popef.rank_ok")],
    max.level = 1)
```

Note that `model_setup()` does **not** calibrate priors. That is
`Prior_Setup_GLMM()`, which consumes these diagnostics.

# Level 1: within-group rank

For each level of the grouping factor, take that group's rows of \(D\) and
ask whether the resulting \(D_j\) has full column rank:

\[
\texttt{groupef.rank}[j] =
\bigl(n_j \ge p_{re}\bigr)
\ \wedge\
\bigl(\mathrm{rank}(D_j) = p_{re}\bigr).
\]

Implemented as `Matrix::rankMatrix(D_j, method = "qr")` at Matrix's default
tolerance, with the row-count precondition checked first — a group with
fewer observations than coefficients cannot be full rank, and testing rows
first avoids the degenerate call.

```{r level1}
table(design$groupef.rank)
```

All 18 subjects here have ten distinct `Days_c` values, so all pass.

## What failure means, and what it does not

A rank-deficient \(D_j\) means there is a direction \(v\) with
\(D_j v = 0\): a linear combination of that group's coefficients the group's
own data say nothing about. In the notation of
`inst/BLOCK_GIBBS_ERGODICITY.md`, the group contributes \(C_b = 0\) in that
direction.

Under a flat prior on \(\mu\) the joint posterior is then improper and the
chain is null recurrent. But that is a statement about a *flat* prior. With
the proper prior this package always supplies, the conditional for
\(\beta_j\) in the deficient direction is simply the prior conditional
\(N(\mathcal{W}_j\gamma, \tau^2)\) — a perfectly ordinary draw.

So the group is **flagged, not dropped**. It stays in \(D\), it stays in
the reference fit, it receives coefficients, and it is shrunk to the
population value in the direction it cannot inform. It is a passenger in
the sampler, not a threat to it.

What it *is* excluded from is what comes next: the level-2 rank check, and
prior calibration in `Prior_Setup_GLMM()`. A group that cannot identify its
own coefficients cannot help identify the population's.

The one context where deficient groups genuinely must be removed is
independent block fitting (`lmbBlock()`/`glmbBlock()` in **lmebayes**),
where there is no population level to fall back on. There, level 1 is the
only condition, and failing it means the block cannot be fit at all. See
Chapter B09.

# The estimability check

Full rank is necessary but not sufficient. `check_identifiability()` runs a
second, separate test, and the separation is deliberate enough to warrant
its own line in the print output.

The reason is that algebraic rank is a statement about \(D_j\) alone, while
what actually matters is whether a finite estimate exists for \(D_j\)
*together with* \(y_j\) and the link. A group can have a perfectly
conditioned design and still admit no finite MLE:

| Family | Failure mode |
|---|---|
| binomial | complete or quasi-complete separation — fitted probabilities pinned at 0 or 1 |
| poisson | all-zero response; the log-link intercept diverges to \(-\infty\) |
| Gamma | non-positive responses, or \(n_j \le p_{re}\) leaving no residual df for dispersion |
| gaussian | \(n_j \le p_{re}\): coefficients exist, but the variance is not estimable |

Gaussian is the instructive case. OLS returns coefficients whenever \(D_j\)
has full rank, but with \(n_j = p_{re}\) the fit is exact, the residual sum
of squares is zero, and \(\sigma^2_j\) has nothing to be estimated from.
Since the calibration in Chapter B07 needs a dispersion estimate per group,
that group is unusable for calibration even though its coefficients are
fine.

Mechanically, `.lmebayes_block_glm_estimable()` runs only on groups that
already passed level 1. Cheap pre-checks come first (single outcome level,
all-zero counts, non-positive Gamma responses, insufficient residual df).
For binomial, Poisson and Gamma it then fits
`stats::glm(y ~ . - 1, family = family, control = glm.control(maxit = 50))`
per group and inspects the result: wrong coefficient count, rank
deficiency, `NA` or non-finite coefficients, non-finite `vcov`, or — for
binomial — any fitted probability outside \([10^{-7},\, 1 - 10^{-7}]\),
which is the separation signature. Gaussian skips the `glm()` fit entirely
and only checks residual df.

The per-group verdicts land in a diagnostic frame:

```{r glm-check}
design$groupef.glm_check
```

`NULL` here, because the Gaussian path with \(n_j = 10 > p_{re} = 2\) has
nothing to report. On a binomial or Poisson fit this is a data frame with
`group`, `n`, `p`, `groupef.rank`, `estimable`, and a `note` giving the
reason for any failure.

# Level 2: across-group rank

The second condition asks whether the groups that *can* speak collectively
span the population predictor space. For each random coefficient \(k\),
restrict the hyper-design to estimable groups and test its rank:

\[
\texttt{popef.rank}[k] =
\Bigl(\mathrm{rank}\bigl(\mathcal{W}_k[\text{estimable}, ]\bigr) = q_k\Bigr).
\]

```{r level2}
design$popef.rank
design$popef.rank_ok
lapply(design$W, dim)
```

Two details are easy to miss. The subset is **estimable**, not merely
full-rank — a separated binomial group carries no usable information about
\(\gamma\) regardless of its design matrix. And row \(j\) of `W[[k]]`
corresponds to `levels(group)[j]` positionally, so the subsetting is by
index rather than by name.

For an intercept-only hyper design — `W[[k]]` a column of ones, which is
the default for any random slope without cross-level moderation — the
condition collapses to "at least one estimable group exists". It becomes
substantive only when group-level covariates enter, which is where a
level-2 failure is a real possibility: a school-level covariate that varies
in the raw data can easily be constant across the *estimable* schools.

`popef.rank_ok` is the scalar gate, and `Prior_Setup_GLMM()` requires it to
derive default hyperpriors automatically.

## Why this one is fatal

The two levels map onto the ergodicity conditions like this:

| Level | Condition | Guarantees |
|---|---|---|
| Within group | \(\mathrm{rank}(D_j) = p_{re}\) | \(\beta_j\) identified by group \(j\)'s data (\(C_b > 0\)) |
| Across groups | \(\mathrm{rank}(\mathcal{W}_k[\text{est.}, ]) = q_k\) | \(\gamma_k\) identified through the data-bearing groups |

Both are required for geometric ergodicity under an improper or near-flat
prior on \(\mu\). If either fails there is a direction in \((\beta, \gamma)\)
along which the chain is null recurrent.

The asymmetry appears once the prior is proper. A level-1 failure is then
repaired by the prior: the missing likelihood information is supplied by
\(N(\mathcal{W}_j\gamma, \tau^2)\), which is a genuine, informative
conditional built from the *other* groups. A level-2 failure is not
repairable that way, because the thing left unidentified is \(\gamma\)
itself — the very quantity the level-1 repair depends on. There is no
further level to borrow from.

# Fixed-effect grammar

The last thing `model_setup()` enforces is which fixed effects are
admissible, and the rule is narrower than `lme4`'s. Exactly three kinds are
allowed:

1. **Level-2 covariates** — constant within each group (`private_school`,
   `title1`).
2. **Population mean slopes** — a main effect whose name matches a random
   slope, as `Days_c` does alongside `(1 + Days_c || Subject)`.
3. **Cross-level moderation** — `moderator:random_slope`, where the
   moderator is group-constant.

Anything else stops:

```
Fixed effects must be constant within {group} (level-2), a population mean
slope (matching a random slope name), or cross-level RE moderation
(moderator:random_slope); not allowed: {cols}
```

This is not arbitrary strictness. It is what makes the hyper-design
\(\mathcal{W}\) well defined: every admissible fixed effect maps onto a
column of some \(\mathcal{W}_k\), and every column of every
\(\mathcal{W}_k\) traces back to an admissible fixed effect. The
construction is:

| Random coefficient | `W[[k]]` |
|---|---|
| `(Intercept)` | all group-constant fixed columns |
| slope with cross-level moderation | `X_group[, c("(Intercept)", moderator)]` |
| any other slope | intercept only, `matrix(1, J, 1)` |

A level-1 fixed effect that varies *within* group has no such home. It is
not that the package refuses to fit it — it is that the centered
parameterization of Chapter B02 has no slot for it, since such a term is
already absorbed into the group's own \(\beta_j\).

Three further restrictions, each with its own message:

**One grouping factor.**
`extract_re_hyper_matrices() requires exactly one grouping factor.`

**Uncorrelated random effects only.** Correlated terms are rejected with a
pointer to `||`:

```
Correlated random effects are not supported (off-diagonal RE covariance).
Use '||' for uncorrelated terms, e.g. (1 + x || group) instead of
(1 + x | group).
```

A diagonal \(\Psi\) is what lets Block 2 factor into independent
per-component updates, which is what the ING machinery of Chapter B05
assumes.

There is also a small `lme4` idiosyncrasy the package catches directly:
`(1 || group)` is not valid `lme4` syntax, so it is intercepted with advice
to write `(1 | group)` instead.

**`dispformula` requires Gaussian:**

```
'dispformula' must be ~1 for family = {family}() (no observation-level
dispersion parameter for per-group measurement-dispersion calibration).
```

# Reading the print output

`print.model_setup()` mirrors the structure above. Section 1 is level 1 —
observation and group counts, then `Full-rank D_j: {n} of {J} groups` and,
when a GLM check ran, `Full-rank & estimable: {n} of {n} full-rank`. Up to
ten offending group names are listed for each.

Section 2 shows the level-2 formulas, one line per random coefficient,
which is the quickest way to confirm that a group-level covariate landed on
the term you intended.

Section 3 reports level-2 rank, headed by
`(Restricted to {n} estimable {group})` so the subset is explicit, then per
term `groups=... predictors=... full-rank|RANK-DEFICIENT`, the deficiency
flags, and `popef.rank_ok`. Deficient terms get a NOTE suggesting either
dropping predictors or merging levels.

# A worked failure

The cleanest way to see both levels is to break one on purpose. Truncate
one subject to a single observation:

```{r broken}
bad <- dat[!(dat$Subject == "308" & dat$Days > 0), ]
bad$Subject <- droplevels(bad$Subject)

design_bad <- suppressWarnings(
  model_setup(Reaction ~ Days_c + (1 + Days_c || Subject), data = bad)
)
design_bad
```

Subject 308 now has one row and two coefficients, so level 1 fails for it,
and — since a failed rank check short-circuits the estimability test — it
drops out of the estimable set too. The report says so on both lines:
`Full-rank D_j: 17 of 18 groups` naming 308, then
`Full-rank & estimable: 17 of 17 full-rank (17 of 18 total Subject)`.

Level 2 still passes, and the header makes the subset explicit:
`(Restricted to 17 estimable Subject)`. With an intercept-only hyper design
seventeen groups are ample, so `popef.rank_ok` is `TRUE`.

This is the survivable case. Subject 308 stays in the model with 171
observations still counted, and it will be estimated almost entirely from
the population — which is the right answer for a subject with one
observation. Nothing about the sampler is compromised.

Contrast what a level-2 failure would take: every estimable group sharing
the same value of some group-level covariate. Then no rearrangement of the
data helps, and the only fixes are the ones the NOTE suggests — drop the
predictor, or merge levels so that more groups become estimable.

# See also

- `vignette("Chapter-B02")` — the ergodicity conditions these tests check.
- `vignette("Chapter-B09")` — level 1 alone, in the independent-block case.
- `vignette("Chapter-02", package = "lmebayes")` — the fixed-effect rules
  for users.
- `?model_setup`, `?Prior_Setup_GLMM`
- `inst/BLOCK_GIBBS_ERGODICITY.md` in the package sources.
