---
title: "Chapter B05: Estimated variance components (ING Block 2)"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B05: Estimated variance components (ING Block 2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B05: Estimated variance components

Chapter B04 held every \(\tau^2_k\) fixed. This appendix documents what
happens when a component's prior is `dIndependent_Normal_Gamma` (ING)
instead of `dNormal`, so that \(\tau^2_k\) is sampled alongside the
population coefficients \(\gamma_k\).

The route becomes `lmm_fixed_estimated`, implemented by
`rLMMNormal_reg_estimated_vcov()` with `needs_pilot = TRUE`.

# Block 2 as a regression on the random effects

The key re-framing is that, conditional on the current Block 1 state, each
random-effect component \(k\) is an ordinary Gaussian regression whose
*response* is the vector of group coefficients.

Fix component \(k\). Let \(b_{jk}\) be group \(j\)'s coefficient, \(X_k\)
the \(J \times p_k\) level-2 design (`design$W[[k]]`), and \(\gamma_k\) the
level-2 coefficients. Then

\[
b_{jk} \sim N\!\bigl(X_k[j,\cdot]^\top \gamma_k,\ \tau^2_k\bigr),
\qquad j = 1, \dots, J,
\]

or in stacked form \(\mathbf{b} = X_k\gamma_k + \varepsilon\) with
\(\varepsilon \sim N(0, \tau^2_k I_J)\).

That is a linear model with \(J\) observations, and \(\tau^2_k\) is its
residual variance. Once you see it this way, the ING prior is just the
standard independent normal–gamma prior for a regression:

\[
\gamma_k \sim N(\mu_\gamma, \Sigma_\gamma), \qquad
\lambda_k = 1/\tau^2_k \sim \mathrm{Gamma}(\mathrm{shape}, \mathrm{rate}),
\]

with the two independent a priori — which is what distinguishes ING from
the conjugate normal-gamma, where \(\Sigma_\gamma\) would be scaled by
\(\tau^2_k\).

Alignment matters: the order of `b_k` must match the rows of `X_k`, which
`two_block_align_b_to_xhyper(b_k, X_k, group_levels)` enforces.

Note what does **not** enter. `mu_all` is a Block 1 object; the \(\tau^2\)
update needs only the residual \(\mathbf{b} - X_k\gamma_k\), which is the
same as subtracting row \(k\) of `mu_all`. Nor does another component's
state appear — a diagonal \(\Psi\) is exactly what makes Block 2 factorize
across \(k\).

# Two ways to update

## Method A: closed-form conditional mode

Freezing \(\gamma_k\) reduces the problem to conjugate Gamma–Normal
algebra. With \(\eta_j = X_k[j,\cdot]^\top\gamma_k\) and
\(\mathrm{RSS} = \sum_j (b_{jk} - \eta_j)^2\), \(J\) independent normals of
variance \(\tau^2_k\) contribute \(J/2\) to the shape and
\(\mathrm{RSS}/2\) to the rate:

\[
\mathrm{shape}_{\mathrm{post}} = \mathrm{shape} + \frac{J}{2},
\qquad
\mathrm{rate}_{\mathrm{post}} = \mathrm{rate} + \frac{\mathrm{RSS}}{2},
\]

and the implementation reports the mode

\[
\widehat{\tau^2_k} = \frac{\mathrm{rate}_{\mathrm{post}}}
                          {\mathrm{shape}_{\mathrm{post}} - 1}.
\]

Neither \(\mu_\gamma\) nor \(\Sigma_\gamma\) appears, because \(\gamma_k\)
is held fixed. This path is not the production sampler; it is the algebra
that Method B collapses to when \(\gamma_k\) is frozen, and it is what the
ICM start uses.

## Method B: `rglmb()`, the production path

The Gibbs Block 2 update in `two_block_block2_one_chain()` fits the full
ING model rather than freezing \(\gamma_k\), by handing the regression to
**glmbayesCore**:

```r
fit_k <- rglmb(
  n       = 1L,
  y       = b_al,       # aligned group coefficients
  x       = X_k,        # level-2 design
  family  = gaussian(),
  pfamily = pf_ing      # dIndependent_Normal_Gamma(mu, Sigma, shape, rate)
)
tau2_k      <- fit_k$dispersion[1L]
gamma_k_new <- fit_k$coefficients[1L, ]
```

So \(\gamma_k\) and \(\tau^2_k\) are updated *jointly*, from their exact
joint conditional, using the same envelope sampler documented in
`vignette("Chapter-A05", package = "glmbayes")`. That joint update is what
keeps the two-block structure intact: Block 2 is still one draw, not a
sub-Gibbs alternation between \(\gamma\) and \(\tau^2\).

The cost shows up in `Cand/draw` in `summary()`: the ING conditional is not
conjugate, so each draw needs several accept-reject candidates.

# Plug-ins and where the prior is calibrated

Before any iteration, \(\tau^2\) has to be set from the prior alone — the
ICM start and the \(\Sigma_{\text{ranef}}\) used by Block 1 both need a
number before there is any \(b\) to condition on:

\[
\tilde\tau^2_k = \frac{\mathrm{rate}}{\mathrm{shape} - 1}
\qquad (\text{requires } \mathrm{shape} > 1),
\]

computed by `.two_block_tau2_plug_in_from_pfamily()` and
`.two_block_tau2_plug_in_vector()`. The conservative rate bounds use a
*different* reference, `.two_block_tau2_ref_from_pfamily()`, which returns
\(\mathrm{rate}/\mathrm{shape}\) — deliberately smaller, since smaller
\(\tau^2\) means a slower chain.

`Prior_Setup_GLMM()` calibrates the prior so that the plug-in reproduces
the marginal hyper-regression center:

\[
\mathrm{shape}_k = \frac{n_0 + 1 + p_k}{2},
\qquad
\mathrm{rate}_k = \hat\tau^2_k\,\frac{n_0 + p_k - 1}{2},
\]

where \(\hat\tau^2_k\) is the marginal center from
`compute_gaussian_prior` applied to
\(b_{\cdot k} \sim N(W_k\gamma_k, \tau^2_k I_J)\) with
\(\gamma_k \sim N(\mu_k, \Sigma_k)\) — stored in
`pop.prior_list[[k]]$dispersion` — and \(n_0\) is the effective prior
sample size for dispersion, set by `pop.dispersion.pwt`. The classical
`VarCorr` plug-in is kept separately in `pop.dispersion.ref`.

By construction \(\mathrm{rate}_k / (\mathrm{shape}_k - 1) = \hat\tau^2_k\),
so the ING prior and the `dNormal` prior agree on the *center* and differ
only in whether the component is allowed to move. That is what makes the
`ptypes` switch a clean comparison: changing it does not silently change
the location of the prior.

```{r calibration}
data(sleepstudy, package = "lme4")
dat <- sleepstudy
dat$Days_c <- dat$Days - mean(dat$Days)
form <- Reaction ~ Days_c + (1 + Days_c || Subject)

ps <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01,
                       pop.dispersion.pwt = 0.2)
str(ps$pop.ing_prior)
```

`pop.ing_prior` is one list per component, each carrying the Gamma
`shape` / `rate` on the precision plus the truncation window.

```{r calibration-check}
ing <- ps$pop.ing_prior
round(data.frame(
  plug_in   = vapply(ing, function(p) p$rate / (p$shape - 1), numeric(1)),
  ref_bound = vapply(ing, function(p) p$rate / p$shape, numeric(1)),
  dNormal   = vapply(ps$pop.prior_list, function(p) p$dispersion, numeric(1))
), 4)
```

The plug-in matches the `dNormal` dispersion exactly; the bound reference
sits below it.

# The truncation window

Each ING component carries `disp_lower` and `disp_upper`, and the sampled
\(\tau^2_k\) is confined to that interval. The bounds are quantiles of the
**limiting posterior** of the variance component — not of the prior.

The distinction is the whole point. A Gamma prior on a precision puts mass
arbitrarily close to \(\tau^2 = 0\), and \(\tau^2 \to 0\) is the
degenerate limit in which every group coefficient is pinned to
\(\mathcal{W}_j\gamma\): the hierarchy collapses, Block 1 and Block 2
become perfectly coupled, and the rate \(\lambda^*\) tends to 1. Truncating
that region away is what keeps the chain geometrically ergodic, and the
argument mirrors the drift-condition reasoning of @JonesHobert2001.

The lower bound does double duty. Because \(\lambda^*\) is monotone
decreasing in \(\tau^2\) for this structure, evaluating the rate at
`disp_lower` gives a bound valid over the entire window:

\[
\sup_{\tau^2 \in [\,\text{disp\_lower},\, \text{disp\_upper}\,]}
   \lambda^*(\tau^2)
\;=\; \lambda^*(\text{disp\_lower}).
\]

That is conservative rather than correct-on-average, which is the right
direction for a certificate: the sweep count it produces is at least as
large as any \(\tau^2\) actually visited would require.

The upper bound is less load-bearing but keeps the envelope construction
well conditioned and prevents a component from wandering into a region the
data cannot support. `inst/ING_TRUNCATION_WINDOW.md` gives the quantile
choices and the derivation.

# What the chain looks like

```{r fit, results = "hide"}
pf_ing <- pfamily_list(ps, ptypes = "dIndependent_Normal_Gamma")
design <- model_setup(form, data = dat)

set.seed(1)
fit <- rlmerb(n = 200L, design = design,
              pfamily_list   = pf_ing,
              dispprior_list = list(dispersion = ps$group.dispersion),
              progbar = FALSE, verbose = FALSE, print_icm_table = FALSE)
```

```{r fit-fields}
c(engine = fit$convergence_info$sim_method_used,
  m      = fit$convergence_info$m_convergence)
names(fit$sweep_history)
round(colMeans(fit$popef.dispersion), 3)
```

Two structural changes relative to Chapter B04. `sweep_history` now has a
`pilot` element as well as `main`, and `popef.dispersion` holds draws
rather than a repeated constant.

# Why a pilot stage becomes necessary

The Theorem 3 bound of Chapter B02 has two terms: one for the distance of
the starting *mean* from the target mean, and one for the variance. Chapter
B04's route sets the first to exactly zero by starting every chain at the
exact posterior mean.

With an ING component there is no exact mean to start from. The joint
target over \((\gamma, \tau^2, b)\) is not Gaussian, so the ICM start is a
conditional mode, and mode \(\neq\) mean. The pilot stage produces a start
point whose distance from the truth can be *bounded* instead: averaging
`n_pilot` independent pilot chains gives a start whose sampling covariance
is \(\Sigma / n_{\text{pilot}}\), so at confidence level \(\alpha\)

\[
D_0 = \frac{\chi^2_{p,\alpha}}{n_{\text{pilot}}}.
\]

`vignette("Chapter-B03")` works through the optimization that trades pilot
sweeps against main sweeps.

# Selecting components individually

`ptypes` accepts a single value or a named vector, so components can be
mixed:

```{r mixed-ptypes}
pf_mixed <- pfamily_list(ps, ptypes = c(
  "(Intercept)" = "dNormal",
  "Days_c"      = "dIndependent_Normal_Gamma"
))
vapply(pf_mixed, function(p) class(p)[1L], character(1))
```

Because Block 2 factorizes across components, this is coherent: the
intercept's \(\tau^2\) stays fixed while the slope's is sampled. The route
key is still `lmm_fixed_estimated` — `pf_summary$all_dNormal` is `FALSE`
as soon as *any* component is ING — so the pilot stage and the conservative
rate apply to the whole model. There is no partial-exactness discount.

# See also

- `vignette("Chapter-B03")` — pilot sizing and the cost optimization.
- `vignette("Chapter-B04")` — the exact route this one generalizes.
- `vignette("Chapter-B06")` — the extended rate diagnostic for ING models.
- `vignette("Chapter-07", package = "lmebayes")` — the user-facing view.
- `inst/TAU2_ING_FORMULAS.md`, `inst/ING_TRUNCATION_WINDOW.md`,
  `inst/BLOCK_ING_RINDEPNORMALGAMMA_REG.md` in the package sources.
