---
title: "Chapter C02: The exact Gaussian iid sampler"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter C02: The exact Gaussian iid sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(lmebayesCore)
```

```{r load-stored}
C02 <- readRDS(system.file(
  "extdata", "Chapter-C02-bwc-iid.rds", package = "lmebayesCore"
))
meta <- C02$meta
```

This chapter fits the model of `Chapter-C01` under the simplest prior
specification the package offers, and shows that the resulting posterior
can be sampled exactly rather than iteratively.

# Prior specification

Two quantities are held constant throughout this chapter: the observation
dispersion \(\sigma^2\) and the between-group variance matrix \(\Psi\).
Both are fixed at the defaults `Prior_Setup_GLMM()` calibrates from the
data, so neither carries a prior and neither is sampled:

```{r fixed}
C02$summary$tau2_prior_overview[, c("Prior", "E[tau2]")]
```

The three `dNormal` rows are the components of \(\Psi\), declared known.
The `Residual` row is \(\sigma^2\), likewise fixed.

That leaves the population effects \(\gamma\) as the only quantity with a
prior of its own. They receive the default multivariate normal prior,
specified separately for each group-effect component, so the intercept
block, the `distracted_ppvt` block and the `distracted_a1` block each get
their own mean vector and covariance matrix:

```{r pop-prior}
round(C02$summary$popef_prior_overview[, c("Prior Mean", "Prior.sd")], 3)
```

The pattern in the mean column is worth noting. Every coefficient is
centered at zero except the grand intercept, so the prior mean vector
describes a null, intercept-only model: no covariate has any effect, and no
group effect varies with anything at level two. The prior therefore carries
the same interpretation the null model does in classical statistics, and
the analysis can be read as asking how far the data move each coefficient
away from that null. What differs is that here the null is a starting point
that the data revise rather than a hypothesis that is accepted or rejected.

This is the same default the **glmbayes** family of packages uses for
ordinary iid regression: a multivariate normal prior on the regression
coefficients, centered at the null model and calibrated from the data
rather than specified by hand. What the mixed model adds is that the
specification is repeated once per group effect.

The setup is built by two calls. `Prior_Setup_GLMM()` performs the
calibration, and `pfamily_list()` wraps the result as the Block 2 prior
families the sampler consumes:

```{r prior-calls, eval = FALSE}
ps <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01)
pf <- pfamily_list(ps)
```

`pop.pwt = 0.01` gives the prior on \(\gamma\) the weight of about one
percent of the data. Because every component of `pf` comes back as
`dNormal`, the variance components are known, which is what the first table
above reports.

# Estimating the model using the default settings

The sampler takes the design and the two prior objects directly:

```{r call, eval = FALSE}
fit <- rlmerb(
  n              = 1000,
  design         = model_setup(form, data = dat),
  pfamily_list   = pf,
  dispprior_list = list(dispersion = ps$group.dispersion),
  sim_method     = "DEFAULT"
)
```

The `design` argument carries the data and both design matrices,
`pfamily_list` supplies the Block 2 priors from the previous section, and
`dispprior_list` supplies the fixed \(\sigma^2\). Passing a bare
`list(dispersion = ...)` rather than a `dGamma()` prior is what declares
that dispersion known.

Under this specification the joint posterior over the group effects and the
population effects is exactly multivariate normal, which admits a direct
sampling scheme with no Markov chain. `sim_method = "DEFAULT"` selects it.

The same model is fitted again later by a two-block Gibbs sampler. Running
both against one posterior is deliberate: the answer here is exact, so it
serves as the standard against which the iterative sampler is judged.

# Properties of the posterior and the iid sampling procedure

Collect the group effects as \(\beta = (\beta_1', \dots, \beta_J')'\). Up
to an additive constant, the log posterior splits into the three pieces of
the model:

\[
-2\log \pi(\gamma, \beta \mid y) =
\underbrace{\frac{1}{\sigma^2}\sum_{j} \left\|y_j - D_j\beta_j\right\|^2}_{\text{(i) log-likelihood}}
+ \underbrace{\sum_{j} (\beta_j - \mathcal{W}_j\gamma)' \Psi^{-1} (\beta_j - \mathcal{W}_j\gamma)}_{\text{(ii) prior on the group effects}}
+ \underbrace{(\gamma - \mu_0)' V^{-1} (\gamma - \mu_0)}_{\text{(iii) prior on the population effects}}.
\]

Because \(\sigma^2\), \(\Psi\) and \(V\) are all known, every term is
quadratic in \((\gamma, \beta)\), and a quadratic log density is a normal
density. The posterior is therefore multivariate normal, and it can be
factorized into a marginal for \(\gamma\) and a conditional for \(\beta\).

## Marginal distribution of the population effects

Integrating the group effects out of (i) and (ii) is a linear-Gaussian
calculation. Substituting \(\beta_j = \mathcal{W}_j\gamma + u_j\) with
\(u_j \sim N(0, \Psi)\) gives

\[
y_j \mid \gamma \sim N\!\left(D_j\mathcal{W}_j\,\gamma,\; V_j\right),
\qquad V_j = \sigma^2 I_{n_j} + D_j \Psi D_j',
\]

so each group contributes a Gaussian likelihood in \(\gamma\) with design
\(X_j = D_j\mathcal{W}_j\). Combining these with (iii) gives the marginal
posterior

\[
\gamma \mid y \sim N\!\left(m_\gamma,\; P_\gamma^{-1}\right),
\qquad
\begin{aligned}
P_\gamma &= V^{-1} + \sum_j X_j' V_j^{-1} X_j, \\
m_\gamma &= P_\gamma^{-1}\Big(V^{-1}\mu_0 + \sum_j X_j' V_j^{-1} y_j\Big).
\end{aligned}
\]

The term \(D_j \Psi D_j'\) in \(V_j\) is what distinguishes this from
ordinary regression: it inflates the variance of the groups that would
otherwise be most influential, which is how the model discounts a group
that is unusual rather than merely large.

## Conditional distribution of the group effects

Given \(\gamma\), terms (i) and (ii) separate across groups, so the group
effects are conditionally independent — each is an ordinary conjugate
normal update, combining that group's own data with the population
prediction \(\mathcal{W}_j\gamma\):

\[
\beta_j \mid \gamma, y \sim N\!\left(b_j,\; \Sigma_j\right),
\qquad
\begin{aligned}
\Sigma_j &= \left(\sigma^{-2} D_j' D_j + \Psi^{-1}\right)^{-1}, \\
b_j &= \Sigma_j\left(\sigma^{-2} D_j' y_j + \Psi^{-1}\mathcal{W}_j\gamma\right).
\end{aligned}
\]

The precision weights in \(b_j\) are the shrinkage: a group with plenty of
data has a large \(D_j' D_j\) and stays near its own fit, while a small
group is pulled toward \(\mathcal{W}_j\gamma\).

## Drawing from the posterior

The two results above factorize the posterior exactly,

\[
\pi(\gamma, \beta \mid y) = \pi(\gamma \mid y)\;\prod_j \pi(\beta_j \mid \gamma, y),
\]

so one draw from the joint posterior is obtained in two steps: draw
\(\gamma\) from its marginal, then draw each \(\beta_j\) from its
conditional given that value. Repeating the pair produces independent
draws from the target itself — no burn-in to discard, no autocorrelation,
and no convergence rate to bound. It also makes the ICM point estimate the
exact posterior mean, since the mode and mean of a normal distribution
coincide.

# Summarizing the output

`summary()` collects the draws into tables:

```{r summary-call, eval = FALSE}
summary(fit, groups = c("30", "15"))
```

The output is organized by block, following the structure of the model.
After the call and a header giving the number of draws, the engine used and
the data dimensions, it reports in turn:

- **Random effects** — the reference `lmer` variance components, for
  orientation.
- **Block 2 variance components** — the \(\tau^2_k\) and the residual
  \(\sigma^2\), with their prior family. Here everything is fixed, so the
  posterior tables are suppressed and a note says so.
- **Block 2 population effects** — three tables for \(\gamma\): the prior
  means and standard deviations alongside the `lmer` reference, then the
  posterior mode, mean, standard deviation and Monte Carlo error, then
  distribution percentiles. Rows are named `component::parameter`,
  following the indexing described in `Chapter-C01`.
- **Block 1 group effects** — the distribution of the per-group posterior
  modes across all `r meta$J` groups, followed by per-group detail for any
  levels named in `groups`.

The `groups` argument is optional; without it the per-group detail is
replaced by a note pointing at `fit$groupef.mode`. The two levels requested
here are the largest and smallest schools in the data.

Two features of the population effect tables follow from the default
`pop.pwt = 0.01`. Because the prior is given the weight of only about one
percent of the data, it barely competes with the likelihood: the posterior
means land close to the `lmer` estimates, displaced slightly toward the
prior means of the previous section. Since those prior means are the null
model, that displacement is a mild shrinkage toward zero for every
coefficient but the intercept.

Read on the scale that matters, the two fits agree throughout. Every
Bayesian estimate falls within a fifth of a standard error of its `lmer`
counterpart, and most within a twentieth, so no coefficient in the table
supports a different reading under one fit than the other. Exact agreement
was never expected: the prior carries a small amount of weight, and this
chapter fixes the variance components where `lmer` estimates them, so the
two are answering slightly different questions. The differences that remain
are too small to be worth interpreting individually.

The `Pr(Prior_tail)` column has a related reading. For every row it measures
the same thing: the posterior mass falling on the opposite side of the prior
mean from the posterior estimate. It is thus a one-sided statement about
how far the data have moved a coefficient away from its prior value, and
how much posterior belief remains on the other side of it.

Because the prior means are the null model, that reference point is zero
for every coefficient except the intercept, so for those rows the column
answers the same question a one-sided p-value does. Across them it stays in
the same range as half the two-sided p-value the corresponding `lmer` fit
would report, typically running somewhat smaller because the posterior
standard deviations are not identical to the `lmer` standard errors, and
the ordering it induces matches the classical one. The intercept is read the
same way, but against the calibrated prior mean rather than the origin, so
it is not comparable to a test against zero.

The output below was produced by `data-raw/make_Chapter-C02.R` with
`n = `r meta$n_draws`` and `set.seed(`r meta$seed`)`, then stored, so that
building this vignette does not rerun the sampler.

```{r summary-print}
C02$summary
```

# Appendix: the reference `lmer` fit

The `lmer` columns used throughout this chapter come from a single REML fit
that `model_setup()` performs on all observations. `Prior_Setup_GLMM()`
returns it as `mer_fit`, and the same object is stored on the design, so
either of these reaches it:

```{r lmer-access, eval = FALSE}
ps$mer_fit
design$lmer
```

It serves two purposes: it calibrates the default priors of the first
section, and it populates the reference columns in the summary tables. Its
own summary is:

```{r lmer-summary}
cat(C02$lmer_summary, sep = "\n")
```

Two points connect this back to the chapter. First, the fixed effects are
reported in `lmer`'s flat parameterization, with the cross-level term
written as `free_reduced_lunch:distracted_a1`. The summary tables carry the
same seven coefficients but regroup them as `component::parameter`, so that
interaction appears as `distracted_a1::free_reduced_lunch` — a level-two
predictor of the `distracted_a1` slope. `Chapter-C01` covers the mapping.

Second, the `Random effects` block is the quantity this chapter holds
fixed, and it is worth seeing that it is not fixed at these values:

```{r vc-compare}
round(C02$summary$tau2_prior_overview[, c("E[tau2]", "lmer")], 2)
```

The residual variance is calibrated close to the REML value, and the
intercept variance is moderately tighter, but the two slope components are
held near a quarter of what `lmer` estimates. This is the one place where
the two fits are far apart, and it is a difference in specification rather
than in estimation: a smaller \(\Psi\) shrinks the group effects harder
toward their population predictions, so the group effects reported in the
previous section are more tightly pooled than `lmer` would make them.

That the population effects nonetheless agree so closely is worth noting.
Holding \(\Psi\) at these values costs little there, which is what makes
this specification a reasonable starting point even though it is not the
one `lmer` fits. The later chapters estimate the variance components rather
than fixing them, and are the place to look for what changes when the
`Random effects` block above is treated as unknown.

# See also

- `vignette("Chapter-C01")` — the data, model and design used here.
- `vignette("Chapter-B04")` — route selection and the compiled kernel
  behind the exact engine.
- `?rlmerb`, `?summary.rlmerb`, `?model_setup`, `?Prior_Setup_GLMM`,
  `?pfamily_list`
- `data-raw/make_Chapter-C02.R` in the package sources for the script that
  produced the stored results.
