---
title: "Chapter B10: Sweep history and convergence plots"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B10: Sweep history and convergence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B10: Sweep history and convergence plots

The sweep history is the engine's audit trail. It records, after every
inner Gibbs sweep, what the *ensemble* of replicate chains looked like —
which is a genuinely different object from the trace of a single chain, and
supports a different kind of inspection.

# Why the ensemble view is the right one

A conventional MCMC trace plot shows one chain's value against iteration.
It is a diagnostic of last resort: you look for a trend, a level shift, or
a sticky patch, and you infer that the burn-in was or was not long enough.

This sampler runs `n` *independent replicate* chains, each of exactly
`m_convergence` sweeps, all started from the same point (Chapter B01).
That means at sweep \(m\) there are `n` independent draws from the
distribution of the chain after \(m\) sweeps — call it \(\pi_m\). Their
mean and variance estimate the mean and variance of \(\pi_m\) directly, and
the convergence claim is a statement about how close \(\pi_m\) is to
\(\pi\).

So the sweep history is not a trace. It is a sequence of snapshots of the
*distribution* being approached, and it maps onto the two terms of the
Theorem 3 bound one-for-one: a mean-error term and a variance term.

# The object

```{r fit, results = "hide"}
data(sleepstudy, package = "lme4")
dat <- sleepstudy
dat$Days_c <- dat$Days - mean(dat$Days)
form <- Reaction ~ Days_c + (1 + Days_c || Subject)

design <- model_setup(form, data = dat)
ps     <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01)

set.seed(1)
fit <- rlmerb(
  n              = 200L,
  design         = design,
  pfamily_list   = pfamily_list(ps),
  dispprior_list = list(dispersion = ps$group.dispersion),
  sim_method     = "TWO_BLOCK_GIBBS",
  progbar = FALSE, verbose = FALSE, print_icm_table = FALSE
)
```

```{r history-structure}
class(fit$sweep_history)
names(fit$sweep_history)
fit$sweep_history$stage
```

**Two API shapes, one object.** At this matrix level `fit$sweep_history`
*is* the history, tagged by `$stage`, and pilot information lives
separately in `fit$pilot`. The formula-level `lmerb()` and `glmerb()` in
**lmebayes** repackage the two stages into a list, so there you write
`fit$sweep_history$main` and `fit$sweep_history$pilot`. Everything below
applies to either, once you have a single-stage object in hand.

```{r history-print}
fit$sweep_history
```

Each Block 2 parameter contributes one `mode` row and one row per sweep.
The `mode` row is the ICM point estimate — the value every chain started
from. The sweep rows carry the cross-chain `mean` and `sd`.

The components are worth knowing individually:

| Field | Contents |
|---|---|
| `stage` | `"main"` or `"pilot"` |
| `n_sweeps` | Sweeps recorded, equal to `m_convergence` for the stage |
| `fixef_mode` | The ICM start point, per component |
| `table` | Long data frame: `re_component`, `covariate`, `sweep`, `mean`, `sd` |
| `cov_by_sweep` | Full cross-chain covariance matrix at each sweep |
| `coef_index` | Component/covariate key for the rows of `cov_by_sweep` |
| `disp_table` | Same shape as `table`, for sampled precisions \(1/\tau^2_k\); empty when they are fixed |

`cov_by_sweep` is the one that carries information the printed table does
not: the *cross-parameter* covariance at each sweep. The Theorem 3 bound is
a statement about the whole covariance, not one coordinate at a time, so
this is the object to reach for when a per-parameter view looks fine but
the certificate seems too optimistic.

# Which engines populate it

| Route | `pilot` | `main` | Note |
|---|---|---|---|
| `lmm_fixed_known`, `sim_method = "DEFAULT"` | — | — | No sweeps, so no history |
| `lmm_fixed_known`, `sim_method = "TWO_BLOCK_GIBBS"` | — | yes | Start is the exact mean |
| `lmm_fixed_estimated` (ING) | yes | yes | B05 |
| `lmm_gamma_*` (sampled \(\sigma^2\)) | yes | yes | B07 |
| `glmm_*` (non-Gaussian) | yes | yes | B09 Block 1 |

The exact-iid engine produces no history at all, and that is correct rather
than a gap: there are no sweeps to record. If you want a history on an
exactly-Gaussian model — to validate the sweep kernel against a known
answer — force `sim_method = "TWO_BLOCK_GIBBS"`, as above.

# Reading the two plots

Both plotting functions take a **sweep-history object**, not a fit, and
both draw base-graphics panels split by intercept-type and slope-type
predictors.

## `plot_mean_convergence()`

```{r plot-mean, fig.alt = "Standardized cross-chain mean by inner sweep"}
plot_mean_convergence(fit$sweep_history)
```

The plotted quantity is

\[
\frac{\bar{x}_m - \bar{x}_{\text{final}}}{\mathrm{sd}_m},
\]

the cross-chain mean at sweep \(m\), centered at the final sweep's mean and
standardized. This is the **Claim 1** quantity: the mean-error term of the
convergence bound.

On this route the line should sit at zero throughout, because every chain
starts at the exact posterior mean and the Gibbs kernel preserves it. A
visible drift means the start point is not what the theory assumes — the
usual causes are a mis-specified `prior_list_block1` or an ICM that did not
converge. On a pilot history, by contrast, drift is *expected*: the pilot
starts at a mode, not a mean, and its whole purpose is to move.

## `plot_var_convergence()`

```{r plot-var, fig.alt = "Ratio of cross-chain variance to final variance by inner sweep"}
plot_var_convergence(fit$sweep_history)
```

The plotted quantity is

\[
\frac{\mathrm{Var}_m}{\mathrm{Var}_{\text{final}}},
\]

the **Claim 3** quantity: the variance term of the bound.

Chains that all start at the same point have zero ensemble spread at sweep
0 and *grow* into the posterior variance, so the interesting direction is
approach to 1 from below. For this sampler that makes the variance plot the
more informative of the two — it is the term that is actually doing work
in the certificate, since the mean term was zeroed by construction.

A ratio still well below 1 at the final sweep means `m_convergence` was
computed from a rate that understates the coupling. On a Gaussian route
that should not happen (the rate is exact); on a GLMM route, where the rate
is a local-Gaussian heuristic, it is the diagnostic that would catch it.

# Multiple chains and `n_chains`

Both functions accept the arguments the underlying history carries. When a
route was run with more than one chain group, pass `n_chains` so the
functions can reshape the stored summaries correctly:

```{r n-chains, eval = FALSE}
plot_mean_convergence(fit$sweep_history, n_chains = 4L)
```

Without it the summaries are treated as a single ensemble, which is right
for the default configuration and wrong when chains were grouped.

# Using the history programmatically

The object is a plain list underneath its class, so the summaries can be
extracted rather than plotted:

```{r programmatic}
h <- fit$sweep_history
head(h$table, 8)
h$cov_by_sweep[[h$n_sweeps]]
```

A minimal automated check — the one worth running in a test suite — is that
the final-sweep mean agrees with the ICM mode to within Monte Carlo error,
and that the variance ratio has reached 1:

```{r check}
gm <- fit$popef.mode[["Days_c"]][[1]]
dr <- fit$popef[["Days_c"]][, 1]
z  <- (mean(dr) - gm) / (sd(dr) / sqrt(length(dr)))
round(c(icm = gm, draws = mean(dr), z = z), 4)
```

# Interpreting a pilot history

```{r pilot, results = "hide"}
pf_ing <- pfamily_list(
  Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01,
                   pop.dispersion.pwt = 0.2),
  ptypes = "dIndependent_Normal_Gamma"
)
set.seed(1)
fit_ing <- rlmerb(
  n = 200L, design = design,
  pfamily_list   = pf_ing,
  dispprior_list = list(dispersion = ps$group.dispersion),
  progbar = FALSE, verbose = FALSE, print_icm_table = FALSE
)
```

```{r main-print}
fit_ing$sweep_history
```

Two things changed. There is now a second block, `Block 2 RE precision`,
tracking the sampled \(1/\tau^2_k\) across sweeps — that is `disp_table`,
which was empty on the fixed-\(\tau^2\) fit above. And the pilot stage has
run, with its sizing recorded in `convergence_info`:

```{r pilot-info}
ci <- fit_ing$convergence_info
c(n_pilot        = ci$n_pilot,
  n_pilot_source = ci$n_pilot_source,
  m_pilot        = ci$m_convergence_pilot,
  m_main         = ci$m_convergence,
  lambda_star    = round(ci$lambda_star, 4),
  method         = ci$method)
```

`n_pilot_source = "cost"` means the pilot size came from the cost
optimization of `vignette("Chapter-B03")` rather than from the
`gap_tol` fallback. `method = "exact+disp_lower_bound"` records that the
rate is exact in its Gaussian part but plugged in at `disp_lower` for the
sampled variance component — the conservatism described in
`vignette("Chapter-B05")`.

The interpretive contrast is the point. A pilot history starts at the ICM
*mode* and its cross-chain mean moves as the chains find the posterior
mean. The main history starts at the pilot's average and barely moves —
which is exactly the condition under which the
\(D_0 = \chi^2_{p,\alpha}/n_{\text{pilot}}\) term is small enough to
certify a short main stage.

If the pilot history has *not* flattened by its last sweep, `n_pilot` or
the pilot sweep count was too small, and the main-stage certificate rests
on a start-distance bound that the pilot did not actually achieve. That is
the single most useful thing to look at when a fit's answers seem off.

# See also

- `vignette("Chapter-B01")` — replicate chains and why the ensemble view
  applies.
- `vignette("Chapter-B02")` — Claim 1 and Claim 3, the two bound terms
  these plots correspond to.
- `vignette("Chapter-B03")` — pilot sizing.
- `vignette("Chapter-06", package = "lmebayes")` — the user-facing version.
- `?plot_mean_convergence`, `?plot_var_convergence`
