---
title: "Chapter B01: The two-block Gibbs architecture"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B01: The two-block Gibbs architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B01: The two-block Gibbs architecture

`Chapter-B00` introduced the two-stage model. This chapter follows one
draw all the way down: from the user-facing call, through the replicate
chain driver and the inner sweep loop, into the block updates and the
`glmbayesCore` iid samplers where the randomness actually happens.

# Why two blocks

The parameter vector splits naturally into

- **Block 1** — the group-level coefficients \(\beta_1, \dots, \beta_J\)
  (dimension \(J \cdot P\)), and
- **Block 2** — the population coefficients \(\gamma\) (dimension \(q\)),
  together with any sampled variance components \(\tau^2_k\).

Given \(\gamma\) and \(\Psi\), the \(\beta_j\) are **conditionally
independent across groups**: group \(j\)'s conditional depends only on
\(y_j\), \(D_j\), and \(\mathcal{W}_j\gamma\). Given all \(\beta_j\), the
Block 2 conditional is a plain normal regression of the \(\beta\)'s on the
level-2 designs, and because \(\Psi\) is diagonal it factorizes further
across the \(P\) coefficients.

So a "two-block" sweep is really \(J + P\) small independent draws,
never a \(J P + q\) dimensional one. The Block 1 precision matrix, which
would be \(J p_{re} \times J p_{re}\), is never formed.

The price is autocorrelation between successive sweeps. The whole of
`Chapter-B02` and `Chapter-B03` is about pricing that autocorrelation
exactly, so the sampler can decide how many sweeps to discard between
stored draws.

# One sweep

```
for each inner sweep m = 1, ..., m_convergence:

  Block 1:  for j = 1, ..., J
              beta_j  ~  p(beta_j | y_j, gamma, Psi, sigma2_j)

  Block 2:  for k = 1, ..., P
              gamma_k ~  p(gamma_k | beta_.k, X_k, tau2_k)
              [ING only: also draw tau2_k, refresh Psi]
```

Two conventions matter downstream:

1. **Block 2 is updated second.** The total-variation theory of
   `Chapter-B02` bounds the distance for the block updated *second* in each
   sweep. The stored Block 1 draw therefore lags by a half-step, which is
   why the sweep calibration adds one extra sweep.
2. **The sampled \(\tau^2_k\) feeds back into Block 1.** When a Block 2
   component is `dIndependent_Normal_Gamma`, the drawn \(\tau^2_k\)
   replaces the corresponding diagonal entry of \(\Psi\) before the next
   Block 1 update.

# Replicate chains, not one long chain

`lmebayesCore` does **not** produce one long correlated chain and thin it.
It runs `n` independent replicate chains, each started at the exact joint
posterior mean (Gaussian case) or the posterior mode (GLMM case), runs
each for `m_convergence` inner sweeps, and stores **one** draw per chain.

This is the same output contract as the iid samplers in `glmbayesCore`:
the `n` stored draws are mutually independent, so ordinary Monte Carlo
standard errors apply and no effective-sample-size correction is needed.
The cost is that every stored draw pays the full `m_convergence` burn-in;
the benefit is that `m_convergence` is small (typically single digits)
because the chain starts at the mean and the TV bound is explicit.

Starting every chain at the posterior mean also kills the mean term of
the convergence bound: \(D_0 = 0\), so only the variance-convergence sum
remains. See `Chapter-B03`.

# The call stack

```
lmerb() / glmerb()                        [lmebayes: formula interface]
  |
  +-- model_setup()                       formula -> y, Z, group, X_hyper
  +-- Prior_Setup_GLMM()                  reference lmer/glmer -> prior scales
  +-- pfamily_list() / dGamma_list()      Block 2 and dispersion priors
  |
  +-- rlmerb() / rglmerb()                [lmebayesCore: matrix level]
        |
        +-- matrix_args_lmm()             design + priors -> route arguments
        +-- lmerb_posterior_mean()        exact Gaussian mean (ICM), or
        |   glmerb_posterior_mode()       GLMM mode -- also the chain start
        |
        +-- route table
              |
              +-- rLMMNormal_reg_known_vcov_iid()      exact iid, no sweeps
              +-- rLMMNormal_reg_known_vcov_two_bg()   two-block Gibbs
              +-- rLMMindepNormalGamma_reg_*()         ING sigma2 routes
              +-- rGLMM_reg_known_vcov()               non-Gaussian, fixed tau2
              +-- rGLMM_reg_estimated_vcov()           non-Gaussian, ING tau2
                    |
                    +-- rGLMM_sweep()     outer sweep loop (pilot and main)
                          |
                          +-- Block 1 across all chains
                          +-- Block 2 across all chains
```

`rlmerb()` and `rglmerb()` are the matrix-level entry points: they take a
`design` object plus prior objects rather than a formula, and they are what
`lmebayes` calls. Everything above them is formula parsing; everything
below is sampling.

## Route selection

Which route runs is decided by two independent questions.

| Measurement dispersion \(\sigma^2\) | Variance components \(\tau^2_k\) | Route |
|---|---|---|
| fixed scalar or fixed per-group vector | all `dNormal` (known) | `rLMMNormal_reg_known_vcov` |
| fixed | any `dIndependent_Normal_Gamma` | `rLMMNormal_reg_estimated_vcov` |
| `dGamma()` / `dGamma_list()` (sampled) | all `dNormal` | `rLMMindepNormalGamma_reg_known_vcov` |
| `dGamma()` / `dGamma_list()` (sampled) | any ING | `rLMMindepNormalGamma_reg_estimated_vcov` |

For non-Gaussian families the same two questions select between
`rGLMM_reg_known_vcov()` and `rGLMM_reg_estimated_vcov()`. The two GLMM
routes differ in **how conservative the eigenvalue bound must be**, not in
whether a pilot stage runs: non-Gaussian models always run a pilot unless
`n_pilot = 0`.

The special case in the first row deserves a name of its own. When
\(\sigma^2\) is fixed *and* every \(\tau^2_k\) is known, the joint
posterior over \((\beta, \gamma)\) is **exactly multivariate normal**, and
`sim_method = "DEFAULT"` draws from it directly — iid, no sweeps, no
burn-in, no autocorrelation. `sim_method = "TWO_BLOCK_GIBBS"` forces the
Gibbs sampler on the same target, which is how the Gibbs path is validated
against ground truth. `Chapter-B04` works through this route.

# Sweep-outer versus chain-outer

Two loop orders exist in the codebase.

**Sweep-outer** (`rGLMM_sweep()`, the reference implementation): for each
inner sweep `m`, *all* chains do Block 1, then *all* chains do Block 2.

```r
for (m in seq_len(inner_sweeps)) {
  batch <- block1_all_chains(batch, ...)
  batch <- block2_all_chains(batch, ...)
  sweep_stats[[m]] <- snapshot_fixef_stats(batch)
}
```

**Chain-outer** (the older C++ driver): each chain runs all of its sweeps
before the next chain starts.

Sweep-outer is preferred for three reasons. It exposes the natural
parallelism — at fixed `m` the chains are independent within a block, so
Block 1 across chains is embarrassingly parallel. It amortizes per-sweep
setup (family functions, prior payloads) across chains rather than
rebuilding it per chain. And it makes `sweep_history` meaningful:
cross-chain means and variances at sweep `m` are a genuine snapshot of the
sampler's distribution after `m` sweeps, which is exactly what
`plot_mean_convergence()` and `plot_var_convergence()` plot
(`Chapter-B10`).

Some Gaussian fixed-\(\sigma^2\), fixed-\(\tau^2\) paths still use the
chain-outer driver and therefore do not populate `sweep_history`; that
gap is tracked in the package README.

# Inside Block 1

Block 1 is split into a **prep** phase and a **draw** phase.

Prep assembles, per chain: the current Block 2 state \(\gamma\), the
implied observation-level prior means \(\mathcal{W}_j\gamma\) (built by
`build_mu_all()`), and the Block 1 prior precision \(\Psi^{-1}\) —
refreshed from the current \(\tau^2\) draw when Block 2 is ING.

Draw then calls, once per chain:

- `rNormal_reg_group()` for Gaussian responses — a conjugate normal draw
  per group, and
- `rNormalGLM_reg_group()` for Poisson, binomial, and Gamma responses —
  the envelope accept-reject sampler per group, via `glmbayesCore`.

Both are **row-block** engines: they take the stacked design, a group
index, and per-group priors, and return one draw per group. The same
engines back `lmebayes`'s `lmbBlock()` / `glmbBlock()` interface, where the
row partition is user-specified rather than a grouping factor
(`Chapter-B09`).

# Inside Block 2

Block 2 does three things per chain.

1. **Align.** The Block 1 draws are stored in `group_levels` order; the
   level-2 design `X_hyper[[k]]` has its own row order. The response
   passed to the Block 2 regression must follow `rownames(X_hyper[[k]])`.
   Skipping this alignment is silently correct on datasets where the two
   orders coincide and silently wrong everywhere else, so it is done
   unconditionally.

2. **Draw \(\gamma_k\)** for each random-effect coefficient \(k\), by
   calling `glmbayesCore::rglmb()` with `pfamily_list[[k]]`:
   - `dNormal` — \(\tau^2_k\) is the known regression dispersion, and the
     draw is a conjugate multivariate normal.
   - `dIndependent_Normal_Gamma` — a Gamma prior sits on the precision
     \(1/\tau^2_k\), and the sampler makes a **joint**
     \((\gamma_k, \tau^2_k)\) draw through the likelihood-subgradient
     envelope [@Nygren2006; @GriffinBrown2010]. This is the route
     documented in `Chapter-B05`.

3. **Write back.** The new \(\gamma_k\) goes into the chain's fixed-effect
   state; a new \(\tau^2_k\) (ING only) goes into \(\Psi\) for the next
   Block 1 prep.

The response for the Block 2 regression of coefficient \(k\) is the vector
of \(J\) Block 1 draws for that coefficient. So Block 2 sees \(J\)
"observations" — one per group — regardless of how many rows the original
data has. This is why the prior-versus-data guard for ING components
requires the effective prior sample size to be at most \(J\).

# Where C++ takes over

| Layer | Implementation |
|---|---|
| Route selection, pilot planning, output assembly | R |
| Outer sweep loop | R (`rGLMM_sweep()`); C++ on the legacy driver |
| Chain loops within a block | R today; C++ migration planned |
| `build_mu_all()`, Block 1 prior refresh, b-alignment | C++ by default |
| Per-group Block 1 draw | C++ (`block_rNormalGLM_cpp`) |
| Envelope construction and accept-reject | C++, in `glmbayesCore` |

The compiled entry points resolve their R-level callbacks through the
`glmbayesCore` namespace rather than a local copy, so there is exactly one
implementation of the envelope machinery in the ecosystem even though two
packages link against it.

Optional OpenCL offload for envelope construction is inherited from the
same layer; see `vignette("Chapter-A10", package = "glmbayes")`.

# What one sweep stores

| Quantity | After `inner_sweeps = 1` |
|---|---|
| `fixef_draws` | Block 2 \(\gamma\), one row per chain |
| `coefficients` | Block 1 \(\beta\) from that sweep |
| `sweep_history` row 1 | cross-chain colMeans of the fixed effects |

With `inner_sweeps = 1` the first `sweep_history` row equals
`colMeans(fixef_draws)` exactly — a useful invariant when checking a new
driver against an old one. For parity work between R and C++ kernels,
compare cross-chain means at large `n` rather than expecting bit-identical
RNG streams: the two implementations draw from different streams.

# See also

- **Chapter B02** — why the sweep loop converges, and how fast.
- **Chapter B03** — how `m_convergence` is chosen.
- **Chapter B09** — the row-block engines used inside Block 1.
- **Chapter B10** — reading `sweep_history`.
- `inst/ARCHITECTURE_glmerb.md` in the package sources, the maintainer-level
  file-by-file map this chapter summarizes.
