---
title: "Chapter B06: The extended local rate diagnostic"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B06: The extended local rate diagnostic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B06: The extended local rate diagnostic

Chapter B02 derived a convergence rate \(\lambda^*\) for a two-block Gibbs
sampler whose one-step map is *linear*. Chapter B05 introduced ING
components, which sample \(\tau^2_k\) and therefore break that linearity.
This appendix is about the gap between those two facts: what the engine
actually certifies when \(\tau^2\) moves, and what the extended diagnostic
adds on top.

The short version, because it is easy to get backwards:

- **Production `m_convergence` never uses the extended rate.** It uses the
  plain rate evaluated at the corners of the truncation window, which is a
  *proved* upper bound for the \((\gamma, \beta)\) sub-chain.
- **The extended rate is a local diagnostic.** It quantifies how much the
  coupling to \(\lambda\) and \(\Omega\) inflates the rate at one reference
  state. It is not a certificate and does not feed calibration.

# Why a single \(\lambda^*\) stops being enough

Under `dIndependent_Normal_Gamma`, Block 2's state is
\((\gamma, \lambda_{\mathrm{ING}})\) rather than \(\gamma\) alone, where
\(\lambda_p := (\Psi^{-1})_{pp} = 1/\tau^2_p\):

\[
\beta_{jp} \mid \gamma_p, \lambda_p \sim N(W_{pj}\gamma_p,\ 1/\lambda_p),
\quad
\gamma_p \sim N(\mu_p^0, V_p),
\quad
\lambda_p \sim \mathrm{Gamma}(a_p^0, r_p^0),
\]

with \(\gamma_p\) and \(\lambda_p\) independent a priori — the
"Independent" in ING.

Claims 1–3 and Theorem 3 of @Nygren2020 require the contraction matrix

\[
A = P_{11}^{-1/2}\,P_{12}\,P_{22}^{-1}\,P_{21}\,P_{11}^{-1/2}
\]

to be **state-independent**, which is what makes the Gibbs map exactly
linear and the TV bound exact. With \(\lambda\) sampled, two things go
wrong at once. First, \(\lambda\) appears inside \(P_{22}\) and
\(P_{11}^{(\gamma\gamma)}\), so \(A\) depends on the current draw. Second,
new cross-derivative blocks appear that involve the residuals
\(u_{jp} = \beta_{jp} - W_{pj}\gamma_p\), which are themselves random.

\(A(\gamma, \beta, \lambda)\) is therefore a **local Jacobian**, not a
global operator, and no certified TV bound carries over without
re-derivation.

# What the engine certifies instead

The recovery is a weaker but genuinely proved statement about a
*sub-chain*.

Freeze \((\lambda, \Omega)\) at fixed values and the original fixed-\(\Psi\)
theory applies verbatim to the \((\gamma, \beta)\) chain conditional on
those values. So it suffices to bound \(\lambda^*(\Lambda, \Omega)\)
uniformly over every \((\Lambda, \Omega)\) the truncated sampler can visit
— and the truncation window makes that set a compact box.

The monotonicity results do the rest. \(\lambda^*\) is non-decreasing in
each \(\lambda_p\) and Loewner-non-increasing in \(\Omega\), so the
supremum sits at a corner:

\[
\sup_{\substack{\lambda_p \in [\lambda_p^{\min}, \lambda_p^{\max}] \\
                \Omega \succeq \Omega_{\min}}}
\lambda^*(\Lambda, \Omega)
\;=\;
\lambda^*\!\left(\Lambda^{\max},\ \Omega_{\min}\right).
\]

Translating through \(\lambda_p = 1/\tau_p^2\) and
\(\Omega_j = 1/\sigma_j^2\), the corner is reached at

| Quantity | Corner | Window field |
|---|---|---|
| RE precision \(\lambda_p^{\max}\) | smallest \(\tau^2_p\) | `disp_lower` |
| Measurement precision \(\Omega_{\min}\) | largest \(\sigma^2_j\) | `disp_upper` |

which is exactly what `two_block_rate_from_pfamily_list()` implements: for
an ING component it substitutes `pl$disp_lower` as the dispersion, and
`.rLMM_measurement_disp_upper_for_rate()` supplies the \(\Omega\) side.

That is the whole justification for the "evaluate at `disp_lower`" rule
stated informally in Chapter B05. It is a proved corner of a proved
monotone function, not a rule of thumb.

**What it does not cover.** The bound certifies the \((\gamma, \beta)\)
sub-chain *conditional on* whatever \((\lambda, \Omega)\) the sampler
visits. It says nothing about how fast \(\lambda\) itself mixes, nor about
the fully joint \((\gamma, \beta, \lambda)\) chain. Closing that gap is
what the extended diagnostic is for.

# The extended rate

`two_block_rate_ing()` enlarges \(x_1\) from \(\gamma\) to
\((\gamma, \lambda_{\mathrm{ING}}, \Omega_{\mathrm{ING}})\), so its
dimension grows from \(q = \sum_k q_k\) to

\[
q_{\mathrm{ext}} = q + |\mathrm{ING}| + n_\Omega,
\qquad
n_\Omega \in \{0,\ 1,\ J\}
\]

according to whether the measurement dispersion is fixed, pooled, or
per-group.

## The new blocks

\(P_{22}\) keeps its form,
\(P_{22} = \mathrm{blockdiag}_j(D_j'\Omega_j D_j + \Psi^{-1})\), but
\(\Psi^{-1}\) and \(\Omega_j\) are now current state rather than constants.
The interesting additions are in \(P_{11}\), which becomes

\[
P_{11} =
\begin{pmatrix}
P_{11}^{(\gamma\gamma)} & P_{11}^{(\gamma\lambda)} & 0 \\
P_{11}^{(\gamma\lambda)\top} & P_{11}^{(\lambda\lambda)} & 0 \\
0 & 0 & P_{11}^{(\Omega\Omega)}
\end{pmatrix}
\]

| Block | Value | Origin |
|---|---|---|
| \(P_{11}^{(\gamma\gamma)}\) | \(\lambda_p W_p'W_p + V_p^{-1}\), block-diagonal in \(p\) | as before, but \(\lambda_p\) is state |
| \(P_{11}^{(\lambda\lambda)}\) | \(\mathrm{diag}_{p}\bigl((J/2 + a_p^0 - 1)/\lambda_p^2\bigr)\) | Gamma prior curvature |
| \(P_{11}^{(\gamma\lambda)}\) | column \(p\) is \(-W_p'u_p\) | residual coupling |
| \(P_{11}^{(\Omega\Omega)}\) | \((n_j/2 + a_j^0 - 1)/\Omega_j^2\) | measurement Gamma curvature |

The two zero blocks are exact: \(\Omega\) is Hessian-orthogonal to
\(\gamma\) and \(\lambda\). And \(P_{12}\) gains a row \(u_{j,\mathrm{ING}}'\)
per ING component and, when \(\Omega\) is sampled, a row
\(-[D_j'e_j]'\).

The extended rate is then
\(\lambda^*_{\mathrm{ext}} = \lambda_{\max}(A_{\mathrm{ext}})\) with
\(A_{\mathrm{ext}}\) built from the extended blocks.

When there are no ING components every new row and column vanishes and the
extended spectrum collapses to the plain one exactly — a useful sanity
check.

## The residuals only make it worse

One result is worth stating on its own, because it explains why the
extended quantity cannot be turned into a certificate by the same trick
that worked for \((\Lambda, \Omega)\). For fixed \(\Lambda, \Omega\),

\[
\lambda_{\max}(A(u)) \;\ge\; \lambda^*_{\mathrm{original}}(\Lambda, \Omega)
\quad \text{for all } u,
\]

with equality exactly at \(u = 0\). Nonzero residuals can only *raise* the
extended top eigenvalue.

So \(u = 0\) is the minimum, not the worst case, and freezing residuals at
zero would understate the extended rate. Unlike \(\lambda\) and \(\Omega\),
the residuals \(u_{jp}\) have no prior-enforced ceiling, so there is no
compact box to take a supremum over. That asymmetry is precisely why the
extended rate remains a diagnostic.

# Using it

`two_block_rate_ing()` is documented but not exported, so it needs the
triple-colon form. It takes the same design and prior arguments as
`two_block_rate()`, plus a reference state.

```{r data}
data(sleepstudy, package = "lme4")
dat <- sleepstudy
dat$Days_c <- dat$Days - mean(dat$Days)
form <- Reaction ~ Days_c + (1 + Days_c || Subject)

design <- model_setup(form, data = dat)
ps     <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01,
                           pop.dispersion.pwt = 0.2)
pf_ing <- pfamily_list(ps, ptypes = "dIndependent_Normal_Gamma")

prior1 <- list(Sigma = ps$group.Sigma, dispersion = ps$group.dispersion)
```

The certified rate first, for comparison:

```{r base-rate}
rate_base <- two_block_rate_from_pfamily_list(
  x                 = design$D,
  group             = design$group,
  x_hyper           = design$W,
  prior_list_block1 = prior1,
  pfamily_list      = pf_ing
)
rate_base$lambda_star
```

That is the corner bound, and the number that becomes `m_convergence`.

The extended version needs a reference state per ING component: a plug-in
precision \(\lambda_p\), the Gamma shape \(a_p^0\), and the RE-level
residuals \(u_{jp}\) named by group level. The residuals come from the
reference fit that `model_setup()` already ran:

```{r ing-u}
u_ref <- lmebayesCore:::.lmebayes_reference_u(
  design$lmer, design$group_name, design$groupef.names
)
head(u_ref, 4)
```

For a plug-in precision, the prior mean \(\lambda_p = (a_p^0 - 1)/r_p^0\)
is the natural "typical state" choice:

```{r ing-spec}
make_lambda_ing <- function(lambda_fun) {
  stats::setNames(
    lapply(names(ps$pop.ing_prior), function(k) {
      g <- ps$pop.ing_prior[[k]]
      list(lambda = lambda_fun(g),
           shape  = g$shape,
           u      = stats::setNames(u_ref[[k]], rownames(u_ref)))
    }),
    names(ps$pop.ing_prior)
  )
}

rate_ext <- lmebayesCore:::two_block_rate_ing(
  x                 = design$D,
  group             = design$group,
  x_hyper           = design$W,
  prior_list_block1 = prior1,
  prior_list_block2 = ps$pop.prior_list,
  lambda_ing        = make_lambda_ing(function(g) (g$shape - 1) / g$rate)
)
rate_ext
```

## Reading the output

```{r fields}
c(base     = rate_ext$lambda_star_base,
  extended = rate_ext$lambda_star,
  delta    = rate_ext$lambda_star - rate_ext$lambda_star_base)
rate_ext$ext_names
```

`lambda_star_base` is the plain rate with no coupling to \(\lambda\);
`lambda_star` is the extended one; the difference is what the print method
reports as `delta`. `ext_names` labels the enlarged coordinate system —
the \(\gamma\) names first, then `lambda::<component>`, then `Omega` or
`Omega::<group>` when the measurement precision is included.

Here the extended rate is roughly `r round(rate_ext$lambda_star /
rate_ext$lambda_star_base, 1)` times the base rate. Both are comfortably
below one, so the qualitative conclusion is unchanged: the joint chain
mixes more slowly than the certified sub-chain bound suggests, but it still
contracts. A large `delta` does not invalidate the certificate — it says
the certificate covers less of the story than you might have assumed.

## Two checks worth running

**The \(u = 0\) floor.** Setting every residual to zero should reproduce
the base rate exactly, since that is the equality case of the inequality
above:

```{r u-zero}
rate_u0 <- lmebayesCore:::two_block_rate_ing(
  x = design$D, group = design$group, x_hyper = design$W,
  prior_list_block1 = prior1, prior_list_block2 = ps$pop.prior_list,
  lambda_ing = stats::setNames(
    lapply(names(ps$pop.ing_prior), function(k) {
      g <- ps$pop.ing_prior[[k]]
      list(lambda = (g$shape - 1) / g$rate, shape = g$shape,
           u = stats::setNames(rep(0, nrow(u_ref)), rownames(u_ref)))
    }),
    names(ps$pop.ing_prior)
  ),
  warn_slow = FALSE
)
all.equal(rate_u0$lambda_star, rate_u0$lambda_star_base)
```

**The pessimistic corner.** Substituting the `disp_lower` corner
precision — the largest \(\lambda_p\) the truncation window allows — while
keeping the reference residuals gives a much harsher picture:

```{r corner, warning = FALSE}
rate_corner <- lmebayesCore:::two_block_rate_ing(
  x = design$D, group = design$group, x_hyper = design$W,
  prior_list_block1 = prior1, prior_list_block2 = ps$pop.prior_list,
  lambda_ing = make_lambda_ing(function(g) 1 / g$disp_lower),
  warn_slow = FALSE
)
rate_corner$lambda_star
```

This exceeds one. It is worth being precise about what that does and does
not mean. It is **not** a contradiction of `rate_base$lambda_star`, which
remains a valid bound for what it certifies. It is a local Hessian
evaluated at a deliberately mismatched state: the smallest \(\tau^2\) the
window permits, combined with residuals from a fit whose \(\tau^2\) was
several times larger. Such a state is reachable but atypical, and the
diagnostic is telling you that the joint chain has no contraction margin
*there*.

This is why `two_block_rate_ing()` uses a non-strict eigensolver and warns
rather than stopping. In `two_block_rate()`, \(\lambda^* \ge 1\) would
violate Remark 8 and indicate a computational bug, so it is an error. Here
it is a legitimate finding about one state.

# Scanning over draws

A single reference state is one sample of a state-dependent quantity. To
see the spread, `.two_block_rate_ing_over_draws()` mirrors the pilot's
`pmax` pattern over stored draws and returns `lambda_star_vec`,
`lambda_star_max`, `i_max`, `max_eigenvalues`, and `n_over_one` — the count
of draws whose extended rate reached one.

This is an empirical maximum over a finite sample, not a supremum, and its
usefulness is bounded by how well the draws explore the residual space. It
is the same species of safeguard as
`.two_block_pilot_ub_from_coefficients()`, which takes the max plain rate
over pilot coefficients on top of the corner bound.

# The marginal route

There is a second, mathematically distinct diagnostic worth knowing about,
because it is easy to confuse with the extended one.

Instead of appending \(\Omega\) as a coordinate, *integrate it out*. With
\(\Omega_j \sim \mathrm{Gamma}(a_j^0, r_j^0)\) integrated
unconditionally out of the group likelihood,

\[
p(y_j \mid \beta_j) \;\propto\;
\Bigl(r_j^0 + \tfrac{1}{2}e_j'e_j\Bigr)^{-(a_j^0 + n_j/2)},
\qquad e_j = y_j - D_j\beta_j,
\]

which is a generalized multivariate \(t\) centered at
\(\hat\beta_j^{\mathrm{ols}}\) with \(\nu_j = 2a_j^0 + n_j - p_{re}\)
degrees of freedom. That the location is the plain OLS estimate, with no
shrinkage toward a calibration prior, is the signature of the *independent*
normal-gamma structure.

This matters beyond the rate, because a multivariate \(t\) is **not
globally log-concave**. Its Hessian is positive definite only inside an
ellipsoid:

\[
-\nabla^2 \log f(x) \succ 0
\iff
(x - \mu)^\top \Sigma^{-1} (x - \mu) < \nu,
\]

which for the ING marginal reads
\(\|\beta_j - \hat\beta_j^{\mathrm{ols}}\|^2_{D_j'D_j}
  \le 2r_j^0 + \mathrm{RSS}_j^{\mathrm{ols}}\).

Outside that region the tangent-plane envelope of @Nygren2006 is no longer
a valid bound, since the construction assumes log-concavity. Chord
majorization on bounded non-concave annuli would restore a valid envelope
and is licensed by the same paper's Fact A.1, but it is not implemented
and the Theorem 2/3 asymptotics would need separate derivation.

The truncation window is what keeps this under control. Under truncation to
\([\omega_L, \omega_U] = [1/\texttt{disp\_upper}, 1/\texttt{disp\_lower}]\)
the moments are exact via `pgamma()`:

\[
H_j(\beta_j) = E_t[\Omega_j]\,D_j'D_j
             - \mathrm{Var}_t[\Omega_j]\,(D_j'e_j)(D_j'e_j)',
\]

and tightening the window mainly *reduces* \(\mathrm{Var}_t[\Omega_j]\),
raising the effective threshold \(E_t/\mathrm{Var}_t\) and making
log-concavity violations harder to reach. So the window is doing three
jobs at once: keeping the chain geometrically ergodic, giving a compact box
for the corner bound, and keeping the envelope sampler valid.

`.two_block_truncated_omega_moments()` and
`.two_block_lambda_star_marginal_over_draws()` implement this route, and
`.two_block_combine_rate_envelopes()` takes a componentwise maximum against
the plug-in bound.

# Theorem or heuristic: the honest table

| Result | Status |
|---|---|
| Claims 1–3, Theorem 3, Corollary 1 for fixed \(\Psi\), Gaussian target | **Proved** |
| Corner bound \(\lambda^*(\Lambda,\Omega) \le \lambda^*(\Lambda^{\max}, \Omega_{\min})\) | **Proved** (\(\Omega\) monotonicity general; \(\lambda_p\) complete for \(p_{re}=1\), cited otherwise) |
| \(\lambda_{\max}(A(u)) \ge \lambda^*_{\mathrm{original}}\), minimized at \(u=0\) | **Proved** |
| Multivariate-\(t\) log-concavity boundary \(d < \nu\) | **Proved** |
| ING marginal is a generalized multivariate \(t\) | **Proved** |
| Truncated Hessian moments via `pgamma()` | Exact algebra |
| **Extended \(\lambda^*\) from `two_block_rate_ing()`** | **Local, uncertified** |
| Non-Gaussian weights at `b_mode` | Local-Gaussian heuristic |
| Max over pilot or main draws | Empirical safeguard |
| Chord majorization for non-log-concave regions | Licensed in principle, not implemented |

The three-way split to keep in mind: production calibration uses the
**certified corner bound**; `two_block_rate_ing()` is a **local joint
diagnostic** answering a question the certificate does not; and the
marginal route is a **third, independent** check used by the `known_vcov`
engine. They are complementary, and none substitutes for another.

# See also

- `vignette("Chapter-B02")` — the fixed-\(\Psi\) theory being extended.
- `vignette("Chapter-B05")` — where the truncation window comes from.
- `vignette("Chapter-B07")` — the \(\Omega\) side, per group.
- `?two_block_rate`, `?two_block_rate_from_pfamily_list`
- `inst/BLOCK_GIBBS_ERGODICITY_ING.md`,
  `inst/omega-ing-marginal-multivariate-t.md`,
  `inst/multivariate-t-log-concavity.md` in the package sources.
