---
title: "Chapter B03: Calibrating m_convergence"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B03: Calibrating m_convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B03: Calibrating `m_convergence`

`Chapter-B02` produced a computable bound on the total-variation distance
between the \(l\)-step kernel and the target. This chapter spends it. The
question is operational: given a total-variation tolerance `tv_tol`, how
many inner Gibbs sweeps should separate the chain start from a stored
draw?

The answer is `m_convergence`, and the user never has to choose it. They
choose a tolerance; the engine derives the sweeps.

# The pipeline

```
pfamily_list + design
        |
        v
two_block_rate_from_pfamily_list()   -->  lambda*, eigenvalue spectrum
        |
        v
two_block_l_for_tv(rate, tv_tol)     -->  smallest l with TV(l) <= tv_tol
        |
        v
m_convergence = l + 1                     (+1: Block 1 lags a half-step)
```

Three properties make this practical.

**The bound is decreasing in \(l\).** So inverting it is exact: a doubling
search followed by bisection finds the smallest sufficient \(l\), with no
tuning constants.

**The cost is logarithmic in \(1/\texttt{tv\_tol}\).** Tightening the
tolerance by three orders of magnitude does not triple the work; it adds a
handful of sweeps. This is what makes the "certify every draw" contract
affordable.

**The chain starts at the posterior mean.** \(D_0 = 0\), the mean term of
the bound vanishes, and only the variance-convergence sum remains.

```{r setup-model}
J      <- 12L
n_j    <- 8L
sigma2 <- 1.0
tau2   <- 0.5

group   <- factor(rep(seq_len(J), each = n_j))
Z       <- matrix(1, J * n_j, 1L, dimnames = list(NULL, "(Intercept)"))
x_hyper <- list("(Intercept)" = matrix(1, J, 1L,
                                       dimnames = list(levels(group), "(Intercept)")))

rate <- two_block_rate(
  x                 = Z,
  group             = group,
  x_hyper           = x_hyper,
  prior_list_block1 = list(Sigma = matrix(tau2), dispersion = sigma2),
  prior_list_block2 = list("(Intercept)" = list(mu = 0, Sigma = matrix(100)))
)
rate$lambda_star
```

The print method lays out the whole calibration table at once — the naive
geometric proxy alongside the two certified bounds:

```{r rate-table}
rate
```

Read the columns as three increasingly good answers to the same question.
The `(lambda*)^m` proxy asks only when the *slowest* eigendirection has
decayed to `tol`, ignoring that the others converged sweeps ago. Corollary
1 keeps the geometric shape but supplies the correct constants. Theorem 3
evaluates the exact per-direction terms and is the sharpest of the three;
it is what the sampler uses.

# What `tv_tol` means

`tv_tol` is a **per-draw** guarantee: each stored draw is within `tv_tol`
of the exact joint posterior in total variation. It is not a statement
about the sample as a whole.

The default is `tv_tol = 0.01`, the conventional threshold of the honest
burn-in literature [@JonesHobert2001]. To certify an entire `n`-draw
sample at level \(\alpha\), pass `tv_tol = alpha / n` — a union bound over
draws. Because the sweep cost is logarithmic, this is cheap:

```{r tv-tol-cost}
tols <- c(1e-2, 1e-3, 1e-4, 1e-6, 1e-8)
data.frame(
  tv_tol = tols,
  sweeps = vapply(tols,
                  function(t) lmebayesCore:::two_block_l_for_tv(rate, t),
                  integer(1))
)
```

Going from a per-draw tolerance of \(10^{-2}\) to a sample-wide
certificate at \(n = 1000\) (i.e. \(10^{-5}\)) costs a few sweeps, not a
few thousand.

The bound itself, evaluated directly:

```{r tv-curve}
l <- 1:8
signif(lmebayesCore:::two_block_tv_bound(rate, l), 3)
```

# Why one extra sweep

The Theorem 3 bound applies to the block updated **second** in each sweep,
which is Block 2 (\(\gamma\)). The stored Block 1 draw (\(b\)) is produced
*before* that Block 2 update, so it lags by a half-step: the bound at
\(l-1\) is what applies to it.

The sampler therefore sets

```
m_convergence = two_block_l_for_tv(rate, tv_tol) + 1
```

so that **both** blocks of every stored draw satisfy the tolerance. If you
compare a hand-computed `two_block_l_for_tv()` against a fit's reported
`m_convergence` and find an off-by-one, this is it.

# Starting points and the \(D_0\) term

Everything so far assumed \(D_0 = 0\). That is exact for the Gaussian
fixed-variance-component case, where `lmerb_posterior_mean()` computes the
joint posterior mean in closed form by iterated conditional modes.

Two situations break it:

1. **Non-Gaussian responses.** The posterior mode is available
   (`glmerb_posterior_mode()`), but the mode is not the mean, so a chain
   started there is at some unknown positive distance from the target
   mean.
2. **Sampled variance components.** The joint target over
   \((\gamma, \tau^2)\) is not Gaussian, so there is no exact mean to start
   from.

In both cases the engine runs a **pilot stage**: a batch of `n_pilot`
short chains whose cross-chain mean is used as the start for the main
chains. The point of the pilot is not to produce draws — they are
discarded — but to produce a *starting point with a certified distance
from the truth*.

If the pilot mean is an average over `n_pilot` independent chains, its
sampling distribution has covariance \(\Sigma_{11}/n_{\mathrm{pilot}}\),
so the squared standardized distance obeys
\[
D_0 \;=\; \frac{\chi^2_{p,\,\alpha}}{n_{\mathrm{pilot}}}
\]
at confidence level \(\alpha\) (default 0.95), where \(p\) is the
dimension of the Block 2 hyper vector. Larger pilots give smaller
\(D_0\), which in turn lowers the main-stage `m_convergence` — but the
pilot itself costs sweeps.

# The pilot / main cost trade-off

That is a genuine optimization. Total inner-sweep cost is
\[
C(n_{\mathrm{pilot}}) \;=\;
n_{\mathrm{pilot}}\, m_p \;+\; n \cdot m_{\mathrm{conv}}(n_{\mathrm{pilot}}),
\]
where \(m_p\) is sweeps per pilot draw and
\(m_{\mathrm{conv}}(\cdot)\) is the main-stage sweep count implied by the
\(D_0\) above. The first term grows linearly in \(n_{\mathrm{pilot}}\);
the second decreases, and flattens out once \(D_0\) is small enough that
the mean term no longer binds. There is an interior minimum.

Four internal helpers form the stack:

| Function | Role |
|---|---|
| `two_block_d0_pilot_start()` | \(D_0 = \chi^2_{p,\alpha}/n_{\mathrm{pilot}}\) |
| `two_block_m_convergence_for_pilot_start()` | main-stage sweeps at that \(D_0\) |
| `two_block_pilot_sampling_cost()` | total cost \(C\) |
| `two_block_optimize_pilot_cost()` | minimize \(C\) over \(n_{\mathrm{pilot}}\) |

```{r pilot-cost}
opt <- lmebayesCore:::two_block_optimize_pilot_cost(
  n                   = 1000L,
  rate                = rate,
  tv_tol              = 0.01,
  m_convergence_pilot = 5L,
  p                   = 1L
)
opt[c("n_pilot_opt", "m_convergence_opt", "total_cost_opt")]
```

```{r pilot-curve}
head(opt$cost_curve, 10)
```

The curve is flat near its minimum, which is good news: the optimum does
not have to be found precisely for the plan to be near-optimal. This is
the Gibbs analogue of the build-versus-sample cost optimization the iid
envelope samplers perform when deciding how finely to partition the
envelope (`glmbayesCore::EnvelopeOpt`; see
`vignette("Chapter-A08", package = "glmbayes")`).

Note the asymmetry with the mode-start minimum. As
\(n_{\mathrm{pilot}} \to \infty\), \(D_0 \to 0\) and
\(m_{\mathrm{conv}}\) approaches the \(D_0 = 0\) sweep count from
`Chapter-B02` — the pilot can never do better than starting at the exact
mean, only approach it.

# From a `pfamily_list`

In practice the Block 2 priors arrive as `pfamily` objects from
`pfamily_list()`, not as raw precision matrices.
`two_block_rate_from_pfamily_list()` is the adapter:

```{r from-pfamily, eval = FALSE}
rate <- two_block_rate_from_pfamily_list(
  x                 = design$Z,
  group             = design$group,
  x_hyper           = design$X_hyper,
  prior_list_block1 = list(Sigma = Sigma_ranef, dispersion = sigma2),
  pfamily_list      = pfamily_list(ps)
)
```

It extracts the dispersion each component implies:

- **`dNormal`** — the fixed `dispersion`, i.e. the known \(\tau^2_k\).
- **`dIndependent_Normal_Gamma`** — the conservative `disp_lower` plug-in,
  the lower end of the truncation window.

The second choice is the one that makes the calibration honest for ING
models. Smaller \(\tau^2\) means tighter cross-block coupling and a larger
\(\lambda^*\), so a bound computed at `disp_lower` is valid for **every**
\(\tau^2\) in the truncated support. It is conservative rather than
correct-on-average, which is the right direction for a certificate. That
conservatism is also why ING routes carry a heavier eigenvalue bound than
their fixed-\(\tau^2\) counterparts, and why `disp_lower` must be set with
some care — see `Chapter-B05`.

# When certification fails

`two_block_l_for_tv()` searches up to `l_max` sweeps. If the bound has
still not reached the tolerance — which happens when \(\lambda^*\) is very
close to 1 — it issues a warning and returns the **uncertified geometric
proxy** \(l\) from `rate$m_for_tol(tol)` rather than `l_max`. The result
is a usable sweep count that is explicitly not a certificate.

Upstream of that, `two_block_rate()` refuses to return at all if
\(\lambda^*\) is numerically indistinguishable from 1: it prints the
\(P_{11}\), \(P_{12}\), \(P_{21}\), \(P_{22}\) blocks and stops. A
\(\lambda^*\) of 1 means the joint posterior is improper in some
direction, so the honest response is a failed rank condition
(`Chapter-B08`), not a larger sweep count.

Between those, a warning fires whenever \(\lambda^*\) exceeds 0.95:
convergence is still geometric, but the coupling is strong enough that the
model is worth a second look. The usual causes are small groups (small
\(C_j\) relative to \(1/\tau^2\)) or a variance component driven toward
zero.

# See also

- **Chapter B02** — where the bound comes from.
- **Chapter B05** — the ING truncation window and `disp_lower`.
- **Chapter B10** — checking the calibration empirically with
  `plot_mean_convergence()` and `plot_var_convergence()`.
- `?two_block_rate`, `?two_block_tv_bound`, `?two_block_optimize_pilot_cost`.
