---
title: "Chapter B09: Row-block engines"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B09: Row-block engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B09: Row-block engines

A row-block engine partitions the rows of a design into groups and runs one
regression per group. That description fits two quite different things in
this ecosystem, and conflating them is the main hazard this chapter exists
to prevent:

- **`rNormal_reg_group()` / `rNormalGLM_reg_group()`** — the compiled block
  engines in this package. They are what Block 1 of the mixed-model sampler
  calls, every sweep.
- **`lmbBlock()` / `glmbBlock()`** in **lmebayes** — independent "by group"
  fits with no population level. Despite the name, they do **not** use the
  engines above; they loop over blocks calling `glmbayes::lmb()`.

Same shape, different code, different statistics.

# The engines

```{r sigs}
args(rNormal_reg_group)
args(rNormalGLM_reg_group)
```

The Gaussian engine draws each block's coefficients from its normal full
conditional; the GLM engine uses the envelope sampler of @Nygren2006 per
block. Both accept either a single `prior_list` shared across blocks or a
`prior_lists` giving one per block.

The compiled path partitions once and loops in C++:

```
rNormalGLM_reg_group()  →  .block_rNormalGLM_cpp()
                        →  block_rNormalGLM_cpp_export()   [src/block_utils.cpp]
                        →  rNormalGLMBlocks()              [src/rNormalGLMBlocks.cpp]
                        →  rNormalGLM()  per block         [glmbayesCore]

rNormal_reg_group()     →  .block_rNormalReg_cpp()
                        →  block_rNormalReg_cpp_export()
                        →  rNormalRegBlocks()
                        →  rNormalReg()  per block         [glmbayesCore]
```

The per-block draw itself is **glmbayesCore**'s. What this package adds is
the partition, the per-block prior plumbing, and the loop — which is
precisely the part that has to be fast, since Block 1 executes it once per
sweep per chain.

A small worked example:

```{r engine-demo}
set.seed(1)
J <- 6; n_j <- 20
g <- factor(rep(seq_len(J), each = n_j))
x <- cbind("(Intercept)" = 1, slope = rnorm(J * n_j))
beta_true <- cbind(rnorm(J, 10, 2), rnorm(J, 2, 0.5))
y <- rowSums(x * beta_true[as.integer(g), ]) + rnorm(J * n_j)

fit <- rNormal_reg_group(
  n = 100, y = y, x = x, group = g,
  prior_list = list(mu = c(0, 0), Sigma = diag(c(100, 100)), dispersion = 1)
)
class(fit)
dim(fit$coefficients)
round(head(fit$coefficients, 3), 3)
```

`coefficients` and `coef.mode` are \(k \times l_1\) matrices — **one row per
block**, one column per predictor. This is worth internalizing, because it
is the opposite of the draws-in-rows convention used elsewhere: here the
block index is the row.

```{r rownames}
rownames(fit$coefficients)
colnames(fit$coefficients)
```

## The partition object

`normalize_group()` turns any of the accepted `group` specifications into a
canonical description:

```{r group-info}
str(fit$group_info, max.level = 1)
```

| Field | Meaning |
|---|---|
| `k` | number of blocks |
| `ids[j]` | block label, in factor-level order |
| `rows[[j]]` | 1-based row indices into the full `y` and `x` |
| `l2_blocks[j]` | `length(rows[[j]])` |
| `starts[j]` | cumulative start position in the stacked layout |

`starts` and `l2_blocks` are what the C++ loop uses; `rows` is what R-side
code uses. Keeping both avoids recomputing the partition on either side.

# Block 1 of the mixed-model sampler

This is where the engines earn their keep. Inside `rGLMM_sweep()`:

```
rGLMM_sweep
  └── .two_block_block1_all_chains()
        ├── .two_block_block1_prep_all_chains()
        │     └── build_mu_all(design, fixef_i)        # γ → per-group prior means
        │     └── .two_block_block1_prior_with_tau2()  # τ² → prior precision
        └── .two_block_block1_draw_all_chains()
              └── .two_block_block1_draw_block()
                    └── rNormal_reg_group() / rNormalGLM_reg_group()
```

The prep step is the whole difference between a Gibbs sweep and a set of
independent fits. Each sweep, for each chain:

```r
fixef_i    <- .two_block_batch_fixef_chain(batch, i)
mu_all     <- build_mu_all(design, fixef_i, batch$group_levels)$mu_all
prior_list <- .two_block_block1_prior_with_tau2(block1_prior, tau2_i, ...)
```

`build_mu_all()` evaluates \(\mathcal{W}_j\gamma\) for every group at the
current Block 2 draw and returns a \(p_{re} \times J\) matrix. That matrix
becomes `prior_list$mu`, in which **column \(j\) is block \(j\)'s prior
mean**, matched by position. Separately,
`.two_block_block1_prior_with_tau2()` refreshes the diagonal of the prior
precision, `P[k, k] <- 1/tau2_k`, from the current \(\tau^2\).

So both halves of each block's prior — mean and precision — are rebuilt
from the population state before every draw. That is the coupling; the
engine itself is stateless.

Comparing the two uses side by side:

| | Independent fits | Block 1 |
|---|---|---|
| Prior source | user's `pfamily` | `build_mu_all()` + current \(\tau^2\) |
| Prior mean | fixed | \(l_1 \times k\) matrix, refreshed each sweep |
| Prior precision | fixed | diagonal refreshed from \(\tau^2\) |
| Draws per call | `n = 1000` typically | **`n = 1L`** per chain per sweep |
| Offset / weights | user-supplied | `rep(0, l2)` / `rep(1, l2)` |
| GLM settings | user defaults | `use_parallel = FALSE`, `n_envopt = 1`, quiet |
| After the draw | nothing | rows permuted to `batch$group_levels` |

Two of these are easy to trip over when reading the source. `n = 1L` is
right because the sweep loop supplies the iteration; asking for more draws
per sweep would be sampling from a conditional that is about to change.
And the Block 1 wrapper deliberately does not forward `design$offset` or
`design$weights` — those are already absorbed upstream, so passing them
again would double-count.

The reordering step matters because `normalize_group()` orders blocks by
factor level while the sweep batch carries its own `group_levels` ordering;
`.two_block_block1_draw_reorder()` reconciles them so that downstream code
can index by position.

# `lmbBlock()` and `glmbBlock()` are not these engines

Here is the part most worth stating explicitly. `lmbBlock()` has exactly
**one** runtime dependency on this package:

```r
lmebayesCore:::normalize_group(block_vec, l2)
```

called from `.blmb_formula_block_meta()`. That is all. The flow is:

1. `.blmb_formula_block_meta()` — build the model frame, resolve the block
   specification, call `normalize_group()`.
2. `.mrglmb_normalize_pfamily_lists()` / `.validate_pfamily_for_rlmb()` —
   prior validation, from **glmbayesCore**.
3. For `b = 1..k`: subset the rows, call `glmbayes::lmb()` or
   `glmbayes::glmb()`.
4. `.blmb_assemble()` / `.bglmb_assemble()` — collect the results.

No call reaches `rNormal_reg_group()`, `rNormalGLM_reg_group()`, or any
compiled code in this package. The block wrappers borrow a partition helper
and otherwise delegate wholesale to **glmbayes**.

The result is a `blmb` (or `bglmb`) object: a named list of `lmb` (or
`glmb`) fits with `call`, `formula`, `block`, `group_info`, `k`, `p`,
`pred_names` and `pfamily_lists` as attributes — plus `family` for the GLM
case.

Why keep them separate at all? Because they answer different questions. The
block wrappers give genuinely independent per-block posteriors, which is
what you want when the blocks are not exchangeable or when you want a
no-pooling baseline to compare a mixed model against. Block 1's draws are
conditional on the population state and are not marginally independent;
they are one component of a Gibbs sweep, not a standalone posterior.

The user-facing treatment is
`vignette("Chapter-12", package = "lmebayes")`.

# Per-block prior calibration

`Prior_SetupGroup()` is the row-block analogue of `Prior_Setup_GLMM()`: it
calls `glmbayesCore::Prior_Setup()` once per block on that block's rows and
returns a named list of per-block results, classed `"Prior_SetupGroup"`,
with `call`, `formula`, `group` and `group_info` attached.

There is no second level here, so there is nothing to calibrate *across*
blocks — each block's prior is derived from its own data and nothing else.
That independence is exactly what makes the identifiability requirement
stricter, which is the next section.

# Identifiability, without a second level

Chapter B08 established that a rank-deficient group in the hierarchical
model is survivable: the population conditional supplies what the data
cannot. Independent block fitting has no population conditional, so that
repair is unavailable.

Consequently only level 1 applies, and it is not advisory. A block whose
design matrix is rank deficient cannot identify its coefficients from its
data, and under a weak prior the posterior is improper in the deficient
direction. Such blocks should be **dropped before fitting**, not passed in
and flagged.

`block_check_identifiability()` in **lmebayes** does the screening. It
returns `level1_table`, `level1_keep`, `level1_drop`, `level2_rank`,
`level2_q`, `level2_ok` and `action`, and takes
`on_failure = c("warn", "stop")`.

The level-2 fields are present for the coupled case rather than the
independent one, and its failure message is the sharpest statement of why
that condition is the fatal one:

```
block_check_identifiability: Level 2 FAILED -- rank of X_nbhd restricted to
Level-1-identified blocks is {l2_rank} (need {q}). The population parameter
mu is not identified by the data-bearing blocks; the coupled Gibbs chain
will be null recurrent in {q - l2_rank} direction(s) of mu even with a
proper prior.
```

"Even with a proper prior" is the operative phrase, and the contrast with
level 1 — where a proper prior *is* the fix — is the whole content of
Chapter B08's asymmetry.

# Who owns what

Since three packages are involved, the division of labor in one table:

| Layer | Package |
|---|---|
| Per-block iid draw (`rNormalGLM`, `rNormalReg`) | **glmbayesCore** |
| Family dispatch (`glmbfamfunc`) | **glmbayesCore** |
| Per-block prior calibration (`Prior_Setup`) | **glmbayesCore** |
| Row partition (`normalize_group`) | **lmebayesCore** |
| Block orchestration and C++ loop | **lmebayesCore** |
| Hierarchical sampler (`build_mu_all`, `rGLMM_sweep`) | **lmebayesCore** |
| Formula interface for iid GLMs (`lmb`, `glmb`) | **glmbayes** |
| Block wrappers (`lmbBlock`, `glmbBlock`) | **lmebayes** |
| Mixed-model front door (`lmerb`, `glmerb`) | **lmebayes** |

The dependency worth noting: **glmbayes has no dependency on
lmebayesCore.** The arrow runs the other way, and `lmbBlock()` reaching in
for `normalize_group()` is the single exception in the whole stack.

# See also

- `vignette("Chapter-B01")` — the sweep loop that drives Block 1.
- `vignette("Chapter-B08")` — the identifiability conditions.
- `vignette("Chapter-12", package = "lmebayes")` — `lmbBlock()` for users.
- `?rNormal_reg_group`, `?rNormalGLM_reg_group`, `?Prior_SetupGroup`
- `inst/BLOCK1_draw_partition_and_labeling.md`,
  `inst/LMBBLOCK_LMEBAYESCORE_DEPENDENCIES.md` in the package sources.
