---
title: "Chapter B04: The known-vcov Gaussian route and the exact iid sampler"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B04: The known-vcov Gaussian route and the exact iid sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B04: The known-vcov Gaussian route

One combination of inputs makes the whole hierarchical posterior collapse
to a multivariate normal, and the engine exploits it aggressively. This
appendix documents that route: when it is selected, why it is exact, and
what the two engines behind `sim_method` actually do.

The conditions are:

1. the observation dispersion \(\sigma^2\) is **fixed** — a scalar
   (`disp_mode = "fixed"`) or a named per-group vector
   (`"fixed_vector"`) — rather than sampled; **and**
2. every Block 2 component uses a `dNormal()` pfamily, so every
   \(\tau^2_k\) is a known constant rather than a sampled quantity.

Under both, no accept-reject envelope, no ICM iteration, and no pilot stage
appears anywhere in the call graph.

# Route selection

`dispersion_ranef` is resolved by `.lmebayes_resolve_dispersion_ranef()`
into one of five modes:

| `disp_mode` | Trigger | Sampled? | Route family |
|---|---|---|---|
| `"none"` | Non-Gaussian family with no free dispersion | — | `glmm_*` |
| `"fixed"` | Single positive scalar | No | `lmm_fixed_*` |
| `"fixed_vector"` | Named length-\(J\) numeric vector | No | `lmm_fixed_*` |
| `"gamma"` | One `dGamma(Inv_Dispersion = TRUE)` | Yes | `lmm_gamma_*` |
| `"gamma_list"` | Named list of per-group `dGamma()` | Yes | `lmm_gamma_*` |

`.lmebayes_reg_route_key()` then appends a suffix from the Block 2 prior
summary: `known` when `pf_summary$all_dNormal` is `TRUE`, `estimated`
otherwise. This appendix is the `lmm_fixed_known` cell, implemented by
`rLMMNormal_reg_known_vcov()`.

The lookup is table-driven. `REG_ROUTE_TABLE` maps route keys to functions
and records whether each needs a pilot stage:

```
lmm_fixed_known      -> rLMMNormal_reg_known_vcov        (needs_pilot = FALSE)
lmm_fixed_estimated  -> rLMMNormal_reg_estimated_vcov    (needs_pilot = TRUE)
lmm_gamma_*          -> rLMMindepNormalGamma_reg_*       (needs_pilot = TRUE)
glmm_*               -> rGLMM_reg*                       (needs_pilot = TRUE)
```

`.lmebayes_run_lmm_engine()` performs `route_key -> route_fn -> do.call()`
and then `.lmebayes_attach_sigma2()`. Only the `lmm_fixed_known` branch
receives a `sim_method` argument; every other route ignores it, because
only this one has a second engine to dispatch to.

# Why the route is exact

With fixed \(\sigma^2_j\) and known \(\tau^2_k\), every density in the
hierarchy is Gaussian:

\[
y_j \mid b_j \sim N(Z_j b_j,\ \sigma^2_j I), \qquad
b_j \mid \gamma \sim N(\mu_j(\gamma),\ \Sigma_b), \qquad
\gamma_k \sim N(\mu_{\gamma_k},\ \Sigma_{\gamma_k}),
\]

with \(\mu_j(\gamma)\) *linear* in the stacked \(\gamma\) through the map
`build_mu_all()` computes, \(\mu_{jk}(\gamma) = [X_k \gamma_k]_j\). A
Gaussian likelihood with a Gaussian prior whose mean is linear in the
hyperparameters gives a jointly Gaussian posterior over
\((\gamma, b_1, \dots, b_J)\). Two consequences follow.

**The posterior mean equals the posterior mode**, and it is one linear
solve away. `lmerb_posterior_mean()` forms, per group,

```
ZtZ_scaled[[j]] = crossprod(Z_j) / sigma2_j
Zty_scaled[[j]] = crossprod(Z_j, y_j) / sigma2_j
```

and eliminates every \(b_j\) by Schur complement. Each \(b_j\) is affine in
the shared \(\gamma\) and never couples to another group's \(b_{j'}\), so
what remains is a single system in \(\gamma\) whose dimension is the total
Block 2 hyperparameter count — independent of \(J\). No \(J \times J\)
matrix is ever formed, and the build cost is \(O(J)\). The returned
`converged = TRUE, iterations = 1L, delta = 0` is literal: there is no
iteration.

**The Theorem 3 convergence rate is exact.** The rate of Chapter B02
depends only on the *linear* Gibbs-sweep operator built from \(P_b\),
\(P_{\gamma_k}\), \(\sigma^2_j\), and the design. With no GLM link and no
non-Gaussian random-effect prior, there is no nonlinearity to bound
conservatively, and `convergence_info$method` is tagged `"exact"` rather
than a `disp_lower` plug-in.

`"fixed_vector"` differs from `"fixed"` only in letting \(\sigma^2_j\)
vary by group. Every formula above already carries the group subscript, so
the mode is not new machinery — only correct group indexing of the
\(\sigma^2\) vector in the ICM solver and in `.two_block_rate_inputs()`,
where a length-\(J\) dispersion is expanded to one working weight per
observation.

# Two engines behind one route

`rLMMNormal_reg_known_vcov()` is a thin dispatcher validated by
`.rLMM_validate_sim_method()`, which accepts exactly `"DEFAULT"` or
`"TWO_BLOCK_GIBBS"`.

| `sim_method` | Engine | Sweeps | `convergence_info$method` |
|---|---|---|---|
| `"DEFAULT"` | `rLMMNormal_reg_known_vcov_iid()` | none | `"exact_iid"` |
| `"TWO_BLOCK_GIBBS"` | `rLMMNormal_reg_known_vcov_two_bg()` | `m_convergence` | `"exact"` |

Both engines validate identically and both reject a non-`dNormal`
`pfamily_list` with a message pointing at `rLMMNormal_reg_estimated_vcov()`.
Both reshape their output through the same `.rLMM_format_v2_out()`, so
downstream staging needs no engine-specific branching.

## The exact iid engine

`rLMMNormal_joint_iid()` is the interesting one. Because the target is
exactly Gaussian, it can be factored once and sampled directly:

1. `.lmerb_posterior_normal_system(design, measurement_prior_list)` builds
   `M`, the posterior precision of the stacked Block 2 vector
   `gamma_full`, and per group `post_P_j`, the conditional precision of
   \(b_j \mid \gamma\). Both are independent of \(\gamma\).
2. `.lmerb_posterior_system_cholesky(system)` Cholesky-factors `M` and
   every `post_P_j` **once**.
3. Each of the `n` draws is then one `backsolve()` against `chol(M)` for
   `gamma_full`, plus one `backsolve()` per group against `chol(post_P_j)`
   for \(b_j\).

No iteration, no burn-in, no autocorrelation. `m_convergence` is set to
`1L` and `draw_engine` to `"rLMMNormal_joint_iid"`.

Note that step 1 reuses exactly the matrices `lmerb_posterior_mean()`
already assembles — the engine that reports the point estimate and the
engine that samples share their algebra, so a discrepancy between
`popef.mode` and `colMeans(popef)` is a genuine bug signal rather than
two independent approximations disagreeing.

## The symmetry precondition

`.lmerb_posterior_system_cholesky()` checks that `M` is numerically
symmetric before factoring, symmetrizes defensively within a relative
tolerance, and stops with a message pointing at
`sim_method = "TWO_BLOCK_GIBBS"` when the asymmetry is too large to ignore.

The check exists because symmetry is only *guaranteed* when `Sigma_ranef`
from `prior_list_block1` is diagonal **and** its \(k\)-th diagonal entry
equals component \(k\)'s `dNormal()` dispersion \(\tau^2_k\) in
`pfamily_list`. That is exactly how `lmerb()` and `glmerb()` construct it,
but it is not enforced at the matrix level, where a caller could supply
inconsistent inputs. The Gibbs engine tolerates the inconsistency; the
Cholesky engine cannot, so it fails loudly rather than returning wrong
draws.

## The Gibbs engine on an exact target

`rLMMNormal_reg_known_vcov_two_bg()` runs the full pipeline of Chapter B01:
`.rLMM_icm_at_start()` for the exact start point,
`.rLMM_calibrate_m_convergence()` for the sweep count, then
`two_block_rNormal_reg()` for `n` replicate chains.

Running Gibbs on a target you can sample exactly is not wasted work — it is
how the sweep machinery is validated. The two engines target the same
distribution, so any disagreement beyond Monte Carlo error localizes a
defect to the sweep kernel. `vignette("Chapter-06", package = "lmebayes")`
shows the comparison from the user side.

# The precision matrix `P`

One design decision is worth flagging because it removes a whole class of
bug. The Block 1 prior precision `P` is **never** a caller-supplied
argument. `.lmebayes_block1_prior_list()` builds only
`list(dispersion = ..., ddef = FALSE)`; every routed export derives its own
`P` from the validated `pfamily_list` via `.rLMM_P_from_pfamily_list()`,
which takes one \(\tau^2_k\) plug-in per component through
`.two_block_tau2_plug_in_vector()`, assembles `diag(tau2)`, and inverts.

Each export then *rejects* a caller-supplied `P` or `Sigma` before calling
`.two_block_validate_block1_prior()`. The invariant is that `P` can never
disagree with `pfamily_list`, because it is always a function of it.

# The compiled kernel

The Gibbs engine's inner loop is C++:

```
.two_block_rNormal_reg_cpp()            [rcpp_wrappers.R]
  -> block1_prior_list()                [twoBlockGibbs.cpp]
  -> MuAllBuilder::build()              [twoBlockGibbs.cpp]
  -> block_rNormalReg_cpp_export()      [block_utils.cpp]
       -> rNormalRegBlocks()            [rNormalRegBlocks.cpp]
            -> rNormalReg()             [rNormalReg.cpp]
  -> (Block 2 loop, is_ing = false)
       -> rNormalReg()
```

Nothing in this chain calls an accept-reject envelope, `rglmb()`, or
`lm.fit()`. Every draw is a direct conjugate-normal sample. Contrast this
with the ING and GLMM routes, where the Block 2 (respectively Block 1) draw
goes through `glmbayesCore::rglmb()` and the envelope sampler of
@Nygren2006.

# Working example

```{r example}
set.seed(1)
J <- 12L; n_j <- 8L
group <- factor(rep(seq_len(J), each = n_j))
Z <- matrix(1, nrow = J * n_j, ncol = 1L, dimnames = list(NULL, "(Intercept)"))
x_hyper <- list("(Intercept)" = matrix(1, nrow = J, ncol = 1L,
                                       dimnames = list(levels(group), "(Intercept)")))

tau2 <- 0.5; sigma2 <- 1
```

The rate for this balanced design is available directly, without running
the sampler:

```{r example-rate}
rate <- two_block_rate(
  x                 = Z,
  group             = group,
  x_hyper           = x_hyper,
  prior_list_block1 = list(Sigma = matrix(tau2), dispersion = sigma2),
  prior_list_block2 = list("(Intercept)" = list(mu = 0, Sigma = matrix(100)))
)
c(computed = rate$lambda_star,
  closed_form = sigma2 / (sigma2 + n_j * tau2))
```

For a balanced random-intercept model with a diffuse Block 2 prior the rate
has the closed form \(\lambda^* = \sigma^2 / (\sigma^2 + n_j \tau^2)\) —
the complement of the shrinkage ratio in
`vignette("Chapter-02", package = "lmebayes")`, seen from the sampler's
side. Larger groups carry more likelihood precision and mix faster; the
rate does not depend on \(J\).

```{r example-print}
rate
```

The `Implied sweeps per tolerance` table is `two_block_l_for_tv()`
evaluated at several tolerances, under all three bounds: the geometric
\((\lambda^*)^m\) proxy, the Corollary 1 envelope, and the Theorem 3 terms
the sampler actually uses.

# When the route does *not* apply

Any of the following moves you off `lmm_fixed_known`:

| Change | Resulting route | What is lost |
|---|---|---|
| Any `dIndependent_Normal_Gamma` component | `lmm_fixed_estimated` | Exactness; pilot stage required (B05, B06) |
| `dispersion_ranef` a `dGamma()` or list | `lmm_gamma_*` | Exactness; pilot stage required (B07) |
| Non-Gaussian `family` | `glmm_*` | Exact rate becomes a local-Gaussian heuristic |

In every one of those cases the posterior is no longer Gaussian, the mode
is no longer the mean, and the convergence rate becomes a plug-in bound
evaluated conservatively rather than a theorem. That progression is the
subject of the next three appendices.

# See also

- `vignette("Chapter-B01")` — the two-block architecture and sweep loop.
- `vignette("Chapter-B02")` — the rate and TV bound this route computes
  exactly.
- `vignette("Chapter-B05")` — what changes when \(\tau^2_k\) is sampled.
- `vignette("Chapter-B07")` — what changes when \(\sigma^2\) is sampled.
- `?rLMMNormal_reg`, `?two_block_rate`
- `inst/README_KNOWN_VCOV_GAUSSIAN.md` in the package sources for the
  function-by-function call graph.
