---
title: "Chapter B02: Ergodicity and total-variation convergence"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B02: Ergodicity and total-variation convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B02: Ergodicity and total-variation convergence

This is the chapter the rest of the engine rests on. A two-block Gibbs
sampler is only useful here if we can say, for a specific model and
dataset, *how many sweeps* separate a stored draw from the target — not
asymptotically, but with a number. For the Gaussian model with fixed
variance components that number is computable, and this chapter derives
it.

The results are stated for the exactly-Gaussian case. Section
"Beyond the Gaussian case" explains precisely what survives when the
likelihood is Poisson, binomial, or Gamma, and what does not.

# The scalar picture

Start with one scalar hyper-mean \(\mu\) and one group coefficient
\(\beta_b\), variance components fixed:

| Symbol | Role |
|---|---|
| \(\mu\) | population mean (Block 2) |
| \(\beta_b\) | group coefficient (Block 1) |
| \(P_{11}\) | prior precision on \(\mu\) |
| \(A = 1/\tau^2\) | coupling precision, \(\beta_b \mid \mu\) |
| \(C_b = x_b^2/\sigma^2\) | likelihood precision from group \(b\) |

Expanding the joint log-density as a quadratic form in \((\mu, \beta)\)
gives the joint precision

\[
P = \begin{pmatrix} P_{11} + A & -A \\ -A & A + C_b \end{pmatrix}.
\]

The off-diagonal is \(-A\): the two blocks are coupled **only through the
hierarchical prior**, never through the data. For a two-dimensional
Gaussian Gibbs sampler the contraction factor — the spectral radius of the
Markov operator on the centered subspace — is the squared canonical
correlation,

\[
r \;=\; \frac{P_{12}^2}{P_{11}^{\text{tot}} P_{22}}
  \;=\; \frac{A^2}{(P_{11}+A)(A+C_b)} .
\]

Reading off the regimes:

| Regime | \(r\) | Behaviour |
|---|---|---|
| \(P_{11} \to \infty\) (tight prior on \(\mu\)) | \(\to 0\) | instant mixing |
| \(P_{11} > 0\), \(C_b = 0\) | \(0 < r < 1\) | geometrically ergodic |
| \(P_{11} \to 0\), \(C_b = 0\) | \(\to 1\) | mixing time \(\to \infty\) |
| \(P_{11} = 0\), \(C_b = 0\) | \(\det P = 0\) | improper joint, **null recurrent** |

The last row is the failure mode. When a group contributes no likelihood
precision in some direction — a rank-deficient design — and the prior on
\(\mu\) is flat in that direction, the joint posterior is improper and the
chain does not converge at all.

**Data rescues the chain.** With \(C_b > 0\), even a single observation,

\[
r = \frac{A^2}{(P_{11}+A)(A+C_b)} \;<\; \frac{A}{P_{11}+A}
\]

for any \(P_{11} \ge 0\). Under a flat prior on \(\mu\) this is still
\(r \approx A/(A + C_b) < 1\): the data pins down \(\beta\), breaking the
deterministic dependence on \(\mu\).

Notice what the formula says about *design*, not just propriety. \(r\)
is close to 1 whenever \(C_b \ll A\), i.e. when each group carries little
information relative to the shrinkage prior — small groups, or a small
\(\tau^2\). Large, well-separated groups mix fast. This intuition is
exactly what the multivariate rate below generalizes.

# The multivariate rate

In the general model of `Chapter-B00`, order the parameters so that the
paper's \(x_1\) is the block updated **second** in each sweep — the Block 2
hyper vector \(\gamma\), of dimension \(q\) — and \(x_2\) is the Block 1
stack \(b\), of dimension \(J p_{re}\). The joint posterior precision
blocks are

\[
\begin{aligned}
P_{11} &= \sum_j H_j' P_b H_j + \mathrm{blockdiag}_k(V_k^{-1}) && (q \times q)\\
P_{22} &= \mathrm{blockdiag}_j\big(Z_j' W_j Z_j + P_b\big) && (Jp_{re} \times Jp_{re})\\
P_{12} &= \begin{bmatrix} -H_1'P_b & \cdots & -H_J'P_b \end{bmatrix} && (q \times Jp_{re})
\end{aligned}
\]

where \(P_b = \Psi^{-1}\) is the Block 1 prior precision, \(V_k\) is the
Block 2 prior covariance for coefficient \(k\), \(W_j\) collects the
per-observation weights (\(1/\sigma^2\) in the Gaussian case), and
\(H_j\) (\(p_{re} \times q\)) maps \(\gamma\) to group \(j\)'s Block 1
prior mean — row \(k\) of \(H_j\) is `X_hyper[[k]][j, ]` placed in
\(\gamma_k\)'s columns.

The **Remark 8 rate matrix** is

\[
A \;=\; P_{11}^{-1/2}\, P_{12}\, P_{22}^{-1}\, P_{21}\, P_{11}^{-1/2},
\]

a \(q \times q\) symmetric matrix with eigenvalues
\(a_1 \ge a_2 \ge \dots \ge a_q\) in \([0, 1)\). The largest,
\[
\lambda^* = a_1,
\]
is the geometric contraction rate of the two-block Gibbs sampler
[@Nygren2020]: the total-variation distance between the \(l\)-step kernel
and the target decays like \((\lambda^*)^l\).

## Computing it without forming \(P_{22}\)

\(P_{22}\) is \(Jp_{re} \times Jp_{re}\) and is never built. Because
\(P_{22}\) is block-diagonal by group and each \(H_j\) has exactly one
non-zero segment per coefficient,

\[
P_{12} P_{22}^{-1} P_{21} \;=\; \sum_{j=1}^{J} H_j'\Big[P_b B_j^{-1} P_b\Big] H_j,
\qquad B_j = Z_j' W_j Z_j + P_b,
\]

accumulates over groups using only \(p_{re} \times p_{re}\) solves. One
\(q \times q\) symmetric eigendecomposition finishes the job. Nothing
larger than \(q \times q\) is ever decomposed, so the cost is linear in
\(J\) and the diagnostic is cheap enough to run before sampling — which is
what `lmerb()` does on every fit.

`two_block_rate()` implements exactly this.

## A worked case with a closed form

For a balanced random-intercept model — \(J\) groups, \(n_j\) observations
each, \(p_{re} = q = 1\), and a flat Block 2 prior — the algebra collapses
to
\[
\lambda^* = \frac{1/\tau^2}{1/\tau^2 + n_j/\sigma^2}
          = \frac{\sigma^2}{\sigma^2 + n_j \tau^2},
\]
which is the scalar formula \(A/(A + C_b)\) with \(C_b = n_j/\sigma^2\).

```{r rate-closed-form}
J      <- 12L   # groups
n_j    <- 8L    # observations per group
sigma2 <- 1.0   # observation variance
tau2   <- 0.5   # random-intercept variance

group <- factor(rep(seq_len(J), each = n_j))
Z     <- matrix(1, nrow = J * n_j, ncol = 1L,
                dimnames = list(NULL, "(Intercept)"))
x_hyper <- list("(Intercept)" = matrix(1, nrow = J, ncol = 1L,
                                       dimnames = list(levels(group),
                                                       "(Intercept)")))

rate <- two_block_rate(
  x                 = Z,
  group             = group,
  x_hyper           = x_hyper,
  prior_list_block1 = list(Sigma = matrix(tau2), dispersion = sigma2),
  prior_list_block2 = list("(Intercept)" = list(mu = 0, Sigma = matrix(100)))
)

c(computed = rate$lambda_star,
  closed_form = sigma2 / (sigma2 + n_j * tau2))
```

The small discrepancy is the Block 2 prior: the closed form above assumes
\(V \to \infty\), while the call used \(V = 100\). Tighten the prior and
\(\lambda^*\) falls, exactly as the scalar table predicted.

The rate depends on \(n_j\) but **not** on \(J\): adding groups adds
information about \(\gamma\) and adds parameters to \(b\) in equal
measure. Making groups larger, on the other hand, helps a great deal.

```{r rate-vs-nj}
lam <- function(n_j) {
  g <- factor(rep(seq_len(J), each = n_j))
  two_block_rate(
    x                 = matrix(1, J * n_j, 1L, dimnames = list(NULL, "(Intercept)")),
    group             = g,
    x_hyper           = list("(Intercept)" = matrix(1, J, 1L,
                                                    dimnames = list(levels(g), "(Intercept)"))),
    prior_list_block1 = list(Sigma = matrix(tau2), dispersion = sigma2),
    prior_list_block2 = list("(Intercept)" = list(mu = 0, Sigma = matrix(100)))
  )$lambda_star
}
round(vapply(c(2L, 4L, 8L, 16L, 32L), lam, numeric(1)), 4)
```

The printed form of a rate object adds the implied sweep counts for
common tolerances, which is the subject of `Chapter-B03`:

```{r rate-print}
rate
```

# From the rate to a total-variation bound

Knowing that TV decays geometrically is not the same as knowing the
constant. The bound actually used is Theorem 3 of [@Nygren2020], which
gives the distance as an explicit sum over eigendirections rather than a
single geometric envelope.

Order the eigenvalues **ascending**, \(a_0 = 0 \le a_1 \le \dots \le a_q\)
(the Lemma 2 ordering). After \(l\) sweeps the bound is

\[
\mathrm{TV}\big(K^l(x^{(0)}, \cdot),\, \pi\big) \;\le\;
\underbrace{\sum_{i=1}^{q} d_i^{(l)}}_{\text{variance convergence}}
\;+\;
\underbrace{(\lambda^*)^{l}\sqrt{D_0/(2\pi)}}_{\text{mean convergence}},
\]

capped at 1, where each \(d_i^{(l)}\) is a closed-form expression in the
ratio
\[
r_i^{(l)} = \frac{1 - a_{i-1}^{2l}}{1 - a_i^{2l}},
\]
evaluated through the generalized error function
\[
\mathrm{erf}_n(x) = P\big(\chi^2_n \le 2x^2\big),
\]
which reduces to the classical error function at \(n = 1\)
[@Nygren2020, Remark 3; @Brown1963]. The looser Corollary 1 form replaces
the exact terms with a geometric envelope decaying like \(a_i^{2l}\) with
explicit constants.

\(D_0\) is the squared standardized distance of the starting point from
the posterior mean,
\[
D_0 = (x^{(0)} - \mu)^{\top}\,\Sigma_{11}^{-1}\,(x^{(0)} - \mu).
\]

**Starting at the posterior mean sets \(D_0 = 0\) and the mean term
vanishes identically.** This is the single most consequential
implementation choice in the package: `lmerb()` computes the exact joint
posterior mean by iterated conditional modes before sampling and starts
every replicate chain there, so only the variance-convergence sum has to
be certified. Chapter B03 shows what this buys in sweeps.

Two internal helpers evaluate and invert the bound:

```{r tv-bound}
lmebayesCore:::two_block_tv_bound(rate, l = 1:6)             # theorem 3, D0 = 0
lmebayesCore:::two_block_l_for_tv(rate, tol = 1e-3)          # sweeps for TV <= 1e-3
```

The exact Theorem 3 terms are considerably sharper than the naive
\((\lambda^*)^m\) proxy — often by a factor of two in sweeps — because the
proxy ignores that all but the leading eigendirection have already
converged.

Both helpers are `\keyword{internal}` topics: reach them with
`lmebayesCore:::`, or read them off the `print()` output of a
`two_block_rate` object, which is the supported route.

A final bookkeeping point. The bound applies to the block updated
**second** in each sweep, i.e. \(\gamma\). The stored Block 1 draw lags by
a half-step, so the bound at \(l-1\) applies to \(b\). The sampler adds
one sweep to compensate; that is why a certified `m_convergence` is
always one more than `two_block_l_for_tv()` returns.

# The full-rank conditions

Geometric ergodicity is not automatic. It requires the joint posterior to
be proper in every direction, and that imposes **two** rank conditions —
one at each level of the hierarchy.

## Level 1: each group's design must be full column rank

A rank-deficient \(D_j\) contributes zero likelihood precision in the
deficient direction, \(C_b = 0\), which is the degenerate scalar case
above. In the \(p\)-dimensional Gaussian model with fixed \(\Sigma\), full
column rank of \(D_j\) means \(C_j = D_j'D_j/\sigma^2\) is positive
definite, so every eigendirection of the data block contributes positively
to the Schur complement. Rank deficiency puts at least one direction back
into the null-recurrent regime.

**Full column rank of every \(D_j\) is therefore necessary for the joint
posterior to be proper in all directions, and necessary for geometric
ergodicity when the prior on \(\gamma\) is weak or improper.**

## Level 2: the retained groups must identify \(\gamma\)

The second condition is subtler and is the caller's responsibility. Take
the hyper-regression \(\beta_b \mid \gamma \sim N(x_b^{\top}\gamma, \tau^2)\)
with group-level covariates \(x_b\). Integrating out a group with no data
(\(C_b = 0\)) removes its contribution **exactly**: the Schur-complement
correction cancels the coupling term it introduced. A data-free group is
informationally equivalent to a group that never existed, and its
covariate vector \(x_b\) is irrelevant.

What is left, after integrating out every non-identified group, is the
marginal precision
\[
P_{\gamma|\mathrm{marg}} \;=\; P_{11} \;+\; \sum_{b:\, C_b > 0}
   x_b x_b^{\top} \cdot \frac{A C_b}{A + C_b}.
\]

With \(P_{11} = 0\) this is positive definite **iff the group-level
covariate vectors of the data-bearing groups span \(\mathbb{R}^q\)**:
\[
\mathrm{rank}\big(X_{\mathrm{hyper}}[\text{data-bearing rows}, ]\big) = q .
\]

| Level | Condition | Ensures |
|---|---|---|
| **Data** (per group) | `qr(D_j)$rank == p_re` | \(C_j \succ 0\): \(\beta_j\) is identified by its own data |
| **Hyper** (across groups) | `qr(X_hyper[kept rows, ])$rank == q` | \(\gamma\) is identified through the data-bearing groups |

Both must hold. If either fails there is a direction in
\((\beta, \gamma)\) space along which the chain is null recurrent — it will
still produce numbers, and they will be meaningless.

The condition is not vacuous: adding a group-level covariate that is
constant across all retained groups introduces a degenerate direction in
\(\gamma\) in exactly the way rank deficiency in \(D_j\) introduces one in
\(\beta_j\). `check_identifiability()` runs both checks; the intercept-only
special case (\(x_b = 1\) for all \(b\)) reduces level 2 to "at least one
full-rank data-bearing group exists".

The rate function also guards the boundary directly: if the computed
\(\lambda^*\) is numerically indistinguishable from 1,
`two_block_rate()` stops and prints the offending precision blocks rather
than returning a rate that would certify nothing.

# Beyond the Gaussian case

For Poisson, binomial, and Gamma responses the joint posterior is not
normal and **no theorem above applies as stated**. What carries over, and
what does not:

**Carries over — the structural argument.** For log-concave GLM
likelihoods, \(C_b\) is replaced by the expected Fisher information at the
current draw. A rank-deficient \(D_j\) still gives zero information in
some direction, still makes that direction's conditional improper under a
weak prior on \(\gamma\), and still breaks ergodicity. Full-rank designs
remain **necessary**. Sufficiency for the coupled chain needs additional
conditions and is not claimed.

**Does not carry over — the numerical rate.** \(\lambda^*\) is only
defined for a Gaussian target. What the package does instead is evaluate
the Gaussian rate at IRLS-style curvature weights taken at the posterior
mode:
\[
w_i = \frac{\mathrm{wt}_i\,\mu'(\eta_i)^2}{V(\mu_i)\,\phi},
\qquad \eta_i = \mathrm{offset}_i + z_i^{\top} b_{\mathrm{mode},\,j(i)} .
\]
For canonical links this is the exact observed Hessian; otherwise it is
the expected (Fisher) information. Supplying these as `weights` to
`two_block_rate()` yields a **local-Gaussian heuristic** rate — a good
practical guide, not a certificate. The print method labels such objects
accordingly, and non-Gaussian fits run a pilot stage rather than trusting
the number outright.

The same caveat applies, for a different reason, when variance components
are sampled rather than fixed: the target is then not Gaussian in
\((\gamma, \tau^2)\) jointly. The package's response is to evaluate the
rate at the **conservative** end of the truncated \(\tau^2\) support —
smaller \(\tau^2\) means tighter coupling and a larger \(\lambda^*\), so a
bound computed at `disp_lower` holds throughout the support.
`Chapter-B05` covers the truncation window, and `Chapter-B06` covers the
extended local rate diagnostic that tracks \(\tau^2\) explicitly.

# See also

- **Chapter B03** — turning \(\lambda^*\) into `m_convergence`, and the
  pilot/main cost trade-off.
- **Chapter B06** — `two_block_rate_ing()`, the local rate when \(\tau^2\)
  is sampled.
- **Chapter B08** — `check_identifiability()` and the rank conditions in
  practice.
- `?two_block_rate` — the function reference, which also documents
  `two_block_rate_from_pfamily_list()` and `two_block_rate_ing()`.
- `inst/BLOCK_GIBBS_ERGODICITY.md` in the package sources, the maintainer
  note this chapter is drawn from.
