Yaml

---
title: "Chapter A4: t-statistics for Bayesian Linear models with normal-gamma priors"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter A4: t-statistics for Bayesian Linear models with normal-gamma priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(glmbayes)
```

** 1.0  Why the Bayesian tail can be smaller

There are three distinct, concrete reasons the Bayesian posterior tail probability for a coefficient can be smaller than the classical t tail

*** 1.1 Different effective degrees of freedom (**v**)

In the conjugate parameterization above, posterior $\boldsymbol{v = 2\alpha_{0} + n}$.
If the prior has positive $\boldsymbol{\alpha_{0}}$ (an informative prior on the variance),
the posterior $\boldsymbol{v}$ will be larger than the classical $n - p$.
Increasing $v$ makes the Student‑t closer to normal (lighter tails), so the tail probability
at the same $t$ value is smaller.

-- Even for vaguely inforative priors, the prior contributes "pseudo-observations" to the posterior **df**. This is often the main cause of smaller Bayesian tails

*** 1.2 Different standard error (scale) used in the denominator

Bayesian marginal variance for $\boldsymbol{\beta_{j}}$ is 
$\boldsymbol{\frac{\beta_{n}}{\alpha_{n}} V_{n}[j,j]}$.

Two effects happen here:  
(a) $\boldsymbol{V_{n}}$ generally shrinks relative to $(X^{T}X)^{-1}$ because of the prior precision $\boldsymbol{V_{0}^{-1}}$, and  
(b) the posterior scale factor $\boldsymbol{\frac{\beta_{n}}{\alpha_{n}}}$ can be smaller than the classical estimate $\hat{s}^{2}$ depending on prior hyperparameters and the data (e.g., prior favors smaller variance).

Either effect reduces the denominator and can increase the raw $t$ value — but combined with larger $\boldsymbol{v}$ the net tail probability can still be smaller.


So you must compare both the numerator (posterior mean $\boldsymbol{m_{n,j}}$) 
vs. the OLS estimate $\hat{\beta}_{j}$, and the denominator 
(posterior scale vs. $\boldsymbol{\hat{s}\sqrt{(X^{T}X)^{-1}_{jj}}}$). 
Prior shrinkage often makes $m_{n,j}$ closer to zero and $V_{n}[j,j]$ smaller, 
both of which reduce evidence against $H_{0}$.

### 1.3 Different centering (posterior mean vs. OLS) and prior shrinkage

- If the prior pulls $\boldsymbol{m_{n,j}}$ toward $\boldsymbol{\beta_{0}}$ (e.g., 0), 
  the numerator of the Bayesian $t$ is smaller in magnitude than $\hat{\beta}_{j}$, 
  further reducing tail probability. In combination with (1) and (2), the posterior 
  tail probability will often be noticeably smaller than the classical one. 
  Griffin & Brown and other shrinkage‑prior literature discuss this tradeoff 
  explicitly for normal–gamma shrinkage priors.


For Bayesian linear models with normal-gamma priors, it seems like standard errors used in t-statistic calculations need to be adjusted relative to the classical estimates. In particular, the adjustment seems to be

The adjusted standard error is:

\[
\text{Std\_err}_t = \text{Std\_err} \cdot \sqrt{ \frac{n - p}{n + n_{\text{prior}}} }
\]

which leads to:


\[
t_{\text{adj},i}
= \frac{n}{n + n_{\text{prior}}}
  \, t_{\text{unadjusted},i}
  \, \sqrt{\frac{n + n_{\text{prior}}}{n - p}}.
\]

where the first ratio shows the skrinkage effect (lower t-value) and the second effect the impact of the adjustment to the standard error (which boosts the t-value).


\[
t_{\text{adj},i} = \sqrt{\frac{n}{n + n_{\text{prior}}}} \cdot \sqrt{\frac{n}{n -p}}  \cdot t_{\text{unadjusted},i} 
\]




\[
t_{\text{adj},i} = \sqrt{\frac{n}{n + n_{\text{prior}}}} \cdot \sqrt{\frac{n}{n -p}}  \cdot t_{\text{unadjusted},i} 
\]

\[
t_{\text{adj},i} = \sqrt{1-pwt} \cdot \sqrt{\frac{n}{n -p}}  \cdot t_{\text{unadjusted},i} 
\]


It can be seen from the above that t-statitic in the Bayesian context is larger than the unadjusted t-statistic (i.e., inflated)  whenver


\[
n_{\text{prior}} < p \cdot \frac{n}{n - p}
\]


which implies tail probabilities smaller than their classical counterparts.


\[
n + n_{\text{prior}}
= n + n\left(\frac{\text{pwt}}{1 - \text{pwt}}\right)
= \frac{n(1 - \text{pwt}) + n(\text{pwt})}{1 - \text{pwt}}
= \frac{n}{1 - \text{pwt}}.
\]




The adjusted standard error is
\[
\text{StdErr}_{t}
= \text{StdErr} \cdot 
\sqrt{\frac{n - p}{n + n_{\text{prior}}}},
\]
which leads to the adjusted t-statistic


\[
t_{\text{adj},i}
= t_{\text{unadjusted},i} \cdot 
\sqrt{\frac{n + n_{\text{prior}}}{n - p}}.
\]

For reference, the same expression can be written as


\[
\text{StdErr}_{t}
= \text{StdErr} \cdot 
\sqrt{\frac{n - p}{n + n_{\text{prior}}}}.
\]

which leads to


\[
t_{\text{adj},i}
= t_{\text{unadjusted},i} \cdot
  \sqrt{\frac{n + n_{\text{prior}}}{n - p}}.
\]





The **glmb** function and related *method functions* that handle the output are designed to be Bayesian versions of the **glm** function and many of its *method functions*. This vignette shows how the basic setup/calling of the functions compare and then walks through how the method functions for glmb can be called to generate similar outputs to those from the glm functions.

**Dobson Randomized Control Data**

To understand how the outputs of the glmb function mirrors those for the glm function, it is useful to take a look at the first portion of the example that is provided with the glm function. The data is based on Randomized Controlled Trial data from Dobson (1990). Here is a view of the data:

```{r dobson}
## Dobson (1990) Page 93: Randomized Controlled Trial :
counts <- c(18,17,15,20,10,20,25,13,12)
outcome <- gl(3,1,9)
treatment <- gl(3,3)
print(d.AD <- data.frame(treatment, outcome, counts))
```

**Calling the two functions**

The example code for the glm function specifies a Poisson regression model for this data. The below code chunks shows how the glmb function can be used in much the same fashion but with some extra requirements. In particular, we need to provide a prior distribution (in this case a Multivariate Normal prior) and related constants (in this case, a prior Mean and and a prior Variance-Covariance matrix) for the coefficients of interest. Optionally, we can also tell the function how many draws to make from the posterior distribution (the default used here generates 1000 iid draws). 

Here is the call to the classical *glm* function:

```{r glm_call,results = "hide"}
## Call to glm
glm.D93 <- glm(counts ~ outcome + treatment, 
              family = poisson())
```

To use the *glmb* function, we first use a call to a function *Prior_Setup" to initialize the prior and to get the variable names needed. We use the default prior from the Prior_Setup function (discussed further in a separate vignette).

```{r Prior_Setup,results = "hide"}
## Using glmb
## Step 1: Set up Default Prior 
ps=Prior_Setup(counts ~ outcome + treatment,family=poisson())
mu=ps$mu
V=ps$Sigma
```

We now call the *glmb* function and include the default prior
distribution in addition to the two required arguments for the *glm* function.

```{r Call_glmb,results = "hide"}
# Step 3: Call the glmb function
glmb.D93<-glmb(counts ~ outcome + treatment, family=poisson(), pfamily=dNormal(mu=mu,Sigma=V))
```


In the above, it is worth noting the additional step that we went through when specifying the prior. We used a call to the function *Prior_Setup* to get the correct dimensions for the mean and variance-covariance matrices and to initialize the constants. The *Prior_Setup* function also provided information on the Variable names in the design matrix (which also corresponds to the names of the coefficients eventually estimated) so we can make informed changes to the prior if so desired.

The next couple of vignette's will discuss the prior specification in more details. Here we instead focus on reviewing the output from the function and how it can be used.

**Printing the output**

Taking a look at the basic printed output, we can see that the two closely mirror each other with the glmb posterior means replacing the glm maximum likelihood estimates.


```{r Printed_Views}
## Printed view of the output from the glm function 
print(glm.D93)
## Printed view of the output from the glmb function 
print(glmb.D93)
```

In addition to the posterior means, the *glmb* printed output also returned three pieces of information that are similar to (but not quite the same) as the classical output. The "Effective Number of Parameters" should in general be close to (but not exactly the same) as the number of parameters estimated while the Expected Residual Deviance in general will be higher than the corresponding maximum likelihood estimate for the Residual Deviance (since the latter is designed to minimize it). Finally, the DIC is essentially a Bayesian version of the AIC. These measures will be discussed in greater detail in one of our later vignettes.

**Methods Available**

In addition to the basic print function output, the glmb function returns an object with an assigned class "glmb" for which a number of generic functions (or methods) are available. The class "glmb" inherits from "glm" and "lm" and as such many functions for those classes work directly for "glmb". For some of the instances where the inherited methods fail and/or could produce incorrect results, we have implemented methods specifically for the *glmb* class. The methods for classes *lm*, *glm*, and *glmb* are listed below. We will use many of these later in this vignette.

```{r lm_Methods}
## Methods for class "lm"
methods(class="lm")
```

```{r glm_Methods}
## Methods for class "glm"
methods(class="glm")
```

```{r glmb_Methods}
## Methods for class "glmb"
methods(class="glmb")
```

**The summary functions**

Let's take a closer look at the outputs of the summary functions.

*glm summary output*

In turn, we see the Call, a list of Deviance residuals, and the estimated coefficients. The coefficients are then followed by some additional model related information. 

```{r glm_summary}
## summary output for the "glm" class
summary(glm.D93)
```

*glmb summary output*

The summary for the glmb function follow a similar structure but adds a table containing information related to the prior and the maximum likelihood above the table with the means for the estimated Bayesian coefficients. The output below the main table with coefficients is also modified to contain similar (but slightly different) pieces of information (the details of which are discussed elsewhere).

```{r glmb_summary}
## summary output for the "glm" class
summary(glmb.D93)
```

**Model Fit, Predictions, Deviance Residuals, Covariance Matrices, and Confidence/Credible Intervals**

Let's next take a look at the outputs from these additional methods to see how they compare. Note that the Bayesian version of these contain random draws tied to the underlying distributions so the column means are mostly used in these comparisons.

*Fitted values*

```{r glm fitted outputs}
## fitted outputs for the glm function
fitted(glm.D93)
```


```{r glmb fitted outputs}
## mean of fitted outputs for the glm function
## works without a "glmb" class specific generic function
colMeans(fitted(glmb.D93))
```

*Predictions(linear predictors)*

```{r glm predictions}
## predictions for the glm function
predict(glm.D93)
```

```{r glmb predictions}
## predictions for the glmb function
colMeans(glmb.D93$linear.predictors) # no current predict function
colMeans(predict(glmb.D93)) 
```

*Residuals*

```{r glm residuals}
## residuals for the glm function
residuals(glm.D93)
```

```{r glmb residuals}
## residuals for the glmb function
colMeans(residuals(glmb.D93))
```

*vcov*

```{r glm vcov}
## vcov for the glm function
vcov(glm.D93)
```

```{r glmb vcov}
## vcov for the glmb function
vcov(glmb.D93)
```

*confint*

Confidence intervals
```{r glm confint}
## confint for the glm function
confint(glm.D93)
```

```{r glmb confint}
## confint for the glmb function
confint(glmb.D93)
```


**AIC/DIC, Deviance, and the Log-Likelihood**

These model statistics are useful when comparing different model specifications. The Bayesian versions of these will be discussed in greater detail in a separate Vignette.

*AIC/DIC*

```{r glm AIC}
## AIC for the glm function (equivalent degrees of freedom and the AIC)
extractAIC(glm.D93)
```

```{r glmb DIC}
## DIC for the glmb function (estimated effective number of parameters and the DIC)
extractAIC(glmb.D93)
```

*Deviance*

```{r glm Deviance}
## Deviance for the glm function
deviance(glm.D93)
```

```{r glmb Deviance}
## Deviance for the glmb function
## works without a "glmb" class specific generic function
mean(deviance(glmb.D93))
```

*Log-Likelihoods*

```{r glm logLik}
## Deviance for the glm function
logLik(glm.D93)
```

```{r glmb logLik}
## Deviance for the glmb function
mean(logLik(glmb.D93))
```

**Model Frame, formula, family, nobs, and show**

These are mostly useful for understanding various aspects of the model.

*Model Frames*

```{r glm Model Frame}
## Model Frame for the glm function
model.frame(glm.D93)
```

```{r glmb Model Frame}
## Model Frame for the glmb function
model.frame(glmb.D93$glm)
```

*formula*

```{r glm formula}
## formula for the glm function
formula(glm.D93)
```

```{r glmb formula}
## formula for the glmb function
formula(glmb.D93)
```
*family*

```{r glm family}
## family for the glm function
family(glm.D93)
```


```{r glmb family}
## family for the glmb function
family(glmb.D93$glm)
```
*nobs*

```{r glm nobs}
## nobs for the glm function
nobs(glm.D93)
```

```{r glmb nobs}
## nobs for the glmb function
nobs(glmb.D93)
```


*show*

The show method returns the same output as the print function