---
title: "Chapter B00: Engine roadmap and notation"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B00: Engine roadmap and notation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B00: Engine roadmap and notation

**lmebayesCore** is the engine package behind
[**lmebayes**](https://github.com/knygren/lmebayes). It owns the
formula-to-design pipeline (`model_setup()`), the Block 2 hyperprior
calibration (`Prior_Setup_GLMM()`), the matrix-level two-block Gibbs
samplers (`rlmerb()`, `rglmerb()`, and the `rLMM_reg` / `rGLMM_reg` route
families), and the convergence theory that decides how many inner Gibbs
sweeps each stored draw needs.

This `B`-series documents that engine. It is written for maintainers,
method developers, and readers who want the derivations behind the
numbers that `lmebayes` prints. If you want to *fit models*, start with
the `lmebayes` curriculum (`Chapter-00` through `Chapter-13`) instead.

## Two vignette series, two audiences

| Series | Package | Audience |
|---|---|---|
| `Chapter-00` … `Chapter-13` | **lmebayes** | Applied users fitting mixed models |
| `Chapter-B00` … `Chapter-B10` | **lmebayesCore** | Engine internals, theory, derivations |
| `Chapter-00` … `Chapter-18`, `Chapter-A01` … `A12` | **glmbayes** | The iid GLM stack this builds on |

The `Chapter-A*` files that ship in the **glmbayes** and **glmbayesCore**
sources describe the *iid* envelope machinery — likelihood subgradient
densities [@Nygren2006], envelope construction, and the `pfamily` prior
system. Every Block 1 and Block 2 draw made by this package ultimately
calls into that machinery through `glmbayesCore`. Read
`glmbayes` `Chapter-A05` (simulation methods) and `Chapter-A08`
(envelope functions) if the accept-reject layer is unfamiliar.

## The B-series

- **B00 — Engine roadmap and notation** (this chapter). Symbol glossary and
  the map between the internal parameterization and `lme4`.
- **B01 — The two-block Gibbs architecture.** Block 1 / Block 2, the
  sweep-outer driver, replicate chains, and where C++ takes over.
- **B02 — Ergodicity and total-variation convergence.** The rate matrix
  \(A = P_{11}^{-1/2} P_{12} P_{22}^{-1} P_{21} P_{11}^{-1/2}\), the
  full-rank conditions, and why the chain is geometrically ergodic.
- **B03 — Calibrating `m_convergence`.** From \(\lambda^*\) to a certified
  sweep count, and the pilot-versus-main cost trade-off.
- **B04 — The known-vcov Gaussian route and the exact iid sampler.**
- **B05 — Estimated variance components (ING Block 2).**
- **B06 — The extended local rate diagnostic.**
- **B07 — Per-group measurement dispersion.**
- **B08 — Design, identifiability and rank conditions.**
- **B09 — Row-block engines.**
- **B10 — Sweep history and convergence plots.**

# The model

For group \(j = 1, \dots, J\) with \(n_j\) observations and \(P\)
group-varying coefficients, the engine works with a **centered**
(Lindley–Smith) parameterization [@LindleySmith1972]:

**Stage 1 (likelihood).**
\[
y_j \mid \beta_j, \sigma^2_j \;\sim\; N\!\big(D_j \beta_j,\; \sigma^2_j I_{n_j}\big)
\]

**Stage 2 (hierarchical prior).**
\[
\beta_j \mid \gamma, \Psi \;\sim\; N\!\big(\mathcal{W}_j \gamma,\; \Psi\big)
\]

The full, non-centered coefficient vector \(\beta_j\) — not a mean-zero
deviation — enters the likelihood directly. This is the single most
important structural difference from `lme4`, and it is what makes the
two-block split natural: **Block 1** draws every \(\beta_j\) given
\(\gamma\), and **Block 2** draws \(\gamma\) given all \(\beta_j\).

For non-Gaussian families the stage-1 normal is replaced by the
corresponding log-concave GLM likelihood with link \(g\), and the Block 1
draw goes through the envelope sampler rather than a conjugate normal.

## Symbol table

| Symbol | Name | Dimension | Role |
|---|---|---|---|
| \(y_j\) | response, group \(j\) | \(n_j \times 1\) | data |
| \(D_j\) | likelihood design, group \(j\) | \(n_j \times P\) | multiplies \(\beta_j\) |
| \(\beta_j\) | group-level coefficients | \(P \times 1\) | **not** mean zero; Block 1 state |
| \(\sigma^2_j\) | measurement (observation) dispersion | scalar | per-group or pooled |
| \(\mathcal{W}_j\) | level-2 design, group \(j\) | \(P \times q\) | block-diagonal; \(\gamma \to E[\beta_j]\) |
| \(\gamma\) | population / hyper-mean coefficients | \(q \times 1\) | Block 2 state |
| \(\Psi\) | covariance of \(\beta_j\) about \(\mathcal{W}_j\gamma\) | \(P \times P\) | diagonal, \(\Psi = \mathrm{diag}(\tau^2_k)\) |
| \(\tau^2_k\) | variance component \(k\) | scalar | known (`dNormal`) or sampled (ING) |
| \(u_j := \beta_j - \mathcal{W}_j\gamma\) | derived deviation | \(P \times 1\) | mean zero; **not** sampled directly |

\(\mathcal{W}_j\) is block-diagonal across the \(P\) coefficient
dimensions,
\[
\mathcal{W}_j = \mathrm{blockdiag}\big(W_{1j}, \dots, W_{Pj}\big),
\qquad W_{kj} \in \mathbb{R}^{1 \times q_k}, \quad q = \sum_{k=1}^{P} q_k ,
\]
so each random-effect coefficient may carry its own set of level-2
predictors. \(q_k = 1\) (intercept only) is the default when a
coefficient has no group-level covariates. This raggedness is exactly why
the level-2 designs are stored as a **list** `X_hyper` with one
\(J \times q_k\) matrix per coefficient rather than as a single padded
array.

## Internal storage

| Quantity | Storage | Shape |
|---|---|---|
| Stacked likelihood design | `Z` (matrix argument `x`) | \(n \times P\) |
| Group index | `group` | length \(n\), \(g_i = j\) |
| Level-2 designs | `X_hyper = list(X_1, …, X_P)` | `X_hyper[[k]]` is \(J \times q_k\) |
| Dimension bookkeeping | `q_k = ncol(X_hyper[[k]])`, `q = sum(q_k)` | — |

Recovery rules: \(D_j\) is `Z[group == j, ]`, and \(\mathcal{W}_j\) is
`blockdiag(X_1[j, ], …, X_P[j, ])`, assembled on demand. The design is
stored row-stacked, never as the block-diagonal \(n \times PJ\) object
`lme4` implicitly uses for \(Z\).

## Full conditionals

**Block 1** (\(\beta_j\), conjugate normal, one draw per group):
\[
\beta_j \mid \cdot \;\sim\; N(m_j, V_j), \qquad
V_j = \Big(\tfrac{1}{\sigma^2_j} D_j' D_j + \Psi^{-1}\Big)^{-1}, \qquad
m_j = V_j\Big(\tfrac{1}{\sigma^2_j} D_j' y_j + \Psi^{-1}\mathcal{W}_j\gamma\Big).
\]

**Block 2** (\(\gamma\), generalized least squares across groups):
\[
\hat\gamma = \Big(\sum_j \mathcal{W}_j' \Psi^{-1} \mathcal{W}_j\Big)^{-1}
             \sum_j \mathcal{W}_j' \Psi^{-1} \beta_j .
\]

Because \(\Psi\) is diagonal and \(\mathcal{W}_j\) is block-diagonal, the
Block 2 update **factorizes across coefficients** \(k\): each
\(\gamma_k\) is drawn from its own \(q_k\)-dimensional normal regression
of \(\beta_{\cdot k}\) on \(X_k\). That is why Block 2 is specified as a
`pfamily_list` — one prior object per random-effect coefficient — rather
than as one joint \(q\)-dimensional prior.

# Correspondence to lme4

Substituting stage 2 into stage 1 recovers the familiar marginal form
[@Bates2015]:
\[
y_j = D_j(\mathcal{W}_j\gamma + u_j) + \varepsilon_j
    = \underbrace{(D_j\mathcal{W}_j)}_{X_j}\gamma
    + \underbrace{D_j}_{Z_j} u_j + \varepsilon_j .
\]

| lmebayesCore | `lme4` / Laird–Ware | Relationship |
|---|---|---|
| \(D_j\) | — | ancestor of **both** \(X_j\) and \(Z_j\) |
| \(D_j \mathcal{W}_j\) | \(X_j\) | fixed-effects design block |
| \(D_j\) | \(Z_j\) | random-effects design block |
| \(\gamma\) | \(\beta\) | fixed effects |
| \(u_j = \beta_j - \mathcal{W}_j\gamma\) | \(b_j\) | random effects (mean zero) |
| \(\Psi\) | per-group block of \(G\) | \(G = I_J \otimes \Psi\) |
| \(\beta_j\) | \(\hat\beta + \hat b_j\) | `lme4::coef()`, when \(\mathcal{W}_j\) is intercept-only |

Note the asymmetry: \(D_j\) has no single `lme4` counterpart. When
\(\mathcal{W}_j\) is intercept-only for every coefficient
(\(q_k = 1\ \forall k\)), the model reduces to a standard random
intercept / random slope `lmer` fit, and the posterior mean of \(\gamma\)
should track `lme4::fixef()`, \(\hat\beta_j - \hat\gamma\) should track
`lme4::ranef()`, and \(\hat\beta_j\) should track `lme4::coef()`. The
package test suite uses exactly this equivalence as a correctness check.

# Model restrictions

The engine deliberately covers a narrower model class than `lme4`:

- **A single grouping factor.** No crossed or nested random effects.
- **Uncorrelated random-effect terms.** Formulas use the `||` form, so
  \(\Psi\) is diagonal. This is what lets Block 2 factorize across
  coefficients and lets the variance components be handled one at a time.
- **Log-concave likelihoods.** Gaussian, Poisson, binomial, and Gamma with
  the standard links. Log-concavity is what the envelope accept-reject
  sampler requires [@Nygren2006].

`model_setup()` enforces these and reports the resulting design in a
printable object. `check_identifiability()` adds the two rank conditions
derived in `Chapter-B08`.

# Naming conventions in the code

| Prefix / suffix | Meaning |
|---|---|
| `rLMM*` | Gaussian (linear) mixed-model replicate-chain routes |
| `rGLMM*` | Non-Gaussian replicate-chain routes |
| `*_known_vcov` | Variance components \(\tau^2_k\) treated as fixed |
| `*_estimated_vcov` | \(\tau^2_k\) sampled (ING Block 2) |
| `*_iid` | Exact independent draws, no Gibbs sweeps |
| `two_block_*` | Convergence theory, rates, TV bounds, pilot costing |
| `.two_block_*` | Internal (unexported) helpers of the above |
| `block_*`, `*_reg_group` | Row-block engines (`Chapter-B09`) |

Names beginning with a dot, and topics marked `\keyword{internal}`, are
not exported; reach them with `lmebayesCore:::` when reproducing the
diagnostics in these chapters.

# See also

- **Chapter B01** — the two-block architecture in detail.
- **Chapter B02** — ergodicity and total-variation convergence.
- `vignette("Chapter-A05", package = "glmbayes")` — the iid simulation
  methods every block draw is built from.
- `inst/notation.md` in the package sources, from which this chapter's
  symbol table is derived.
