---
title: "Chapter B07: Per-group measurement dispersion"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter B07: Per-group measurement dispersion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmebayesCore)
```

# Chapter B07: Per-group measurement dispersion

Every model so far has assumed a single measurement variance
\(\sigma^2\) shared by all groups. This appendix relaxes that to one
\(\sigma^2_j\) per group and works through what has to change: the
reference fit, the prior calibration, the truncation window, and the
Block 1 draw.

The user-facing view is
`vignette("Chapter-08", package = "lmebayes")`. This chapter is about how
the numbers are produced.

# The model, and what `dispformula` selects

Pooled (`dispformula = ~1`, the default):

\[
y_j \mid \beta_j \sim N(D_j\beta_j,\ \sigma^2 I_{n_j}),
\qquad
\beta_j \mid \gamma, \Psi \sim N(\mathcal{W}_j\gamma, \Psi).
\]

Per group (`dispformula = ~<group_name>`):

\[
y_j \mid \beta_j, \Omega_j \sim N(D_j\beta_j,\ \Omega_j^{-1} I_{n_j}),
\qquad
\Omega_j = 1/\sigma^2_j \sim \mathrm{Gamma}(a_j^0, r_j^0)
\ \text{(truncated)},
\]

with the second level untouched. The likelihood already factorized by
group; now the *prior* does too, and each group carries its own ING prior
on \((\beta_j, \sigma^2_j)\).

`dispformula` accepts only `~1` or `~<group_name>` naming the random-effects
grouping factor, and only for Gaussian responses — a Poisson or binomial
GLMM has no free observation-level dispersion to calibrate, so
`.lmebayes_dispformula_kind()` rejects it.

Switching it on changes five things at once:

| | `dispformula = ~1` | `dispformula = ~group` |
|---|---|---|
| Reference fit | `lme4::lmer()` | `glmmTMB::glmmTMB()` with the same `dispformula` |
| `calibration_source` | `"lme4"` | `"glmmTMB"` |
| `group.ing_prior` | one flat list | named list, one entry per level |
| `group.dispersion.pwt` | scalar | scalar or length-\(J\) |
| `group.max_disp_perc` | scalar | scalar or length-\(J\) |

```{r data}
data(sleepstudy, package = "lme4")
dat <- sleepstudy
dat$Days_c <- dat$Days - mean(dat$Days)
form <- Reaction ~ Days_c + (1 + Days_c || Subject)

ps_pooled <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01)
ps_group  <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01,
                              dispformula = ~Subject)

c(pooled = ps_pooled$calibration_source,
  group  = ps_group$calibration_source)
```

The pooled setup carries one number; the grouped setup carries one list per
subject:

```{r compare-structure}
ps_pooled$group.dispersion
length(ps_group$group.ing_prior)
names(ps_group$group.ing_prior)[1:4]
```

# Calibrating \(\sigma^2_j\): why the within-group RSS is the wrong target

The obvious estimator for group \(j\) is its own residual variance,

\[
\hat\sigma^2_{\mathrm{ols},j}
= \frac{\mathrm{RSS}_{\mathrm{ols},j}}{n_j - p_{re}},
\qquad
\mathrm{RSS}_{\mathrm{ols},j}
= \sum_i (y_{ij} - x_{ij}'\hat\beta_j^{\mathrm{ols}})^2,
\]

and it is the wrong target here, for a reason specific to the hierarchical
setting. \(\mathrm{RSS}_{\mathrm{ols},j}\) is computed at group \(j\)'s
*own* least-squares fit. The sampler never uses that \(\beta_j\): it uses a
shrunk one, pulled toward \(\mathcal{W}_j\gamma\). Residuals around a
shrunk coefficient are larger than residuals around the OLS coefficient, so
calibrating from the OLS RSS systematically understates the dispersion the
sampler will actually encounter.

## The marginal sum of squares

The fix is to integrate \(\beta_j\) out before measuring the residual.
With prior mean \(\mu_j\), dispersion-free prior covariance
\(\Sigma_{0j}\), and

\[
M_j = \bigl(\Sigma_{0j} + (X_j'X_j)^{-1}\bigr)^{-1},
\]

the marginal sum of squares is

\[
S_{\mathrm{marg},j}
= \mathrm{RSS}_{\mathrm{ols},j}
+ (\hat\beta_j - \mu_j)'\,M_j\,(\hat\beta_j - \mu_j),
\qquad
\hat\sigma^2_j = \frac{S_{\mathrm{marg},j}}{n_j - p_{re}}.
\]

The added term is a quadratic form in a positive definite matrix, hence
always non-negative, so \(\hat\sigma^2_j \ge \hat\sigma^2_{\mathrm{ols},j}\)
with equality only when \(\hat\beta_j = \mu_j\). The correction is exactly
the shrinkage penalty: groups whose own fit disagrees most with the
population get the largest upward adjustment, which is right, because those
are the groups whose shrunk residuals will be largest.

## The Gamma parameters

With \(n_{\mathrm{prior},j} = \dfrac{w_j}{1-w_j}\,n_j\) and
\(w_j =\) `group.dispersion.pwt` (default `0.01`):

\[
\mathrm{shape}_j = \frac{n_{\mathrm{prior},j} + 1}{2} + \frac{p_{re}}{2},
\qquad
\mathrm{rate}_j
= \frac{S_{\mathrm{marg},j}}{2}
  \cdot \frac{n_{\mathrm{prior},j} + p_{re} - 1}{n_j - p_{re}}.
\]

These satisfy the identity that makes the calibration interpretable:

\[
\frac{\mathrm{rate}_j}{\mathrm{shape}_j - 1} = \hat\sigma^2_j
\quad\text{exactly.}
\]

The Gamma prior mean *is* the marginal dispersion estimate. Check it:

```{r identity}
g1 <- ps_group$group.ing_prior[[1L]]
c(sigma2_hat = g1$sigma2_hat,
  prior_mean = g1$rate / (g1$shape - 1))
```

## `rate` versus `rate_gamma`

Each entry stores two rates, and using the wrong one is a real bug rather
than a stylistic choice.

`rate_gamma` is the fixed-\(\beta\) alternative: the RSS evaluated at the
Zellner blend \(\beta^\star_j = (1-w_j)\hat\beta_j + w_j\mu_j\) rather than
the marginal \(S_{\mathrm{marg},j}\). It is a coherent quantity, but it
answers a different question.

Which one is correct depends on what the sampler does, and here the sampler
is decisive. Block 1 draws \((\beta_j, \sigma^2_j)\) *jointly* through
`rindepNormalGamma_reg()`, which integrates \(\beta_j\) out before drawing
\(\sigma^2_j\). A prior calibrated at a fixed \(\beta\) is then mismatched
to the draw it parameterizes. So **`rate` is what reaches the sampler**;
`rate_gamma` is retained for diagnostics only.

The gap is not academic — on groups whose \(\hat\beta_j\) departs sharply
from \(\mu_j\), differences of 40–55% in the rate have been observed.

```{r rate-compare}
rates <- t(vapply(ps_group$group.ing_prior, function(g) {
  c(rate = g$rate, rate_gamma = g$rate_gamma,
    pct_diff = 100 * (g$rate - g$rate_gamma) / g$rate_gamma)
}, numeric(3)))
round(head(rates[order(-abs(rates[, "pct_diff"])), ], 5), 2)
```

## Folding in hyperparameter uncertainty

There is a second correction, applied unconditionally in current releases.
\(\mu_j\) is a stand-in for \(\mathcal{W}_j\gamma\), and \(\gamma\) is
itself uncertain. Propagating that uncertainty inflates the prior
covariance:

\[
\Omega_j[k,k] = \mathcal{W}_j[k]\,\Sigma_{\mathrm{fixef},k}\,\mathcal{W}_j[k]',
\qquad
M_j' = \bigl(\Sigma_{0j} + \Omega_{0,j} + (X_j'X_j)^{-1}\bigr)^{-1},
\]

with \(\Omega_{0,j} = \Omega_j / \mathrm{dispersion}_j^{\mathrm{classical}}\),
and \(S_{\mathrm{marg},j}'\) recomputed with \(M_j'\). The shape is
unchanged; only the rate moves. A larger \(\Omega_j\) means a larger
\(M_j^{-1}\), hence a *smaller* \(M_j'\), hence a smaller penalty — being
less certain about \(\gamma\) makes a given deviation
\(\hat\beta_j - \mu_j\) less surprising, so less of it is charged to
\(\sigma^2_j\).

The \(\Omega_j\) matrix is stored per group for inspection:

```{r omega}
ps_group$group.ing_prior[[1L]]$omega_j
```

# The truncation window

Chapter B05 introduced `disp_lower`/`disp_upper` as the interval that keeps
the ING chain geometrically ergodic. For measurement dispersion there is a
wrinkle worth knowing, because it is a genuine asymmetry in the package.

The window is **not** a quantile interval of the prior
\(\mathrm{Gamma}(\mathrm{shape}_j, \mathrm{rate}_j)\). It uses a
*posterior*-shape Gamma, mean-matched at the same center:

\[
\mathrm{shape}^{\mathrm{post}}_j = \mathrm{shape}_j + \frac{n_j}{2},
\qquad
\mathrm{rate}^{\mathrm{post}}_j
= \hat\sigma^2_j\,(\mathrm{shape}^{\mathrm{post}}_j - 1),
\]

and then, on the precision scale with \(Q\) the `qgamma` quantile function,

\[
\mathrm{disp\_lower}_j = 1 / Q(\mathtt{max\_disp\_perc}),
\qquad
\mathrm{disp\_upper}_j = 1 / Q(1 - \mathtt{max\_disp\_perc}).
\]

The `+ n_j/2` matches the envelope builder's `shape2 = Shape + n_w/2` in
`EnvelopeDispersionBuild.cpp`, so the window brackets where the *posterior*
will live rather than where the diffuse prior would. With
\(w_j = 0.01\), the prior alone carries almost no information and its
quantiles would be uselessly wide.

To be explicit, since the two are easy to conflate: `shape` and `rate`
handed to `dGamma()` are the **prior** parameters; only `disp_lower` and
`disp_upper` are computed from `shape_post`/`rate_post`.

```{r window}
w <- t(vapply(ps_group$group.ing_prior, function(g) {
  c(n_j = g$n_j, sigma2_hat = g$sigma2_hat,
    lower = g$disp_lower, upper = g$disp_upper,
    shape_post = g$shape + g$n_j / 2)
}, numeric(5)))
round(head(w, 5), 2)
```

Two consequences of the `+ n_j/2` inflation:

- **The window narrows as \(n_j\) grows.** A group with many observations
  gets a tight bracket, because its posterior is tight. That is the
  intended behavior — a wide window on a well-determined group only weakens
  the ergodicity bound for nothing.
- **The pooled path does not do this.** With `dispformula = ~1` the window
  is still built from prior quantiles with no `+ n/2` term. This is a
  documented gap, not a design principle.

# `dGamma_list()`

`dGamma_list()` turns a calibrated per-group setup into one `dGamma()`
pfamily per group — the object Block 1 consumes.

```{r dgamma-list}
dl <- dGamma_list(ps_group)
class(dl)
length(dl)
dl[[1L]]
```

Its one interesting argument is `max_disp_perc_measurement`. Left `NULL`,
stored bounds are reused verbatim. Given a scalar or a length-\(J\) vector
in \((0.5, 1)\), the window is recomputed for every group whose stored
`max_disp_perc` differs:

```{r retune}
dl_tight <- dGamma_list(ps_group, max_disp_perc_measurement = 0.80)
wd  <- attr(dl, "window_diagnostics")
wdt <- attr(dl_tight, "window_diagnostics")
head(data.frame(
  group    = wd$group,
  lower_99 = round(wd$disp_lower),  upper_99 = round(wd$disp_upper),
  lower_80 = round(wdt$disp_lower), upper_80 = round(wdt$disp_upper)
), 5)
```

Tightening from `0.99` to `0.80` narrows every window. That is a real
trade-off rather than a free improvement: a narrower window gives a better
ergodicity bound and a smaller `m_convergence`, but truncates more genuine
posterior mass. `0.99` is the default because it is close to no truncation
at all.

The remaining attributes are diagnostics:

```{r attrs}
names(attributes(dl))
str(attr(dl, "measurement_prior_group"), max.level = 1)
```

Note what `measurement_prior_group` deliberately omits: no `mu` and no
`Sigma`. The Block 1 prior *mean* is \(\mathcal{W}_j\gamma\), recomputed
every sweep from the current Block 2 state, and the RE covariance comes
from `pfamily_list`. Only the dispersion half of the ING prior is fixed at
calibration time.

# What Block 1 does with it

The per-group route keeps the v1 joint structure: for each group, each
sweep, one call to `rindepNormalGamma_reg()` draws \((\beta_j, \sigma^2_j)\)
together from the ING conditional, dispatched through
`.two_block_block1_ing_group_draw_one_chain()`.

A v2 partition is derived in `inst/DERIVATION_sigma2_with_block2_v2.md` —
draw \(\beta \mid (\gamma, \Psi, \Omega)\) as a plain Gaussian, then

\[
\Omega_j \mid \beta_j \sim
\mathrm{Gamma}\Bigl(a_j^0 + \frac{n_j}{2},\
                    r_j^0 + \frac{1}{2}\|y_j - D_j\beta_j\|^2\Bigr),
\]

which would sidestep envelope rejection entirely. It is theory and stubs
only; v1 remains the production route. Its cost is that it needs the
extended \(P_{11}\) blocks for \(\Omega\) from Chapter B06 before the rate
could be certified.

# Effect on the rate

\(\Omega_j\) enters the ergodicity story through \(P_{22}\) alone:

\[
P_{22} = \mathrm{blockdiag}_j\bigl(D_j'\Omega_j D_j + \Psi^{-1}\bigr).
\]

This is a structural fact worth isolating. The measurement residual
\(e_j = y_j - D_j\beta_j\) depends on \(\beta_j\) only, never on
\(\gamma\), so \(\Omega\) is Hessian-orthogonal to the Block 2 coordinates
and contributes nothing to \(P_{11}\) or \(P_{12}\). Contrast \(\lambda_p\)
from Chapter B06, which enters through the RE link
\(u_p = \beta_{\cdot p} - \mathcal{W}_p\gamma_p\) and therefore touches all
four blocks.

Because \(\lambda^*\) is Loewner-non-increasing in \(\Omega\), the corner
is at \(\Omega_{\min}\), i.e. at the *largest* \(\sigma^2_j\) the window
allows — `disp_upper`. That is what
`.rLMM_measurement_disp_upper_for_rate()` supplies to
`two_block_rate_from_pfamily_list()`, and it completes the pair with
Chapter B06's `disp_lower` corner on \(\tau^2\).

Under the marginal route the plug-in becomes a Student-\(t\)-type Hessian
with a rank-one correction,

\[
H_j^{(\Omega)}(\beta_j)
= E_t[\Omega_j]\,D_j'D_j
- \mathrm{Var}_t[\Omega_j]\,(D_j'e_j)(D_j'e_j)',
\]

positive definite only inside the ellipsoid
\(q_j(\beta_j) \le E_t[\Omega_j]/\mathrm{Var}_t[\Omega_j]\) where
\(q_j = \|\beta_j - \hat\beta_j^{\mathrm{ols}}\|^2_{D_j'D_j}\). Tightening
the window shrinks \(\mathrm{Var}_t[\Omega_j]\) and so raises the
threshold, which is the third job the window does. See Chapter B06.

# Caveats

Worth stating plainly, since per-group dispersion is the most
approximation-laden path in the package.

**Empirical-Bayes double counting.** Each group's prior is calibrated from
that group's own data, so its data are used twice. The default
\(w_j = 0.01\) keeps the effect at \(O(1/J)\) but does not remove it.

**Calibration is conditional on Block 2.** The marginal integrates
\(\beta_j\) out given the current \(\mathrm{sd}_\tau\) and
\(\Sigma_{\mathrm{fixef}}\); it is not joint over \(\tau^2\) uncertainty.
This matches the two-block convention, but it is a conditional statement.

**The `glmmTMB` reference fit drives everything.** When
`calibration_source` is `"glmmTMB"`, that fit's estimates *are* the
calibration. `glmmTMB` can converge to a boundary \(\tau^2 \approx 0\)
where `lmer` finds a positive estimate, and there is no automatic fallback.
Comparing against the pooled setup's `Sigma_ranef` is the cheap check:

```{r sanity}
round(rbind(pooled = diag(ps_pooled$group.Sigma),
            group  = diag(ps_group$group.Sigma)), 3)
```

**Small groups.** With \(n_j\) barely above \(p_{re}\), the RE covariance
can be near-singular and the ING envelope can converge poorly. The
`group.alpha_target` argument (default `0.01`) searches for the smallest
per-group `group.dispersion.pwt` meeting an ellipsoid-violation target,
floored at the user's supplied weight; results land in
`ps$group.pwt_calibration`.

**Runtime diagnostics live in lmebayes.** `summary_sigma2()` is a method on
fitted `lmerb` objects in **lmebayes**, not part of this package. See
`vignette("Chapter-08", package = "lmebayes")`.

# See also

- `vignette("Chapter-B05")` — the truncation window for \(\tau^2\).
- `vignette("Chapter-B06")` — where \(\Omega\) enters the extended rate.
- `vignette("Chapter-08", package = "lmebayes")` — the user-facing view.
- `?dGamma_list`, `?Prior_Setup_GLMM`
- `inst/DGAMMA_LIST_MARGINAL_AND_BOUNDS.md`,
  `inst/P_MATRIX_MARGINAL_PRECISIONS.md`,
  `inst/DERIVATION_sigma2_with_block2_v2.md`,
  `inst/PREFLIGHT_dGamma_list.md` in the package sources.
