---
title: "Chapter C04: The two-block Gibbs sampler"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: REFERENCES.bib
reference-section-title: References
vignette: >
  %\VignetteIndexEntry{Chapter C04: The two-block Gibbs sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(lmebayesCore)
```

```{r load-stored}
.c_extdata_rds <- function(name) {
  p <- system.file("extdata", name, package = "lmebayesCore")
  if (nzchar(p) && file.exists(p)) return(p)
  for (root in c(
    file.path("..", "inst", "extdata", name),
    file.path("inst", "extdata", name)
  )) {
    if (file.exists(root)) return(normalizePath(root))
  }
  stop(
    "Cannot find ", name, " in inst/extdata/. ",
    "Run devtools::load_all() or reinstall the package.",
    call. = FALSE
  )
}

C04 <- readRDS(.c_extdata_rds("Chapter-C04-bwc-twoblock.rds"))
meta <- C04$meta
conv <- C04$convergence
```

This chapter fits the same model and prior specification as `Chapter-C02`,
but forces the two-block Gibbs route instead of the exact iid sampler.
Both chapters target the same multivariate normal posterior; here the draws
come from independent chains, each run through `r conv$m_convergence` inner
sweeps before being stored.

# Prior specification

Two quantities are held constant throughout this chapter: the observation
dispersion \(\sigma^2\) and the between-group variance matrix \(\Psi\).
Both are fixed at the defaults `Prior_Setup_GLMM()` calibrates from the
data, so neither carries a prior and neither is sampled:

```{r fixed}
C04$summary$tau2_prior_overview[, c("Prior", "E[tau2]")]
```

The three `dNormal` rows are the components of \(\Psi\), declared known.
The `Residual` row is \(\sigma^2\), likewise fixed.

That leaves the population effects \(\gamma\) as the only quantity with a
prior of its own. They receive the default multivariate normal prior,
specified separately for each group-effect component:

```{r pop-prior}
round(C04$summary$popef_prior_overview[, c("Prior Mean", "Prior.sd")], 3)
```

Every coefficient is centered at zero except the grand intercept, so the
prior mean vector describes a null, intercept-only model — the same default
as in `Chapter-C02` and the **glmbayes** family of packages. The setup is
built by two calls:

```{r prior-calls, eval = FALSE}
ps <- Prior_Setup_GLMM(form, data = dat, pop.pwt = 0.01)
pf <- pfamily_list(ps)
```

`pop.pwt = 0.01` gives the prior on \(\gamma\) the weight of about one
percent of the data.

# Estimating the model using the default settings

The call matches `Chapter-C02` except for the simulation method:

```{r call, eval = FALSE}
fit <- rlmerb(
  n              = 10000,
  design         = model_setup(form, data = dat),
  pfamily_list   = pf,
  dispprior_list = list(dispersion = ps$group.dispersion),
  sim_method     = "TWO_BLOCK_GIBBS"
)
```

The `design`, `pfamily_list` and `dispprior_list` arguments carry the same
meaning as in the previous chapter. `sim_method = "TWO_BLOCK_GIBBS"` forces
the iterative route on this known-variance-components specification; with
`sim_method = "DEFAULT"` the package would draw directly from the closed-form
posterior instead.

Under this specification the joint posterior is exactly multivariate normal,
as in `Chapter-C02`. Each stored draw is one independent chain after
`r conv$m_convergence` inner Gibbs sweeps. That budget is not a guess: it is
certified from the convergence rate
(`lambda_star` = `r round(conv$convergence_info$lambda_star, 4)`) against a
total variation tolerance, by the theory of `vignette("Chapter-C03")`. The
derivation is in the section below.

# Properties of the posterior and the two-block Gibbs procedure

The log posterior splits into the same three components as in
`Chapter-C02`: the log-likelihood, the prior on the group effects, and the
prior on the population effects. Because \(\sigma^2\), \(\Psi\) and \(V\)
are all known, the posterior is multivariate normal and factorizes into the
same conditional distributions.

## Block 1: group effects given the population effects

Given \(\gamma\), the group effects are conditionally independent across
groups:

\[
\beta_j \mid \gamma, y \sim N\!\left(b_j,\; \Sigma_j\right),
\qquad
\begin{aligned}
\Sigma_j &= \left(\sigma^{-2} D_j' D_j + \Psi^{-1}\right)^{-1}, \\
b_j &= \Sigma_j\left(\sigma^{-2} D_j' y_j + \Psi^{-1}\mathcal{W}_j\gamma\right).
\end{aligned}
\]

Block 1 of the Gibbs sampler draws each \(\beta_j\) from this conditional,
holding the current \(\gamma\) fixed.

## Block 2: population effects given the group effects

Given the group effects, the population effects receive a conjugate normal
update that combines the prior on \(\gamma\) with the information each
group's current \(\beta_j\) contributes through \(\mathcal{W}_j\). The
result is another multivariate normal conditional,
\(\gamma \mid \beta, y\), with precision that adds the prior precision
to a sum of group-level contributions. Block 2 draws \(\gamma\) from this
conditional, holding the current \(\beta\) fixed.

## One sweep and one stored draw

One inner sweep alternates the two blocks: update all \(\beta_j\), then
update \(\gamma\). Repeating the pair moves the chain closer to the target.
After `r conv$m_convergence` sweeps the chain's distribution is close enough
that the final state can be treated as a draw from the posterior. Replicating
that procedure across `r meta$n_draws` independent chains produces the stored
sample.

Unlike the iid route of `Chapter-C02`, which draws \(\gamma\) from its
marginal and then each \(\beta_j\) from its conditional in one pass, the
Gibbs route needs the sweeps because it never integrates the group effects
out analytically. Both routes target the same posterior; the Gibbs route
trades the closed-form factorization for an iterative scheme whose
convergence can be diagnosed from the sweep history.

# Where the sweep count comes from

`vignette("Chapter-C03")` develops @Nygren2020's total variation bounds for
a two-block Gibbs sampler on a multivariate normal target in the abstract.
This model is exactly that setting, so the results apply verbatim and fix
the sweep count that produced the draws above. This section instantiates
them.

## The rate matrix for this design

Take \(\gamma\) — the `r nrow(C04$rate$P11)`-vector of population effects —
as the first block, updated second in each sweep, and the group effects
\(\beta\) as the second block, updated first. Claim 2 says the mean bias
contracts by the matrix
\(P_{11}^{-1}P_{12}P_{22}^{-1}P_{21}\) each sweep, and Remark 8 says the
relevant eigenvalues are those of the symmetrized form
\(A = P_{11}^{-1/2}P_{12}P_{22}^{-1}P_{21}P_{11}^{-1/2}\) — which is why
the rate can be read off the precision blocks without ever forming
\(\Sigma\).

Here \(P_{11}\) is the prior precision of \(\gamma\), \(P_{22}\) is block
diagonal with \(j\)th block \(\sigma^{-2}D_j'D_j + \Psi^{-1}\), and
\(P_{12}\) collects the \(-\Psi^{-1}\mathcal{W}_j\) coupling terms.
`two_block_rate()` assembles these and solves the generalized eigenproblem:

```{r rate-print}
C04$rate
```

The spectrum splits sharply. Three eigenvalues sit near 0.87–0.93 and four
near 0.17–0.20. The large ones belong to the intercept and the two slope
directions, where the prior on \(\gamma\) is weak relative to the
information the group effects carry, so \(\gamma\) and \(\beta\) are
strongly coupled and the sampler moves through that subspace slowly. The
rate is \(\lambda^{*} =\)
`r round(C04$rate$lambda_star, 4)`, the largest of them.

## From the rate to a certified sweep count

Every chain starts at the exact posterior mean, so
\(D_0 = (x_1^{(0)}-\mu_1)'\Sigma_{11}^{-1}(x_1^{(0)}-\mu_1) = 0\) and the
mean term of Theorem 3 vanishes identically. What remains is the variance
sum \(\sum_i d_i^{(l)}\), evaluated at the Remark 11 ratios
\(r_i^{(l)} = (1-a_{i-1}^{2l})/(1-a_i^{2l})\):

```{r tv-bound}
l <- c(1, 5, 10, 15, 20, 24, 30)
data.frame(
  sweeps     = l,
  theorem3   = signif(lmebayesCore:::two_block_tv_bound(C04$rate, l, "theorem3"), 3),
  corollary1 = signif(lmebayesCore:::two_block_tv_bound(C04$rate, l, "corollary1"), 3)
)
```

Both columns bound the same total variation distance, and the ordering
between them is the price of Corollary 1's closed form. Theorem 3 evaluates
the error function differences exactly; Corollary 1 replaces each by its
Remark 5 and Remark 17 envelope, which is loose at small \(l\) — it exceeds
1, and so says nothing, until sweep 12 or so — but decays at the same
geometric rate afterwards. The `(lambda*)^m` column of the printout above
is neither: it is the naive proxy \((\lambda^{*})^{m}\), which ignores that
the variance converges at \(a_i^{2l}\) rather than \(a_i^{l}\) and so is
conservative by roughly a factor of two in sweeps.

At the package default tolerance of `r C04$convergence$convergence_info$tv_tol`,
Theorem 3 certifies
`r lmebayesCore:::two_block_l_for_tv(C04$rate, C04$convergence$convergence_info$tv_tol, "theorem3")`
sweeps. The sampler adds one, giving
`r C04$convergence$m_convergence`:

```{r m-convergence}
c(
  theorem3      = lmebayesCore:::two_block_l_for_tv(C04$rate, 0.01, "theorem3"),
  stored        = C04$convergence$m_convergence
)
```

The extra sweep is a half-step correction, not slack. The bound is a
statement about the block updated *second* in each sweep — \(\gamma\) — and
the stored Block 1 draw \(\beta\) lags it by half a sweep, so one more
sweep brings both blocks inside the tolerance.

Two things are worth noting about the cost. Because the bound is geometric,
tightening the tolerance is cheap: the printed table above shows that going
from `r C04$convergence$convergence_info$tv_tol` to \(10^{-6}\) — four
orders of magnitude — costs roughly three and a half times the sweeps, not
four orders of magnitude more. And because \(D_0 = 0\) removes the mean
term outright, the entire budget here is paid for variance convergence;
routes that cannot start at the posterior mean (non-Gaussian families, or
sampled variance components) must certify \(D_0\) from a pilot stage and
carry the Lemma 1 mean term as well.

# Summarizing the output

`summary()` collects the draws into the same tables as in `Chapter-C02`:

```{r summary-call, eval = FALSE}
summary(fit, groups = c("30", "15"))
```

The output is organized by block: reference random effects, Block 2 variance
components (fixed here, so posterior tables are suppressed), Block 2
population effects in three tables, and Block 1 group effects with optional
per-group detail.

Two features of the population effect tables follow from the default
`pop.pwt = 0.01`. The posterior means land close to the `lmer` estimates,
displaced slightly toward the null prior means. Read on the scale that
matters, the two fits agree throughout: every Bayesian estimate falls
within a fifth of a standard error of its `lmer` counterpart.

The `Pr(Prior_tail)` column measures the posterior mass on the opposite
side of the prior mean from the estimate — a one-sided statement about how
far the data have moved each coefficient off its prior value. Because the
prior means are the null model, that reference point is zero for every
coefficient except the intercept, so for those rows the column tracks half
the two-sided p-value the corresponding `lmer` fit would report.

The output below was produced by `data-raw/make_Chapter-C04.R` with
`n = `r meta$n_draws`` and `set.seed(`r meta$seed`)`, then stored, so that
building this vignette does not rerun the sampler.

```{r summary-print}
C04$summary
```

# Variance convergence during the main phase

The exact iid route of `Chapter-C02` produces no sweep history because it
never runs a Markov chain. The Gibbs route does, and the history is stored on
the fit as `fit$sweep_history`. The variance chart measures Claim 3 of
`vignette("Chapter-C03")` empirically. That claim is an equality,
\(\Sigma_{11}^{-1/2}\Sigma^{(l)}_{11}\Sigma_{11}^{-1/2} = I - A^{2l}\), so
after \(l\) sweeps the cross-chain covariance of the Block 2 fixed effects
should sit below the exact posterior covariance by exactly \(A^{2l}\) on the
whitened scale, and rise toward it as \(l\) grows.

Because \(\sigma^2\) and \(\Psi\) are known here, that target covariance is
known as well. `plot_var_convergence()` resolves it automatically when both
the design and the measurement prior list are supplied — it calls
`lmerb_posterior_covariance()` internally and uses the result as
\(\mathrm{Var}_{\mathrm{ref}}\), not the empirical variance at the final
sweep. The artifact stores a stripped design and a measurement prior list
built the same way the sampler does: Block~1 precision from
`pfamily_list()` (the fixed \(\tau^2_k\) plug-ins), not the REML
`group.Sigma` row of `Prior_Setup_GLMM()` (which lists the reference `lmer`
variances in its `lmer` column). Using the wrong \(\Psi\) in the reference
would make the slope traces sit far above 1 even when the chain has
converged.

```{r plot-var, fig.alt = "Cross-chain variance ratio by inner sweep, unwhitened coefficients"}
plot_var_convergence(
  C04$sweep_history,
  design                 = C04$design,
  measurement_prior_list = C04$measurement_prior_list,
  whitened               = FALSE,
  n_chains               = C04$meta$n_chains
)
```

With `whitened = FALSE`, each trace is a named population effect and the
y-axis is \(\mathrm{Var}(l) / \mathrm{Var}_{\mathrm{ref}}\) for that
coefficient, where \(\mathrm{Var}(l)\) is the cross-chain variance at inner
sweep \(l\) and \(\mathrm{Var}_{\mathrm{ref}}\) is the corresponding
diagonal entry of the exact posterior covariance. Under Claim 3 the ratio is
bounded above by 1 and rises toward it as \(l\) grows, so a well-calibrated
run shows every trace flattening just below 1 by sweep
`r conv$m_convergence`. The dotted band, when `n_chains` is supplied, is a
naive Monte Carlo ruler for how much departure from 1 is sampling noise in
the numerator alone — appropriate here because the denominator is exact.

The chart uses the main-phase history only. This route has no pilot stage;
every chain starts at the exact posterior mean and the sweeps carry the
cross-chain spread up to the target variance.

# See also

- `vignette("Chapter-C01")` — the data, model and design used here.
- `vignette("Chapter-C02")` — the same posterior sampled exactly by the iid
  route.
- `vignette("Chapter-C03")` — @Nygren2020's total variation theory, of which
  the rate, the sweep budget and the variance chart above are instances.
- `vignette("Chapter-B10")` — reading the sweep-history plots.
- `?rlmerb`, `?summary.rlmerb`, `?plot_var_convergence`, `?two_block_rate`,
  `?Prior_Setup_GLMM`, `?pfamily_list`
- `data-raw/make_Chapter-C04.R` in the package sources for the script that
  produced the stored results.
